### 2014-10-22 paper/paper.tex
First paper commit

\documentclass[11pt,letterpaper,twoside,english]{article}

\usepackage[margin=1.4in]{geometry} % controls the size of the margins

% Special symbols, etc.
\usepackage{amssymb,amsbsy,latexsym}
\usepackage{amsmath}
\usepackage{graphics, subfigure, float} 
\usepackage{cancel}
%\usepackage{todonotes}
% Encoding settings
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
\usepackage[T1]{fontenc}

\usepackage{titling} % allows posttitle command

% AMS Math packages

\usepackage{amscd,amsthm}

\usepackage{verbatim, comment} % can comment out text 
\usepackage{mdwlist}

% Graphics
\usepackage[dvips]{graphicx,epsfig,color}
\usepackage{subfigure}
%\usepackage{pst-all}
%\usepackage{pstricks-add}
\usepackage{hyperref}  % can only be used with pdflatex - gives hyperlinks
\usepackage{bm} % bold math font
\usepackage{bbm}
\usepackage{todonotes}

\newtheoremstyle{theorem}{1em}{1em}{\slshape}{0pt}{\bfseries}{.}{ }{}
\theoremstyle{theorem}
\newtheorem{theorem}{Theorem}
\newtheorem*{theorem*}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{definition}[theorem]{Definition}
\newtheorem*{claim*}{Claim}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem*{remark*}{Remark}
\newtheorem{algorithm}{Algorithm}
\newtheorem*{question*}{Question}
\newtheorem{question}{Question}
\newtheorem{example}{Example}

\providecommand{\setN}{\mathbb{N}}
\providecommand{\setZ}{\mathbb{Z}}
\providecommand{\setQ}{\mathbb{Q}}
\providecommand{\setR}{\mathbb{R}}
\providecommand{\E}{\mathrm{E}}
\providecommand{\Pr}{\mathrm{Pr}}
\providecommand{\Var}{\mathrm{Var}}

\makeatother

\title{18.821 Project 2}

\author{Yajit Jain, Deepak Narayanan, Leon Zhang}

\begin{document}

\maketitle

\section{Terms in the sequence $s_n(312)$}

We claim that the sequence of numbers $s_n(312)$ is in fact the sequence of Catalan numbers. We state this result formally as the following theorem,

\begin{theorem}
The total number of permutations of $\{1,2,3,...,n\}$ that avoid the order $312$ as a subsequence is $C_n$ where $C_n$ is the $n^{th}$ Catalan number.
\end{theorem}

Before proving the theorem, we state and prove the following lemma, that will be used in our proof of the theorem.
\begin{lemma}
All permutations of $\{1,2,...,k,k+1\}$ ending in $i$ that avoid the order $312$ as a sub-sequence must be of the form,
$$\pi_1 \pi_2 i$$
where $\pi_1$ is a permutation of $A$ that avoids the order $312$ as a sub-sequence and $\pi_2$ is a permutation of $B$ that avoids the order $312$ as a sub-sequence.
\end{lemma}

\begin{proof}
It is clear that the permutations $\pi_1$ and $\pi_2$ themselves must avoid the order $312$ as a subsequence.

Now, we need to prove that the ordering between $\pi_1$ and $\pi_2$ is necessary. We proceed with a proof by contradiction.

For the sake of contradiction, let us assume that there exists some permutation $\pi$ of $\{1,2,...,k,k+1\}$ that ends with value $i$ such that some integer $x < i$ is to the right of some integer $y > i$. Clearly, this permutation is not of the form described above. It is also easy to see that $\pi$ does \textbf{not} avoid the order $312$ since the triple $(y, x, i)$ satisfies the condition $y > x > i$ and hence is in the order $312$.

\end{proof}

With this lemma proven, we move on to the proof of our theorem.

\begin{proof}
The inductive hypothesis holds for our base case of $\{1\}$, since the only permutation of $\{1\}$ avoids $312$.

Now, we need to prove the inductive case. Let us first assume that for all $i$ from $1$ to $k$, the number of permutations of $\{1,2,...,i\}$ that avoid the order $312$ as a subsequence is $C_i$.

Now, we want to prove the inductive hypothesis for $\{1,2,...,k,k+1\}$ as well, that is the number of permutations of $\{1,2,...,k,k+1\}$ that avoid the order $312$ as a subsequence is $C_{k+1}$.

We count the number of permutations of $\{1,2,...,k,k+1\}$ that avoid $312$ by enumerating through all possible values of the last term of a valid permutation.
If the last term of the permutation is $i$ (where $i \in \{1,2,...,k,k+1\}$), then let us define the subsets $A$ and $B$ of the  set $\{1,2,...,k+1\} \setminus \{i\}$ as the set of integers less than $i$ and the set of integers greater than $i$ respectively. It is clear from the definition of $A$ and $B$ that $A$ and $B$ are disjoint from each other.

Now, from the above lemma, we know that all permutations of $\{1,2,...,k,k+1\}$ ending in $i$ that avoid $312$ must be of the form,

$$\pi = \pi_1 \pi_2 i$$

where $\pi_1$ is a permutation of $A$ that avoids the order $312$ as a sub-sequence and $\pi_2$ is a permutation of $B$ that avoids the order $312$ as a sub-sequence. It is clear that the above permutation contains all integers between $1$ and $k+1$, from the definitions of the subsets $A$ and $B$, which implies that $\pi_1\pi_2i$ is permutation of the set $\{1,2,...,k,k+1\}$.

Now, the total number of permutations $\pi$ is,
$$n_{\pi_1} \cdot n_{\pi_2} = C_{i-1} \cdot C_{k-i+1}$$
since the total number of valid permuations $\pi_1$ is simply going to be $C_{i-1}$ (total number of valid permutations of length $i-1$ that avoid the order $312$ as a sub-sequence is $C_{i-1}$, and correspondingly $n_{\pi_2} = C_{k-1+1}$)

Now, summing over all valid $i$, we see that the total number of permutations of $\{1,2,...,k+1\}$ that avoid $312$ is equal to,
$$\sum_{i=1}^{k+1} C_{i-1} \cdot C_{k-i+1} = \sum_{i=0}^k C_i \cdot C_{k-i}$$ which is in fact $C_{k+1}$, and we are done.

\end{proof}

\end{document}

### 2014-10-23 paper/paper.tex
Added figure for generating tree / updated proof

@@
 \usepackage[T1]{fontenc} 
+\usepackage{tikz}
 
@@
 % Graphics
-\usepackage[dvips]{graphicx,epsfig,color}
-\usepackage{subfigure}
+%\usepackage[dvips]{graphicx,epsfig,color}
+%\usepackage{subfigure}
 %\usepackage{pst-all}
@@
 
+\tikzset
+{
+    treenode/.style = {circle, draw=black, align=center, minimum size=1cm},
+}
+
 \makeatother
@@
 
+\begin{figure}
+\begin{tikzpicture}[level distance=0.5cm, growth parent anchor={south}, nodes={anchor=north}]
+\tikzstyle{hollow node}=[circle,draw,inner sep=0.1mm, text width=1.36cm,align=center]
+\tikzstyle{level 1}=[sibling distance=6cm] 
+\tikzstyle{level 2}=[sibling distance=3cm] 
+\tikzstyle{level 3}=[sibling distance=1.8cm]
+\tikzstyle{level 4}=[sibling distance=0.9cm]
+\node [hollow node]{(1)}
+	child {
+		node[hollow node] {(1,2)}
+		child{
+			node[hollow node]{(1,2,3)}
+			child{node[hollow node]{(1,2,3,4)}}
+			child{node[hollow node]{(1,2,4,3)}}
+			child{node[hollow node]{(1,4,2,3)}}
+			child{node[hollow node]{(4,1,2,3)}}
+		}
+		child{
+			node[hollow node]{(1,3,2)}
+		}
+	}
+    	child {
+		node[hollow node] {(2,1)}
+    		child {
+			node[hollow node] {(2,1,3)}
+			child{node{\vdots}}
+		}
+      		child {
+			node[hollow node] {(2,3,1)}
+			child{node{\vdots}}
+		}
+		child {
+			node[hollow node] {(3,2,1)}
+			child{node{\vdots}}
+		}
+    	};
+
+\end{tikzpicture}
+\caption{Generating tree of the sequence $s_n(312)$} \label{fig:M1}
+\end{figure}
+
 We claim that the sequence of numbers $s_n(312)$ is in fact the sequence of Catalan numbers. We state this result formally as the following theorem,
@@
 \begin{lemma}
-All permutations of $\{1,2,...,k,k+1\}$ ending in $i$ that avoid the order $312$ as a sub-sequence must be of the form,
+All permutations of $\{1,2,\dots,k,k+1\}$ ending in $i$ that avoid the order $312$ as a sub-sequence must be of the form,
 $$\pi_1 \pi_2 i$$
-where $\pi_1$ is a permutation of $A$ that avoids the order $312$ as a sub-sequence and $\pi_2$ is a permutation of $B$ that avoids the order $312$ as a sub-sequence.
+where $\pi_1$ is a permutation of $\{1,2,\ldots,(i-1)\}$ that avoids the order $312$ as a sub-sequence and $\pi_2$ is a permutation of $\{(i+1),\ldots,(k+1)\}$ that avoids the order $312$ as a sub-sequence.
 \end{lemma}
@@
 \begin{proof}
-It is clear that the permutations $\pi_1$ and $\pi_2$ themselves must avoid the order $312$ as a subsequence.
+It is clear that any subsequences of the permutation $\pi = \pi_1 \pi_2 i$ must avoid $312$ if the entire permutation $\pi$ is to avoid $312$ as well; this implies that the permutations $\pi_1$ and $\pi_2$ must avoid $312$ as well.
 
-Now, we need to prove that the ordering between $\pi_1$ and $\pi_2$ is necessary. We proceed with a proof by contradiction.
+We proceed with a proof by contradiction.
 
-For the sake of contradiction, let us assume that there exists some permutation $\pi$ of $\{1,2,...,k,k+1\}$ that ends with value $i$ such that some integer $x < i$ is to the right of some integer $y > i$.
+Before proceeding, we define the sets $A$ and $B$ to be $\{1,2,\ldots,(i-1)\}$ and $\{(i+1),(i+2),\ldots,(k+1)\}$ respectively.
+For the sake of contradiction, let us assume that there exists some permutation $\pi$ of $\{1,2,...,k,k+1\}$ that ends with value $i$ such that some integer $x < i$ (that is, $x \in A$) is to the right of some integer $y > i$ ($y \in B$).
 Clearly, this permutation is not of the form described above.
-It is also easy to see that $\pi$ does \textbf{not} avoid the order $312$ since the triple $(y, x, i)$ satisfies the condition $y > x > i$ and hence is in the order $312$.
+It is also easy to see that $\pi$ does \emph{not} avoid the order $312$ since the triple $(y, x, i)$ satisfies the condition $y > x > i$ and is in the order $312$.
+
+From this we conclude that only permutations of the form described above can avoid $312$.
 
@@
 \begin{proof}
-The inductive hypothesis holds for our base case of $\{1\}$, since the only permutation of $\{1\}$ avoids $312$.
+The inductive hypothesis holds for our base case of $\{1\}$, since the only permutation of $\{1\}$ trivially avoids $312$.
 
@@
 $$n_{\pi_1} \cdot n_{\pi_2} = C_{i-1} \cdot C_{k-i+1}$$
-since the total number of valid permuations $\pi_1$ is simply going to be $C_{i-1}$ (total number of valid permutations of length $i-1$ that avoid the order $312$ as a sub-sequence is $C_{i-1}$, and correspondingly $n_{\pi_2} = C_{k-1+1}$)
+since the total number of valid permuations $\pi_1$ is simply going to be $C_{i-1}$ (total number of valid permutations of length $i-1$ that avoid the order $312$ as a sub-sequence is $C_{i-1}$; similarly $n_{\pi_2} = C_{k-1+1}$)
 
-Now, summing over all valid $i$, we see that the total number of permutations of $\{1,2,...,k+1\}$ that avoid $312$ is equal to,
+Now, summing over all possible values of $i$, we see that the total number of permutations of $\{1,2,...,k+1\}$ that avoid $312$ is equal to,
 $$\sum_{i=1}^{k+1} C_{i-1} \cdot C_{k-i+1} = \sum_{i=0}^k C_i \cdot C_{k-i}$$ which is in fact $C_{k+1}$, and we are done.

### 2014-10-24 code/compute_tn.py
Added code to compute T_n

@@ -1,0 +1,43 @@
+import sys
+
+
+def does_permutation_satisfy_constraint(permutation):
+    prev_odd_ele = -1
+    i = 0
+    for ele in permutation:
+        if ele % 2 == 1:
+            if ele < prev_odd_ele:
+                return False
+            if (ele + 1) not in set(permutation[i+1:]):
+                return False
+            prev_odd_ele = ele
+        i += 1
+    return True
+
+
+def compute_all_permutations(input_list):
+    all_permutations = list()
+    for i in xrange(len(input_list)):
+        ele = input_list[i]
+        new_input_list = input_list[:i] + input_list[i+1:]
+        smaller_list_permutations = compute_all_permutations(new_input_list)
+        if not smaller_list_permutations:
+            all_permutations.append([ele])
+        for smaller_list_permutation in smaller_list_permutations:
+            all_permutations.append([ele] + smaller_list_permutation)
+    return all_permutations
+
+
+def compute_tn(n):
+    all_permutations = compute_all_permutations(range(1, n+1))
+    tot_count = 0
+    for permutation in all_permutations:
+        if does_permutation_satisfy_constraint(permutation):
+            tot_count += 1
+    return tot_count
+
+
+if __name__ == '__main__':
+    n = int(sys.argv[1])
+    for i in xrange(n):
+        print compute_tn(2*i)

### 2014-10-25 code/compute_sn.py
Bug fixes

@@ -18,8 +18,10 @@ def get_subsets(permutation, length):
 def process_subsets(subsets):
     processed_subsets = list()
     for subset in subsets:
-        processed_subset = range(1, len(subset) + 1)
-        processed_subset = sorted(processed_subset, key = lambda x: subset[x-1])
+        mapping = {i : subset[i] for i in xrange(len(subset))}
+        subset_copy = list(subset)
+        subset_copy.sort()
+        processed_subset = [1 + subset_copy.index(mapping[i]) for i in xrange(len(subset))]
         processed_subsets.append(tuple(processed_subset))
     return processed_subsets
 

### 2014-10-25 code/compute_tn.py

@@ -1,6 +1,46 @@
 import sys
 
 
+def get_subsets(permutation, length):
+    if length == 0:
+        return []
+    if len(permutation) == length:
+        return [list(permutation)]
+    subsets = get_subsets(permutation[1:], length)
+    smaller_subsets = get_subsets(permutation[1:], length - 1)
+    if not smaller_subsets:
+        subsets.append([permutation[0]])
+    for smaller_subset in smaller_subsets:
+        subsets.append([permutation[0]] + smaller_subset)
+    return subsets
+
+
+def is_sorted(sequence):
+    for i in xrange(len(sequence) - 1):
+        if sequence[i] > sequence[i+1]:
+            return False
+    return True
+
+
+def process_subsets(subsets):
+    processed_subsets = list()
+    for subset in subsets:
+        mapping = {i : subset[i] for i in xrange(len(subset))}
+        subset_copy = list(subset)
+        subset_copy.sort()
+        processed_subset = [1 + subset_copy.index(mapping[i]) for i in xrange(len(subset))]
+        processed_subsets.append(tuple(processed_subset))
+    return processed_subsets
+
+
+def is_order_in_permutation(permutation, order):
+    all_subsets = get_subsets(permutation, len(order))
+    all_subsets = set(process_subsets(all_subsets))
+    if tuple(order) in all_subsets:
+        return True
+    return False
+
+
 def does_permutation_satisfy_constraint(permutation):
     prev_odd_ele = -1
     i = 0
@@ -28,16 +68,20 @@ def compute_all_permutations(input_list):
     return all_permutations
 
 
-def compute_tn(n):
+def compute_tn(n, order):
     all_permutations = compute_all_permutations(range(1, n+1))
     tot_count = 0
     for permutation in all_permutations:
         if does_permutation_satisfy_constraint(permutation):
-            tot_count += 1
+            if not is_order_in_permutation(permutation, order):
+                tot_count += 1
     return tot_count
 
 
 if __name__ == '__main__':
     n = int(sys.argv[1])
+    order = [1, 3, 2]
     for i in xrange(n):
-        print compute_tn(2*i)
+        if (2*i < len(order)):
+            continue
+        print compute_tn(2*i, order)

### 2014-10-25 paper/paper.tex
Paper additions

@@
 
+\section{Avoidance of permutations of $T_n$}
+We now impose further restrictions on the set of permutations $S_n$. This new set of permutations, $T_n$ is defined follows.
+
+\begin{definition}
+For even $n$, $T_n$ is defined as the set of all permutations $\sigma \in S_n$ for which $1,3,5,\ldots,2n-1$ appear in increasing order, and $2i$ always appears to the right of $2i-1$.
+\end{definition}
+
+Since $T_n$ is only defined for even $n$, henceforth we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $0$.
+
+Given the above definition of $T_{2m}$, we prove the following result about the cardinality of $T_{2m}$.
+
+\begin{theorem}
+$|T_{2m}| = 1 \cdot 3 \cdot 5 \cdot \ldots \cdot (2m-1)$
+\end{theorem}
+
+\begin{proof}
+Observe that since $1,3,5,\ldots,2m-1$ must appear in increasing order in $T_{2m}$, our problem is now reduced to determining the relative order of $2,4,6,\ldots,2m$ with respect to each other, as well as with respect to $1,3,5,\ldots,2m-1$.
+
+Let us first try to insert $2m$ into the sequence $1,3,5,\ldots,2m-1$.
+Clearly, the way $T_{2m}$ is defined, $2m$ can be inserted into only $1$ slot, the one following $2m-1$.
+(Here we define a slot as a gap between two existing elements of the sequence, or the gap that follows the last element of the sequence or the gap that precedes the first element of the sequence)
+
+Now, let us try to insert $(2m-2)$ into the sequence.
+$(2m-2)$ can be inserted into $3$ slots in the sequence -- the one between $2m-3$ and $2m-1$, the one between $2m-1$ and $2m$ or the one following $2m$.
+
+Now, if we try to insert $(2m-4)$ into this incompletely formed sequence, we will see that there are $5$ possible locations into which this number can be inserted.
+Continuing this for all even numbers upto $2$, we observe that the total number of ways such a sequence can  be created is equal to $1 \cdot 3 \cdot 5 \cdot \ldots \cdot (2m-1)$, as desired.
+\end{proof}
+
 \end{document}

### 2014-10-25 code/compute_sn.py
Code cleanup

@@ -18,10 +18,9 @@ def get_subsets(permutation, length):
 def process_subsets(subsets):
     processed_subsets = list()
     for subset in subsets:
-        mapping = {i : subset[i] for i in xrange(len(subset))}
         subset_copy = list(subset)
         subset_copy.sort()
-        processed_subset = [1 + subset_copy.index(mapping[i]) for i in xrange(len(subset))]
+        processed_subset = [1 + subset_copy.index(subset[i]) for i in xrange(len(subset))]
         processed_subsets.append(tuple(processed_subset))
     return processed_subsets
 
@@ -58,7 +57,7 @@ def compute_sn(n, order):
 
 if __name__ == '__main__':
     n = int(sys.argv[1])
-    order = [5, 4, 1, 2, 3]
+    order = [1, 2, 3, 4]
     k = len(order)
     for i in xrange(k, n):
         print compute_sn(i, order)

### 2014-10-25 code/compute_tn.py

@@ -25,10 +25,9 @@ def is_sorted(sequence):
 def process_subsets(subsets):
     processed_subsets = list()
     for subset in subsets:
-        mapping = {i : subset[i] for i in xrange(len(subset))}
         subset_copy = list(subset)
         subset_copy.sort()
-        processed_subset = [1 + subset_copy.index(mapping[i]) for i in xrange(len(subset))]
+        processed_subset = [1 + subset_copy.index(subset[i]) for i in xrange(len(subset))]
         processed_subsets.append(tuple(processed_subset))
     return processed_subsets
 
@@ -80,7 +79,7 @@ def compute_tn(n, order):
 
 if __name__ == '__main__':
     n = int(sys.argv[1])
-    order = [1, 3, 2]
+    order = [3, 1, 2]
     for i in xrange(n):
         if (2*i < len(order)):
             continue

### 2014-10-27 paper/paper.tex
added introduction

@@
 
-\section{Terms in the sequence $s_n(312)$}
+\section{Introduction}
 
-\begin{figure}
+In this paper we will consider investigate pattern avoidance in permutations of finite sets of positive integers.
+
+\begin{definition}
+Two finite sequences $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_n\}$ and $\{b_1,\cdots, b_n\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
+\end{definition}
+
+\begin{definition}
+A finite sequence of numbers $a_1a_2\cdots a_n$ avoids a sequence $b_1\cdots b_k$ with $n\ge k$ if no subsequence $a_{i_1}a_{i_2}\cdots a_{i_k}$ of $a_1a_2\cdots a_n$ has its terms in the same relative order as $b_1\cdots b_k$.
+\end{definition}
+
+\begin{example}
+The sequence 15432 avoids 123 while the sequence 13425 does not. The latter example fails because 134 appears in 15342 with the same relative order as 123. 
+\end{example}
+
+\begin{remark}
+Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$.
+There are two conventions commonly used to write elements of $S_n$.
+The first is $\emph{one-line}$ notation where permutations in $S_n$ are written explicitly, e.g.
+43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.
+
+The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
+For example, the cycle $(14)(23)$ sends $12345$ to $43215$.
+Unless otherwise stated, we will refer to permutations in one-line notation.
+\end{remark}
+
+\begin{definition} 
+When $n\ge k$, a permutation $\sigma\in S_n$ avoids a permutation $\pi\in S_k$ if $pi$ and $\sigma$ avoid each other when written as sequences in one-line notation. 
+\end{definition}
+
+\begin{definition}
+For a permutation $\pi$, we write $s_n(\pi$ for the number of permutations in $S_n$ that avoid $\pi$. 
+\end{definition}
+
+In this paper we concern ourselves with the sequences $s_n(\pi)$ for various $\pi$.
+In section \ref{S3} we will consider sequence $s_n(\pi)$ for $\pi\in S_3$, and in section \ref{S4} we will consider sequence $s_n(\pi)$ for $\pi\in S_4$.
+In section \ref{Tn} we consider a variation on the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of $S_n$.
+
+\subsection{Generating Trees}
+
+Before continuing we introduce generating trees, a tool we can use to study pattern avoidance.
+The generating tree for the sequence $s_n(\pi)$ is the infinite tree whose vertices are elements of $S_n$ that avoid $\pi$.
+To construct the tree, we define the children of an arbitrary node.
+Let $\omega\in S_k$ be a node of the generating tree of $s_n(\pi)$.
+Then the children of $\omega$ are elements of $S_{k+1}$ that avoid $\pi$ and that can be obtained by inserting the integer $k+1$ between integers in $\omega$.
+Figure \ref{fig:M1} is an example of the generating tree of $s_n(312)$.
+
+\begin{figure}[h!]
 \begin{tikzpicture}[level distance=0.5cm, growth parent anchor={south}, nodes={anchor=north}]
@@
 \end{figure}
+
+\section{Terms in the sequence $s_n(312)$}
+\label{S3}
 
@@
 
+\section{Conjectures on $s_n(\pi)$ for $\pi\in S_4$}
+\label{S4}
+
 \section{Avoidance of permutations of $T_n$}
+\label{Tn}
 We now impose further restrictions on the set of permutations $S_n$. This new set of permutations, $T_n$ is defined follows.

### 2014-10-27 paper/paper.tex
added the Young tableaux stuff

@@
 % Special symbols, etc.
-\usepackage{amssymb,amsbsy,latexsym}
+\usepackage{amssymb,amsbsy,latexsym,ytableau}
 \usepackage{amsmath}
@@
 
+\section{Terms in the sequence $s_n(321)$}
+
+Unfortunately, there is no clear way to reformulate Lemma 2 for the case of avoiding the permutation 321 or 123.
+As a consequence, the proof in the previous section simply does not hold.
+Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping principle).
+Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number.
+In order to prove this result, however, we will need the machinery of Young tableaux.
+What follows is a brief exposition of the key results, adapted from Stanley's \emph{Algebraic Combinatorics}.
+
+\begin{definition}
+A Young diagram of a partition $\lambda=\{\lambda_1, \lambda_2, \ldots, \lambda_n\}$ of the integer $\sum_{i=1}^n \lambda_i$ is a left-justified array of squares, with $\lambda_i$ squares in the $i$th row.
+\end{definition}
+\begin{example}
+The Young diagram of $(4, 3, 1, 1)$ looks like:
+\[
+\ytableausetup{centertableaux}
+\begin{ytableau}
+\ & \ & \ & \ \\
+\ & \ & \ \\
+\ \\
+\
+\end{ytableau}
+\]
+\end{example}
+
+\begin{definition}
+A standard Young tableau (or SYT) consists of the Young diagram $D$ of some partition $\lambda$ of an integer $n$, together with the numbers $1, 2, \ldots, n$ inserted into the squares of $D$, so that each number appears exactly once, and every row and column is \emph{increasing}.
+We call $\lambda$ the \emph{shape} of the SYT.
+\end{definition}
+
+\begin{example}
+There are five SYT of the shape $(2, 2, 1)$. They are given by
+\[\ytableausetup{aligntableaux=top}
+\begin{ytableau}
+1 & 2 \\
+3 & 4\\
+5
+\end{ytableau}\ \ \ \ \
+\hfill
+\begin{ytableau}
+1 & 2 \\
+3 & 5\\
+4
+\end{ytableau}\ \ \ \ \
+\hfill
+\begin{ytableau}
+1 & 3 \\
+2 & 4\\
+5
+\end{ytableau}\ \ \ \ \
+\hfill
+\begin{ytableau}
+1 & 3 \\
+2 & 5\\
+4
+\end{ytableau}\ \ \ \ \
+\hfill
+\begin{ytableau}
+1 & 4 \\
+2 & 5\\
+3
+\end{ytableau}.\]
+\end{example}
+
+\begin{definition}
+Let $u$ be a square of the Young diagram of the partition $\lambda$.
+Then the hook $H(u)$ is the set of all squares directly to the right of $u$ or directly below $u$, including $u$ itself.
+The size of $H(u)$ is called the hook length of $u$, and is denoted $h(u)$.
+\end{definition}
+
+\begin{example}
+In the Young diagram of the partition $(4, 2, 2)$ below, each square $u$ contains its hook length $h(u)$.
+\[\ytableausetup{centertableaux}
+\begin{ytableau}
+6 & 5 & 2 & 1\\
+3 & 2\\
+2 & 1
+\end{ytableau}\]
+\end{example}
+
+\begin{theorem}
+Let $\lambda$ be a partition of $n$. Then $f^\lambda$, the number of SYT of shape $\lambda$, is given by
+\[f^\lambda=\frac{n!}{\prod_{u\in \lambda} h(u)},\]
+where the notation $u\in \lambda$ means that $u$ ranges over all squares of the Young diagram of $\lambda$.
+\end{theorem}
+
+The theorem above is called the \emph{hook-length formula}.
+
+\begin{example}
+The diagram above for the hook lengths of $\lambda=(4, 2, 2)$ tells us that the number of SYT of shape $\lambda$ is given by
+\[\frac{8!}{6\cdot 5\cdot 2 \cdot 1 \cdot 3 \cdot 2 \cdot 2\cdot 1}=56.\]
+\end{example}
+
+It is an easy application of the hook-length formula to prove the following theorem:
+
+\begin{theorem}
+The number of SYT of shape $(n, n)$ is given by $C_n$, the $n^{th}$ Catalan number.
+\end{theorem}
+
+We need only one more result before we are ready to prove our central result.
+
+\begin{theorem}
+There exists an algorithm (called the \emph{RSK algorithm}) which defines a bijection between the set of permutations in $S(n)$ avoiding a decreasing subsequence of length 3 and the set of pairs of size-$n$ SYT of the same shape and at most two rows.
+\end{theorem}
+Using all this machinery and all these results, we are finally able to prove the closed-form expression for $s_n(321)$.
+\begin{theorem}
+The total number of permutations of $\{1,2,3,\ldots, n\}$ that avoid the order 321 as a subsequence is $C_n$, where $C_n$ is the $n^{th}$ Catalan number.
+\end{theorem}
+\begin{proof}
+By the first part of Theorem 12, there are $s_n(321)$ elements in the set $A$ of pairs of size-$n$ SYT of the same shape and at most two rows.
+By Theorem 11, the set $B_n$ of SYT of shape $(n, n)$ has its size given by $C_n$.
+We will construct a bijection between the sets $A_n$ and $B_n$: because $|A_n|=s_n(321)$ and $|B_n|=C_n$, it will therefore follow that $s_n(321)=C_n$, as desired.
+
+Given any pair of size $n$ Young tableaux of the same shape and at most two parts, take the second tableau and invert its numbers: that is, send $i$ to $2n+1-i$ for all $i$.
+Now rotate the Young tableau and `fit' it into the first tableau.
+We demonstrate this process for a pair of Young tableaux with $n=6$: we begin with
+\[\ytableausetup{aligntableaux=top}
+\begin{ytableau}
+1 & 2 &3&4\\
+5 & 6
+\end{ytableau} \ \ \ \ \ 
+\hfill
+\begin{ytableau}
+1 & 3 & 4 & 5\\
+2 & 6
+\end{ytableau}.\]
+Inverting the numbers of the second tableau gives us
+\[\ytableausetup{aligntableaux=top}
+\begin{ytableau}
+1 & 2 &3&4\\
+5 & 6
+\end{ytableau} \ \ \ \ \ 
+\hfill
+\begin{ytableau}
+12 & 10 & 9 & 8\\
+11 & 7
+\end{ytableau}.\]
+Finally, rotating the second Young tableau and fitting it into the first gives us
+\[\ytableausetup{aligntableaux=top}
+\begin{ytableau}
+1 & 2 &3&4&7&11\\
+5 & 6&8&9&10&12
+\end{ytableau},\]
+a valid Young tableau of shape $(6, 6)$.
+
+Beginning with a pair of size $n$ Young tableaux of shape $\lambda$ and at most two parts, it is easy to see that this process always yields a valid Young tableau of shape $(n, n)$.
+Indeed, we can also define the inverse: given any Young tableau for the partition $(n, n)$, we can find the shape formed by all numbers greater than $n$: split it off from the original tableau, rotate it, and invert the numbers to get a pair of size $n$ Young tableau of the same shape and at most two parts.
+These processes can be easily checked to be inverses: hence the two sets are of the same size, and the proof follows.
+\end{proof}
+
 \section{Conjectures on $s_n(\pi)$ for $\pi\in S_4$}

### 2014-10-27 paper/paper.tex
S4 conjectures

--- 2014-10-27 paper/paper.tex
@@
 
+\subsection{Flipping}
+
 \section{Terms in the sequence $s_n(312)$}
@@
 
+In this section we will present our findings and conjectures about sequences $s_n(\pi)$ for $\pi\in S_4$.
+We computed values up to $s_{10}$.
+It appears that there are three sequences that appear, and we have sorted the permutations according to which sequence they produce and paired them with their flips.
+
+We define the following sequences based on the first 8 terms:
+$$
+A:=1,2,6,23,103,512,2740,15485,91245\ldots
+$$
+$$
+B:=1,2,6,23,103,513,2761,15767,94359\ldots
+$$
+$$
+C:=1,2,6,23,103,513,2762,15793,94776\ldots
+$$
+
+The table below lists the elements of $S_4$ generate sequences $A,B,C$ for the first 8 terms.
+
+\begin{center}
+\begin{tabular}{|c|c|c|}
+B &A&C\\
+\hline
+1234, 4321&4132, 1423&4231,1324\\
+1243, 4312&4213, 1342&\\
+1432, 4123&2431,3124&\\
+2134, 3421&2413, 3142&\\
+2143, 3412&2314, 3241&\\
+2341, 3214&&\\
+\end{tabular}
+\end{center}
+
+There is no guarantee that these sequences will not diverge and produce more than three distinct sequences, however for now we hypothesize that $A$, $B$, and $C$ are the only sequences that arise for $s_n(\pi)$ with $\pi\in S_4$.
+Assuming this hypothesis, if we write each length four permutation in cycle notation, a pattern begins to emerge.
+
+\begin{center}
+\begin{tabular}{|c|c|c|}
+B &A&C\\
+\hline
+(1),(14)(23)&(243),(142)&(23),(14)\\
+(34),(1423)&(234),(143)&\\
+(24),(1432)&(124),(132)&\\
+(12),(1324)&(123),(134)&\\
+(12)(34),(13)(24)&(1243),(1342)&\\
+(1234),(13)&&\\
+\end{tabular}
+\end{center}
+
+So we conjecture that sequence $A$ is characterized by elements of $S_4$ that are three cycles or four cycles when written in cycle notation.
+Sequence $B$ seems to be characterized by either a pair of disjoint two cycles paired with a flip that is a 2-cycle, or a single 4-cycle paired with a flip that is a 2-cycle.
+Sequence $C$ appears to be characterized by a 2-cycle paired with a 2-cycle as a flip.
+
 \section{Avoidance of permutations of $T_n$}

### 2014-10-27 paper/paper.tex

--- 2014-10-27 paper/paper.tex
@@
 \end{figure}
+
+\subsection{Flipping}
 

### 2014-10-27 paper/paper.tex
Paper improvements

@@
 
-In this paper we will consider investigate pattern avoidance in permutations of finite sets of positive integers.
+In this paper we will investigate pattern avoidance in permutations of finite sets of positive integers.
+
+Before, beginning our discussion, we present the following definitions.
 
 \begin{definition}
-Two finite sequences $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_n\}$ and $\{b_1,\cdots, b_n\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
+Two finite sequences $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
 \end{definition}
@@
 \begin{definition} 
-When $n\ge k$, a permutation $\sigma\in S_n$ avoids a permutation $\pi\in S_k$ if $pi$ and $\sigma$ avoid each other when written as sequences in one-line notation. 
+When $n\ge k$, a permutation $\sigma\in S_n$ avoids a permutation $\pi\in S_k$ if $\pi$ and $\sigma$ avoid each other when written as sequences in one-line notation. 
 \end{definition}
@@
 \begin{definition}
-For a permutation $\pi$, we write $s_n(\pi$ for the number of permutations in $S_n$ that avoid $\pi$. 
+For a permutation $\pi$, $s_n(\pi)$ represents the number of permutations in $S_n$ that avoid $\pi$. 
 \end{definition}
 
-In this paper we concern ourselves with the sequences $s_n(\pi)$ for various $\pi$.
-In section \ref{S3} we will consider sequence $s_n(\pi)$ for $\pi\in S_3$, and in section \ref{S4} we will consider sequence $s_n(\pi)$ for $\pi\in S_4$.
+In this paper we concern ourselves with the sequence of numbers $s_n(\pi)$ for various $\pi$.
+In section \ref{S3} we will consider sequence $s_n(\pi)$ for $\pi\in S_3$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
+In section \ref{S4} we will consider sequence $s_n(\pi)$ for $\pi\in S_4$.
 In section \ref{Tn} we consider a variation on the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of $S_n$.
@@
 \subsection{Flipping}
+Another concept that we will use in this paper is the concept of flipping.
+We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with elements in the opposite order.
+More concretely, if $a = a_1, a_2, \dots, a_n$ for some arbitrary $n$, then the flip of the sequence $a$ will be the sequence $b = a_n, a_{n-1}, \dots, a_2, a_1$.
+
+We will talk about how the flip of a sequence is important to this avoidance problem later in the paper.
 
@@
 
-From this we conclude that only permutations of the form described above can avoid $312$.
+From this we conclude that all $x < i$ must be to the left of all $y > i$.
+This information coupled with the fact that we are only considering permutations that end with $i$ complete the claim that only permutations of the form described above can avoid $312$.
 
@@
 \end{proof}
+
+It is easy to see that the above proof can be modified slightly to prove that the sequences $s_n(132)$, $s_n(213)$ and $s_n(231)$ are in fact the sequence of catalan numbers as well.
+In fact, we claim that if $\sigma$ and $\pi$ are both permutations in $S_k$ for some arbitrary $k$, then if $\sigma = flip(\pi)$, $s_n(\pi) = s_n(\sigma)$.
+This is easy to see since if $a$ is a permutation that avoids $\sigma$, then $flip(a)$ is a permutation that avoids $\pi$ -- since this can be done both ways, it is bijective and we can conclude that the cardinalities of the two sets are equal, or $s_n(\sigma) = s_n(\pi)$.
 
@@
 
+Studying avoidance of permutations in $T_n$ seems an interesting exercise.
+In the following section, we consider the sequence of numbers $t_n(\pi)$, defined as the number of permutations in $T_n$ that avoid the permutation $\pi$.
+In particular, we consider permutations $\pi \in S_3$.
+
+Before stating our conjectures, we present the computed sequences. Note that since the sequence $T_n$ is defined only for even $n$, we only present the even terms of the sequence $t_n$.
+
+$$t_n(123) = 1, 0, 0, 0, 0, \ldots$$
+$$t_n(132) = 1, 1, 1, 1, 1, \ldots$$
+$$t_n(213) = 1, 2, 4, 8, 16, \ldots$$
+$$t_n(231) = 1, 2, 4, 8, 16, \ldots$$
+$$t_n(312) = 1, 3, 12, 55, 273, \ldots$$
+$$t_n(321) = 1, 3, 12, 55, 273, \ldots$$
+
+It is easy to see why $t_n(123)$ and $t_n(132)$ behave the way they do -- since all odd numbers in $T_n$ are increasing, it is impossible to have a permutation in $T_n$ that avoids $123$ and the permutation consisting of all increasing integers is the only permutation of $T_n$ that avoids $132$.
+
+More interesting is the equivalences between $213$ and $231$, and $312$ and $321$. Proving these equivalences remains future work.
+
 \end{document}

### 2014-10-28 paper/paper.tex
made a few changes

@@
 
-Before, beginning our discussion, we present the following definitions.
+Before beginning our discussion, we present the following definitions.
 
@@
 \begin{definition} 
-When $n\ge k$, a permutation $\sigma\in S_n$ avoids a permutation $\pi\in S_k$ if $\pi$ and $\sigma$ avoid each other when written as sequences in one-line notation. 
+A permutation $\sigma\in S_n$ avoids a permutation $\pi\in S_k$ if the sequence of numbers for the one-line notation for $\sigma$ avoids the sequence of numbers for the one-line notation for $\pi$.
 \end{definition}
@@
 \begin{definition}
-For a permutation $\pi$, $s_n(\pi)$ represents the number of permutations in $S_n$ that avoid $\pi$. 
+For a permutation $\pi\in S_k$, $s_n(\pi)$ represents the number of permutations in $S_n$ that avoid $\pi$. 
 \end{definition}
@@
 \subsection{Flipping}
-Another concept that we will use in this paper is the concept of flipping.
+Another notion that we will use in this paper is the concept of flipping.
 We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with elements in the opposite order.

### 2014-10-28 paper/paper.tex
changed flipping definition

@@
 \subsection{Flipping}
-Another notion that we will use in this paper is the concept of flipping.
-We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with elements in the opposite order.
-More concretely, if $a = a_1, a_2, \dots, a_n$ for some arbitrary $n$, then the flip of the sequence $a$ will be the sequence $b = a_n, a_{n-1}, \dots, a_2, a_1$.
-
-We will talk about how the flip of a sequence is important to this avoidance problem later in the paper.
+Another notion that we will use in this paper is the concept of `flipping'.
+We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with the largest element swapped with the smallest element, the second largest element swapped with the second smallest element, etc.
+For example, the flip of 1324 is 4321.
+We will discuss the flip as it becomes useful.
 

### 2014-11-01 paper/paper.tex
Added t_n proof to paper

--- 2014-10-27 paper/paper.tex
@@
 \tikzstyle{level 4}=[sibling distance=0.9cm]
-\node [hollow node]{(1)}
+\node [hollow node]{1}
 	child {
-		node[hollow node] {(1,2)}
+		node[hollow node] {12}
 		child{
-			node[hollow node]{(1,2,3)}
-			child{node[hollow node]{(1,2,3,4)}}
-			child{node[hollow node]{(1,2,4,3)}}
-			child{node[hollow node]{(1,4,2,3)}}
-			child{node[hollow node]{(4,1,2,3)}}
+			node[hollow node]{123}
+			child{node[hollow node]{1234}}
+			child{node[hollow node]{1243}}
 		}
 		child{
-			node[hollow node]{(1,3,2)}
+			node[hollow node]{132}
+			child{node{\vdots}}
 		}
@@
     	child {
-		node[hollow node] {(2,1)}
+		node[hollow node] {21}
     		child {
-			node[hollow node] {(2,1,3)}
+			node[hollow node] {213}
 			child{node{\vdots}}
@@
       		child {
-			node[hollow node] {(2,3,1)}
+			node[hollow node] {231}
 			child{node{\vdots}}
@@
 		child {
-			node[hollow node] {(3,2,1)}
+			node[hollow node] {321}
 			child{node{\vdots}}
@@
 \begin{lemma}
+\label{lemma1}
 All permutations of $\{1,2,\dots,k,k+1\}$ ending in $i$ that avoid the order $312$ as a sub-sequence must be of the form,
@@
 
-Unfortunately, there is no clear way to reformulate Lemma 2 for the case of avoiding the permutation 321 or 123.
+Unfortunately, there is no clear way to reformulate Lemma \ref{lemma1} for the case of avoiding the permutation 321 or 123.
 As a consequence, the proof in the previous section simply does not hold.
@@
 
-More interesting is the equivalences between $213$ and $231$, and $312$ and $321$. Proving these equivalences remains future work.
+More interesting is the equivalences between $213$ and $231$, and $312$ and $321$. Before proving these equivalences, we state and prove the following lemmas which will be useful in our proof.
+
+\begin{lemma}
+\label{tn_avoidance}
+If $a$ and $b$ are two elements that appear in a permutation in $T_n$ such that $a$ is to the left of $b$, then if $a > b$, $b$ must be even.
+\end{lemma}
+
+\begin{proof}
+For the sake of contradiction, let us assume that $b$ is odd. Then there exist two possibilities that need to be considered -- one, that $a$ is odd, and two, that $a$ is even.
+
+\begin{itemize}
+\item \textbf{Case 1:} $a$ is odd.
+
+Note that all permutations in $T_n$ must have odd elements monotonically increasing.
+So if $a$ appears to the left of $b$, and both $a$ and $b$ are odd, then $a$ must be less than $b$ -- a contradiction.
+Hence, we conclude that $a$ can't be odd.
+
+\item \textbf{Case 2:} $a$ is even.
+
+For all permutations $\sigma$ in $T_n$, every even element $j$ must be to the right of all odd elements less than $j$.
+From this, we conclude that since $a$ is even, it must be to the right of all odd numbers less than it.
+But $b$ is an odd number less than $a$ -- we conclude that $a$ can't be even either.
+\end{itemize}
+
+Since $a$ can't be odd or even if $b$ is odd, we conclude that $b$ must be even, as desired.
+\end{proof}
+
+\begin{lemma}
+\label{swapping_lemma}
+If $x, y, z$ is a three-element subsequence of a permutation $\pi$ in $T_n$ such that $x > y > z$ or $x > z > y$, then swapping $y$ and $z$ in the permutation $\pi$ yields a permutation $\pi'$ that also belongs to $T_n$.
+\end{lemma}
+
+\begin{proof}
+From Lemma \ref{tn_avoidance} we see that both $y$ and $z$ must be even. Then there exist two cases that need to be covered.
+\begin{itemize}
+\item \textbf{Case 1:} $x$ is odd.
+
+If $x$ is odd, then it is clear that the permutation $\pi'$ obtained from swapping $y$ and $z$ is still valid, since both $y$ and $z$ are to the right of the odd numbers $y-1$ and $z-1$ (since $y-1$ and $z-1$ are themselves to the left of $x$ since they are strictly less than $x$)
+
+\item \textbf{Case 2:} $x$ is even.
+
+If $x$ is even, then $x-1$ which is an odd number greater than both $y-1$ and $z-1$, must be to the left of $x$.
+Since $y-1$ and $z-1$ are already to the left of $x-1$, we conclude that swapping $y$ and $z$ still ensures that $y$ and $z$ are both to the right of $y-1$ and $z-1$.
+\end{itemize}
+Hence, we see that regardless of whether $x$ is even or odd, swapping $y$ and $z$ in a permutation $\pi \in T_n$ if $x > y > z$ or $x > z > y$ yields a permutation $\pi'$ that also belongs to $T_n$.
+\end{proof}
+
+Given these lemmas, we now present a proof for $t_n(312) = t_n(321)$.
+
+\begin{theorem}
+$t_n(312) = t_n(321)$
+\end{theorem}
+
+\begin{proof}
+First observe from Lemma \ref{tn_avoidance} that if $x, y, z$ is a subsequence of three elements of a permutation in $T_n$ in the relative order $312$ or $321$, then both $y$ and $z$ must be even.
+
+We now construct a bijective mapping between all permutations in $T_n$ that avoid $312$ and all permutations in $T_n$ that avoid $321$.
+
+Consider a permutation $\pi$ in $T_n$ that avoids $312$.
+For each permutation $\pi \in T_n$ that avoids $312$, we want to show that there exists a sequence of transformation moves that transforms $\pi$ to a unique permutation $\pi'$ that avoids $321$.
+
+If the original permutation $\pi$ avoids $321$ already, then clearly $\pi' = \pi$ avoids $321$.
+
+Now if $\pi$ does not avoid $321$, there must exist some sub-sequence $x, y, z$ in $\pi$ in the relative ordering $321$.
+From Lemma \ref{swapping_lemma}, we see that swapping $y$ and $z$ in the permutation $\pi$ yields another valid permutation in $T_n$ -- note that the permutation produced after this swap has $x, y, z$ in the relative ordering $312$ instead of $321$.
+If we do this for every sub-sequence in the relative ordering $321$, we see that we will produce a permutation in $T_n$ that avoids $321$.
+This process is completely invertible, since we can take a permutation $\pi'$ in $T_n$ that avoids $321$, and similarly swap all elements that violates $\pi'$'s avoidance of $321$ to obtain a permutation that avoids $312$.
+\textbf{Make this more rigorous}
+
+Since there exists a bijective mapping between permutations in $T_n$ that avoid $312$ and permutations in $T_n$ that avoid $321$, we can conclude that $t_n(312) = t_n(321)$.
+
+\end{proof}
 

### 2014-11-01 code/compute_tn.py

@@ -73,6 +73,7 @@ def compute_tn(n, order):
     for permutation in all_permutations:
         if does_permutation_satisfy_constraint(permutation):
             if not is_order_in_permutation(permutation, order):
+                print " ".join([str(ele) for ele in permutation])
                 tot_count += 1
     return tot_count
 

### 2014-11-01 paper/paper.tex

@@
 \subsection{Flipping}
-Another concept that we will use in this paper is the concept of flipping.
-We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with elements in the opposite order.
-More concretely, if $a = a_1, a_2, \dots, a_n$ for some arbitrary $n$, then the flip of the sequence $a$ will be the sequence $b = a_n, a_{n-1}, \dots, a_2, a_1$.
-
-We will talk about how the flip of a sequence is important to this avoidance problem later in the paper.
+Another notion that we will use in this paper is the concept of `flipping'.
+We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with the largest element swapped with the smallest element, the second largest element swapped with the second smallest element, etc.
+For example, the flip of 1324 is 4321.
+We will discuss the flip as it becomes useful.
 

### 2014-11-02 paper/paper.tex
Some more changes to paper
More edits to paper -- talked a lit more about how bijection is established

@@
 
-\title{18.821 Project 2}
+\title{Avoidance in permutations}
 
@@
 
-In this paper we will investigate pattern avoidance in permutations of finite sets of positive integers.
+In this paper we will investigate pattern avoidance in permutations of finite sets of the first $n$ positive integers for arbitrary $n$.
 
@@
 
-More interesting is the equivalences between $213$ and $231$, and $312$ and $321$. Before proving these equivalences, we state and prove the following lemmas which will be useful in our proof.
+More interesting is the equivalences between $213$ and $231$, and $312$ and $321$.
+We will prove the equivalence between $312$ and $321$ below, but before doing so, we first state and prove the following lemmas which will be useful in our proof.
 
@@
 
+\textbf{May want to skip the transformation stuff, and directly talk about the bijection being established.}
+
 If the original permutation $\pi$ avoids $321$ already, then clearly $\pi' = \pi$ avoids $321$.
@@
 Now if $\pi$ does not avoid $321$, there must exist some sub-sequence $x, y, z$ in $\pi$ in the relative ordering $321$.
-From Lemma \ref{swapping_lemma}, we see that swapping $y$ and $z$ in the permutation $\pi$ yields another valid permutation in $T_n$ -- note that the permutation produced after this swap has $x, y, z$ in the relative ordering $312$ instead of $321$.
-If we do this for every sub-sequence in the relative ordering $321$, we see that we will produce a permutation in $T_n$ that avoids $321$.
-This process is completely invertible, since we can take a permutation $\pi'$ in $T_n$ that avoids $321$, and similarly swap all elements that violates $\pi'$'s avoidance of $321$ to obtain a permutation that avoids $312$.
-\textbf{Make this more rigorous}
+From Lemma \ref{swapping_lemma}, we see that swapping $y$ and $z$ in the permutation $\pi$ yields another valid permutation $\pi_1$ in $T_n$ -- note that the permutation produced after this swap has $x, y, z$ in the relative ordering $312$ instead of $321$.
+Note that $\pi_1$ may still have sub-sequences in the relative ordering $321$, so we may have to perform the above swapping operation again to get some permutation $\pi_2 \in T_n$ such that a sub-sequence $x_1, y_1, z_1$ in $\pi_1$ that has a relative ordering $321$ is in the relative ordering $312$ in $\pi_2$.
+Continuing this process produces the following chain of permutations,
+$$\pi \rightarrow \pi_1 \rightarrow \pi_2 \rightarrow \ldots \rightarrow \pi_k$$
+Where the last permutation in the chain $\pi_k$ does not have a single three-element sub-sequence in the relative ordering $321$, that is $\pi' = \pi_k$ is a permutation in $T_n$ that avoids $321$.
+
+This process is completely invertible, since we can take a permutation $\pi'$ in $T_n$ that avoids $321$, and traverse through the above chain in reverse order to obtain a permutation that avoids $312$.
+
+What remains to be proven is that the order of swapping operations in the chain of transformations does not matter -- that is, any valid transformation chain starting with some permutation $\pi \in T_n$ that avoids $312$, will always produce the same permutation $\pi' \in T_n$ that avoids $321$.
+Note that since we're only swapping even elements, the odd elements are in exactly the same position in both $\pi$ and $\pi'$.
+In addition, for every element $i \in \pi$, all even numbers in $\pi$ to the right of $i$ and less than $i$ must be in decreasing order; in $\pi'$, for every element $i \in \pi'$, all even numbers in $\pi'$ to the right of $i$ and less than $i$ must be in increasing order.
 
@@
 
+Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above, partially due to the fact that the element corresponding to $2$ and $3$ in the above sub-sequences can be either even or odd, which means we now also have to account for the positions of odd elements too.
+Proving this equivalence remains future work.
+
 \end{document}

### 2014-11-02 project1/paper/paper.tex
Added project1 stuff to this repo

\documentclass[11pt,letterpaper,twoside,english]{article}

\usepackage[margin=1.4in]{geometry} % controls the size of the margins

% Special symbols, etc.
\usepackage{amssymb,amsbsy,latexsym}
\usepackage{amsmath}
\usepackage{graphics, subfigure, float} 
\usepackage{cancel}
%\usepackage{todonotes}
% Encoding settings
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
\usepackage[T1]{fontenc}

\usepackage{titling} % allows posttitle command

% AMS Math packages

\usepackage{amscd,amsthm}

\usepackage{verbatim, comment} % can comment out text 
\usepackage{mdwlist}

% Graphics
\usepackage[dvips]{graphicx,epsfig,color}
\usepackage{subfigure}
%\usepackage{pst-all}
%\usepackage{pstricks-add}
\usepackage{hyperref}  % can only be used with pdflatex - gives hyperlinks
\usepackage{bm} % bold math font
\usepackage{bbm}
\usepackage{todonotes}

\newtheoremstyle{theorem}{1em}{1em}{\slshape}{0pt}{\bfseries}{.}{ }{}
\theoremstyle{theorem}
\newtheorem{theorem}{Theorem}
\newtheorem*{theorem*}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{definition}[theorem]{Definition}
\newtheorem*{claim*}{Claim}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem*{remark*}{Remark}
\newtheorem{algorithm}{Algorithm}
\newtheorem*{question*}{Question}
\newtheorem{question}{Question}
\newtheorem{example}{Example}

\providecommand{\setN}{\mathbb{N}}
\providecommand{\setZ}{\mathbb{Z}}
\providecommand{\setQ}{\mathbb{Q}}
\providecommand{\setR}{\mathbb{R}}
\providecommand{\E}{\mathrm{E}}
\providecommand{\Pr}{\mathrm{Pr}}
\providecommand{\Var}{\mathrm{Var}}

\makeatother

\title{Greedy sequences of integers}

\author{Yajit Jain, Deepak Narayanan, Leon Zhang}

\begin{document}

\maketitle

\section{Introduction} \label{sec:intro}

In this paper, we look at the sequence of integers $A=(a_i)_{i\in\mathbb{Z}_{\ge 0}}$ defined inductively as follows.

\begin{definition} \label{def:seqA}
The sequence $A$ is defined as the sequence of numbers $a_0, a_1, a_2, \ldots$ such that $a_0 = 0$, and all subsequent terms $a_{n}$ are defined as the least term greater than $a_{n-1}$ such that no three distinct terms in the set $\{a_0, a_1, a_2, \ldots, a_n\}$ form a $3$-term arithmetic progression.
\end{definition}

It turns out that this sequence $A$ has a number of extremely interesting properties. We state and prove some of these conjectures below.

Before we begin, consider the beginning of the sequence $A$. All numbers less than 40 that are not in the sequence have been crossed out.

\

\noindent 0, 
\\1, \xcancel2,
\\3, 4, \xcancel5, \xcancel6, \xcancel7, \xcancel8,
\\9, 10, \xcancel{11}, 12, 13, \xcancel{14}, \xcancel{15}, \xcancel{16}, \xcancel{17}, \xcancel{18}, \xcancel{19}, \xcancel{20}, \xcancel{21}, \xcancel{22}, \xcancel{23}, \xcancel{24}, \xcancel{25}, \xcancel{26},
\\27, 28, \xcancel{29}, 30, 31, \xcancel{32}, \xcancel{33}, \xcancel{34}, \xcancel{35}, 36, 37, \xcancel{38}, 39, 40

 \

The terms above are formatted to draw attention to a very specific pattern: all powers of $3$ less than $a_{15}=40$ appear in the sequence. This observation is at the heart of many of the results we shall prove about this sequence and its variants. The main result regarding the sequence $A$ is stated below and proven in Section \ref{sec:AP}. In what follows $(\cdot)_3$ will denote the base three expansion, and $(\cdot)_2$ will denote the base two expansion.

\

 {\itshape If the binary representation of $n$ is $(b_kb_{k-1}\ldots b_0)_2$, where $b_0, b_1, b_2, \ldots, b_k \in \{0, 1\}$, then $a_n = (b_k b_{k-1}\ldots b_0)_3.$}

\

In Sections \ref{sec:GP}, \ref{sec:firstTerm}, and \ref{sec:diffEquation}, we consider variations of this sequence $A$. First we consider the problem of avoiding geometric sequences of length 3. We will show that the problem of avoiding 3-term geometric sequences is very closely related to the problem of avoiding 3-term arithmetic sequences, and the closed form of the sequence $A$ can be used to characterize the geometric progression avoiding sequence $G$.

In Section \ref{sec:firstTerm}, we will show that the behavior of the sequence $A$ is heavily dependent on the initial conditions of the sequence. In particular, if we let $a_0=0$ and $a_1=m$, the sequence is significantly altered.

Finally, we note that avoiding three-term arithmetic progressions is equivalent to the problem of avoiding triplets $x,y,z\in S$ that are solutions to the equation $x+z=2y$. In Section \ref{sec:diffEquation}, we consider the problem of avoiding solutions to the equation $x+z =y$.

\emph{Yajit wrote and edited this section. Leon made close edits to this section. Deepak edited this section.}

\section{Avoiding three-term arithmetic progressions} \label{sec:AP}

As motivation, notice that the number preceding any number of the form $3^n$ in $A$ is exactly $\frac{3^n-1}{2}$. This can be factored as 
\[\frac{3^n-1}{2}=\frac{(3-1)(3^{n-1}+3^{n-2}+\ldots+3+1)}{(3-1)}=3^{n-1}+3^{n-2}+\ldots+3+1,\]
which is the sum of all smaller powers of $3$.

Also notice that if we consider the entries of each row modulo the power of $3$ at the beginning of the row, the sequence appears to regenerate itself:

\

\noindent 0, 
\\0, 
\\0, 1, 
\\0, 1, 3, 4, 
\\ 0, 1, 3, 4, 9, 10, 12, 13,
\\0, 1, 3, 4, 9, 10, 12, 13, 27, 28, 30, 31, 36, 37, 39, 40

 \

This recurrence is the motivating example for our first claim. Informally stated, this claim is that the base $3$ representation of each element in the sequence $A$ contains only $0$'s and $1$'s. Once this claim is proven, we can conveniently write a closed form expression for each $a_n\in A$.

%As we observed in the introduction, the sequence $A$ seems to contain all powers of $3$. Furthermore, the preceding term to a power of $3$ seems to be the sum of all smaller powers of 3. Our main result of this section implies that this is true, and more generally characterizes all terms of the sequence.

\begin{definition} \label{def:S_A}
$S_A$ is defined to be the set of integers greater than or equal to zero, of the form $\alpha_k \cdot 3^k + \alpha_{k-1} \cdot 3^{k-1} + \alpha_{k-2} \cdot 3^{k-2} + \ldots + \alpha_1 \cdot 3 + b_0$ where $\alpha_k, \alpha_{k-1}, \alpha_{k-2}, \ldots, \alpha_1, \alpha_0 \in \{0, 1\}$ (that is, all numbers whose representation in base $3$ consists of only $0$s and $1$s). 
\end{definition}

Before proving our main result for this section, we first consider a lemma, which will be of use both in the proof of our main theorem and in future sections.

\begin{lemma} \label{lemma1}
Let $n$ a positive integer such that $n\not\in S_A$. Then there exist specific $\ell, m\in S_A$ whose base $3$ representations are $(\ell_k \ell_{k-1}\dots \ell_0)_3$ and $(m_km_{k-1}\ldots m_0)_3$ such that $\ell, m, n$ is a strictly increasing arithmetic progression and, for each $0\leq i \leq k$, either $\ell_i=m_i$ or $\ell_i=0, m_i=1$. Furthermore, for each $i$ such that $\ell_i=m_i$, the $i^{th}$ digit of the base 3 representation of $n$ is also $\ell_i$. For each $i$ such that $\ell_i=0, m_i=1$, the $i^{th}$ digit of the base 3 representation of $n$ is $2$.
\end{lemma}
\begin{proof}
Because $n$ does not belong to $S_A$, its base 3 representation contains at least one 2. We can therefore express $n$ as
\[n=2\cdot (3^{y_1}+3^{y_2}+\ldots + 3^{y_i})+n_{x_1}\cdot 3^{x_1}+n_{x_2}\cdot 3^{x_2}+\ldots+n_{x_j}\cdot 3^{x_j}\]
where the exponents are pairwise distinct, and each $x_1,\ldots, x_j\in \{0,1\}$. (In essence, we are separating the powers of 3 with digit 2 in the base 3 representation of $n$ from the powers of 3 with digit either 0 or 1)

Define $\ell$ as
\[\ell=0\cdot (3^{y_1}+3^{y_2}+\ldots + 3^{y_i})+n_{x_1}\cdot 3^{x_1}+n_{x_2}\cdot 3^{x_2}+\ldots+n_{x_j}\cdot 3^{x_j}\]
and $m$ as
\[m=1\cdot (3^{y_1}+3^{y_2}+\ldots + 3^{y_i})+n_{x_1}\cdot 3^{x_1}+n_{x_2}\cdot 3^{x_2}+\ldots+n_{x_j}\cdot 3^{x_j}.\]
Clearly, $\ell$ and $m$ have all digits in their base 3 representation equal to 0 or 1, so $\ell, m\in S_A$. It is also plain to see that $\ell, m, n$ form a strictly increasing arithmetic progression. Finally, it is clear by construction that $\ell,m,$ and $n$ satisfy the properties specified in the statement of the lemma.
\end{proof}

\begin{theorem} \label{thm:mainThmA}
The set $S_A$ and the sequence $A$ have the same elements, that is $S_A = A$. If the binary representation of $n$ is $(b_kb_{k-1}\ldots b_0)_2$, where $b_k, b_{k-1}, \ldots, b_0 \in \{0, 1\}$, then $a_n$ is given by the expression
$a_n = (b_k b_{k-1} \ldots b_0)_3.$

\end{theorem}

\begin{proof}
We prove this theorem by induction.

We have two base cases, $0$ and $1$. These numbers have base $3$ representations $(0)_3$ and $(1)_3$, respectively, and both numbers are in $A$: $0$ by definition, and $1$ because it does not form an arithmetic progression with $0$ alone.

Now we prove the inductive case. Let us assume that all numbers up to $n-1$ satisfy the inductive hypothesis: that is, all numbers up to $n-1$ in $S_A$ are in $A$, and all numbers up to $n-1$ not in $S_A$ are {not} in $A$.

We need to show that the inductive hypothesis holds for $n$ as well. There are two simple cases that need to be considered.

\begin{itemize}

\item \textbf{Case 1:} $n \in S_A$

We write $n = n_k \cdot 3^k + n_{k-1} \cdot 3^{k-1} + n_{k-2} \cdot 3^{k-2} + \ldots + n_1 \cdot 3 + n_0$, where each $n_i \in \{0, 1\}$.

We need to prove that $n$ is a part of the sequence: that is, $n$ does not form a strictly increasing arithmetic progression with any two existing terms in the sequence.

For the sake of contradiction, assume otherwise. Then $n$ must form a strictly increasing arithmetic progression with two other terms that already exist in the sequence. Let the larger of these two terms be $m$, and let the smaller be $\ell = \ell_k \cdot 3^k + \ell_{k-1} \cdot 3^{k-1} + \ell_{k-2} \cdot 3^{k-2} + \ldots + \ell_1 \cdot 3 + \ell_0$. By our induction hypothesis, we see that the base 3 representation of $\ell$ must consist of only 0's and 1's; that is, each $\ell_i\in\{0,1\}$. Similarly, we see that the base 3 representation of $m$ also consists of only 0's and 1's.

Note that $\ell, m, n$ being a strictly increasing arithmetic progression implies that $\ell+n=2m$. Since each digit of the base 3 representation of $m$ is either 0 or 1, we see that each of the terms $(n_k + \ell_k), (n_{k-1} + \ell_{k-1}), \ldots, (n_1 + \ell_1), (n_0 + \ell_0)$ must equal either $0$ or $2$ (since they must equal $2 \cdot m_i$ where $m_i \in \{0, 1\}$). Because each $n_i$ and $\ell_i$ equals either 0 or 1, it follows that for all $i$, we have $n_i=\ell_i$. But then $\ell=n$, a contradiction as $\ell, m, n$ are all distinct by assumption. Hence such an arithmetic progression cannot exist, and it follows that $n\in A$, as desired.
\end{itemize}

\begin{itemize}
\item \textbf{Case 2:} $n \not \in S_A$

From Lemma \ref{lemma1}, we see that we can obtain $\ell, m\in S_A$ such that $\ell, m, n$ is a strictly increasing arithmetic progression. By the inductive hypothesis, $\ell, m$ are also in $A$. Hence $n$ is not an element of $A$, as desired.
\end{itemize}

Since we have checked the inductive hypothesis for both possibilities for $n$, it follows by induction that $S_A=A$.

Now, since the sequence is monotonically increasing, we see that $a_0 = (0)_3$, $a_1 = (1)_3$, $a_2 = (10)_3$, $a_3 = (11)_3$ and so on.

Observe that $0 = (0)_2$, $1 = (1)_2$, $2 = (10)_2$, $3 = (11)_2$ and so on. Combining the above two facts, we obtain the following closed form expression for $a_n$. If the binary representation of $n$ is $(b_k b_{k-1} \ldots b_0)_2$, where $b_k, b_{k-1}, \ldots, b_0 \in \{0, 1\}$, then $a_n$ is equal to
$$a_n = (b_k b_{k-1} \ldots b_0)_3.$$
That is, the base 3 representation of $a_n$ is the same sequence of $0$'s and $1$'s as $n$ in binary.

\end{proof}

\emph{Leon rewrote parts of the proof and made close edits to this section. Yajit wrote parts of this section and made close edits. Deepak wrote parts of this section, and edited closely.}
\section{Avoiding three-term geometric progressions} \label{sec:GP}

It is natural to ask if there are simple variations of the sequence $A$ with similarly nice closed form expressions for the $n^{\text{th}}$ term. For instance, instead of avoiding three-term arithmetic progressions, what if we avoided three-term geometric progressions instead?

We now consider the problem of avoiding geometric progressions of length $3$. As it turns out, this problem is closely related to our original problem. Before we proceed further, however, let us formally define this new sequence $G$.

\begin{definition} \label{def:seqG}
$G$ is defined as the sequence of numbers $g_0, g_1, g_2, \ldots$ such that $g_0 = 1$, and all subsequent terms $g_{n+1}$ are defined as the least term greater than $g_n$ such that no three distinct terms in the sequence including $g_{n+1}$ form a geometric progression.
\end{definition}

Here are the first one hundred terms of $G$:

\

1, 2, 3, 5, 6, 7, 8, 10, 11, 13, 14, 15, 16, 17, 19, 21, 22, 23, 24, 26, 27, 29, 30, 31, 33, 34, 35, 37, 38, 39, 40, 41, 42, 43, 46, 47, 48, 51, 53, 54, 55, 56, 57, 58, 59, 61, 62, 65, 66, 67, 69, 70, 71, 73, 74, 77, 78, 79, 80, 81, 82, 83, 85, 86, 87, 88, 89, 91, 93, 94, 95, 97, 101, 102, 103, 104, 105, 106, 107, 109, 110, 111, 112, 113, 114, 115, 118, 119, 120, 122, 123, 125, 127, 129, 130, 131, 133, 134, 135, 136.

\

While this sequence lacks a pattern as distinguished as the pattern involving powers of three in $A$, we notice that {\itshape{most}} integers appear in this sequence. Indeed, if we plot this sequence and compare it to a plot of the sequence $A$, we see that $A$ grows much faster.

\begin{figure}[!h]
  \centering
    \includegraphics[resolution=120]{GvsS.png}
    \caption{A plot of $a_n$ and $g_n$ versus $n$. Clearly $A$ grows much faster than $G$}
\end{figure}

In this case it is interesting to observe which integers do not appear in this sequence. In particular, the integers

\

4, 9, 12, 18, 20, 25, 28, 32, 36, 44, 45, 49, 50, 52, 60, 63, 64, 68, 72, 75, 76, 84, 90, 92, 96, 98, 99, 100, 108, 116, 117, 121, 124, 126, 128, 132

\

do not appear in the sequence. If we present the prime factorizations of these numbers we notice a few patterns.

$$
2^2, 3^2, 2^3\cdot 3, 2\cdot3^2, 2^2\cdot 5, 5^2, 2^2\cdot7, 2^5, 2^2\cdot 3^2,\cdots
$$

First, note that squares of primes don't occur in the sequence. It is easy to see why: 1 is in the sequence, and any prime $p$ should appear in the sequence since $p$ cannot be written as a product of smaller integers, and therefore $p$ cannot be in a geometric progression. It then follows that no $p^2$ can be in the sequence.

In fact, if we ask what powers of $p$ can appear in the sequence, we quickly realize that to avoid geometric progressions in the powers of a single prime we must avoid arithmetic progressions in the exponents of the prime. It is now clear why the original arithmetic sequence in Section \ref{sec:AP} is so crucial to the geometric sequence problem.

This discussion motivates the following theorem. Before stating the theorem, we first define the set $S_G$.
\begin{definition} \label{def:S_G}
$S_G$ is defined to be set of positive integers that have a prime factorization of the form $p_1^{a_1} p_2^{a_2} \ldots p_k^{a_k}$ where $p_1, p_2, \ldots, p_k$ are distinct primes and $a_1$, $a_2$, \ldots, $a_k$ are elements of the sequence $A$.
\end{definition}

\begin{theorem} \label{thm:mainThmG}
 The set $S_G$ and the sequence $G$ have the same elements, that is $S_G=G$. 
\end{theorem}

\begin{proof}
We prove this result using induction. Our inductive hypothesis is that all numbers $i \leq k$ of the form described above (that is, all integers $i \leq k$ in the set $S_G$) belong to the sequence $G$, and that all numbers not of this form (that is, all integers $i \leq k$ not in the set $S_G$) do not belong to the sequence $G$.

Our base case is $k = 2$. $2$ has a prime factorization of $2^1$, and hence is clearly in $S_G$. $1$ and $2$ are also both part of the sequence $G$, since no three-term geometric progression can be formed from $1$ and $2$.

Now, let us assume that the inductive hypothesis holds up to some $k-1$. We want to prove the inductive hypothesis holds for $k$ as well.

Let us factor $k$ as $p_1^{\alpha_1} p_2^{\alpha_2} \ldots p_n^{\alpha_n}$ where $p_1, p_2, \ldots, p_n$ are distinct primes. We consider the following two cases.

\begin{itemize}

\item \textbf{Case 1:} $\alpha_1$, $\alpha_2$, \ldots, $\alpha_n$ are elements of the sequence $A$.

In this case, we want to prove that $k$ is a part of the sequence $G$, that is $k$ does not form a geometric progression with any two $\ell$ and $m$ already in $G$.

Suppose for the sake of contradiction that there exist $\ell, m\in G$ strictly less than $k$ such that $\ell, m, k$ is a three-term geometric progression. Let $\ell = p_1^{\beta_1} p_2^{\beta_2} \ldots p_n^{\beta_n}$ and let $m = p_1^{\gamma_1} p_2^{\gamma_2} \ldots p_n^{\gamma_n}$. Note that some of the $\alpha_i$'s, $\beta_i$'s and $\gamma_i$'s may be equal to $0$, but we do \emph{not} allow $\alpha_i = \beta_i = \gamma_i = 0$ for any $i \in \{1,2,\ldots,n\}$.

Now, by the inductive hypothesis, all the $\beta_i$'s and $\gamma_i$'s are members of the sequence $A$ (including the $0$'s). In addition we just assumed that all the $\alpha_i$'s are members of the sequence $A$ as well.

Without loss of generality, let $\ell < m$. Since $\ell$, $m$ and $k$ form a three-element geometric progression, we see that $\ell \cdot k$ must equal $m^2$. This is equivalent to saying,
$$(p_1^{\beta_1} p_2^{\beta_2} \ldots p_n^{\beta_n}) \cdot (p_1^{\alpha_1} p_2^{\alpha_2} \ldots p_n^{\alpha_n}) = (p_1^{\gamma_1} p_2^{\gamma_2} \ldots p_n^{\gamma_n})^2$$

By comparing the exponents of primes on the left and right sides of the above expression, we obtain the following system of equations that need to hold,

\begin{align}
\alpha_1 + \beta_1 & = 2 \gamma_1 \nonumber \\
\alpha_2 + \beta_2 & = 2 \gamma_2 \nonumber \\
&\mathrel{\makebox[\widthof{=}]{\vdots}} \nonumber \\
\alpha_n + \beta_n & = 2 \gamma_n \nonumber
\end{align}

But note that $\alpha_i$, $\beta_i$ and $\gamma_i$ are members of the sequence $A$, so it is impossible for each of these equations to hold unless $\alpha_i = \beta_i = \gamma_i$. But if for all $i \in \{1, 2, \ldots, n\}$, $\alpha_i = \beta_i = \gamma_i$, then $\ell = m = k$, which is impossible since $\ell$ and $m$ were strictly less than $k$. Therefore by induction $S_G\subseteq G$.

\item \textbf{Case 2:} At least one of $\alpha_1$, $\alpha_2$, \ldots, $\alpha_n$ is not an element of the sequence $A$.

In this case, we want to prove that $k$ is not an element of the sequence $G$, that is, $k$ forms a three-element geometric progression with some distinct $\ell$ and $m$, where $\ell$ and $m$ are elements of the sequence $G$.

We will now separate the set of indices $I = \{1, 2, \ldots, n\}$ into two disjoint sets $X = \{x_1, x_2, \ldots, x_{r}\}$ and $Y = I \setminus X = \{y_1, y_2, \ldots, y_{n - r}\}$ such that $X$ contains all indices $i$ such that $\alpha_i$ is not an element of the sequence $A$, and $Y$ contains all indices $i$ such that $\alpha_i$ is an element of the sequence $A$.

Using the set of indices as defined above, we see that we can express the integer $k$ as,
$$k = (p_{x_1}^{\alpha_{x_1}} p_{x_2}^{\alpha_{x_2}} \ldots p_{x_{r}} ^ {\alpha_{x_{r}}}) (p_{y_1}^{\alpha_{y_1}} p_{y_2}^{\alpha_{y_2}} \ldots p_{y_{n - r}}^{\alpha_{y_{n - r}}})$$

Without loss of generality, assume that $\ell < m$. Since we want $\ell$, $m$ and $k$ to be in a geometric progression, $\ell$, $m$ and $k$ must satisfy the equation $\ell \cdot k = m^2$.

Consider the following definitions of $\ell$ and $m$,
$$\ell = (p_{x_1}^{\beta_{x_1}} p_{x_2}^{\beta_{x_2}} \ldots p_{x_{r}} ^ {\beta_{x_{r}}}) (p_{y_1}^{\alpha_{y_1}} p_{y_2}^{\alpha_{y_2}} \ldots p_{y_{n - r}}^{\alpha_{y_{n - r}}})$$
$$m = (p_{x_1}^{\gamma_{x_1}} p_{x_2}^{\gamma_{x_2}} \ldots p_{x_{r}} ^ {\gamma_{x_{r}}}) (p_{y_1}^{\alpha_{y_1}} p_{y_2}^{\alpha_{y_2}} \ldots p_{y_{n - r}}^{\alpha_{y_{n - r}}})$$

If we now impose the condition that for all $x \in X$, $\beta_x$ and $\gamma_x$ are elements of the sequence $A$, then it is easy to see that by our inductive hypothesis, $\ell$ and $m$ as defined above are in the sequence $G$ (since they're in the set $S_G$).

Furthermore, if the following system of equations hold, $\ell$, $m$ and $k$ will form a three-term geometric progression.

\begin{align}
\alpha_{x_1} + \beta_{x_1} & = 2 \gamma_{x_1} \nonumber \\
\alpha_{x_2} + \beta_{x_2} & = 2 \gamma_{x_2} \nonumber \\
&\mathrel{\makebox[\widthof{=}]{\vdots}} \nonumber \\
\alpha_{x_{r}} + \beta_{x_{r}} & = 2 \gamma_{x_{r}} \nonumber
\end{align}

Since $\alpha_{x_1}$, $\alpha_{x_2}$, \ldots, $\alpha_{x_{r}}$ are not elements of the sequence $A$, it is possible to find corresponding $\beta_i$'s and $\gamma_i$'s that are elements of the sequence $A$, such that $\alpha_i$, $\beta_i$ and $\gamma_i$ form a three-element arithmetic progression. Therefore by induction $G\subseteq S_G$. 
\end{itemize}

With the above two cases proven, our inductive argument is complete. We conclude that $S_G=G$.

\end{proof}

\emph{Deepak wrote this section and edited closely. Leon made edits to this section, and produced Figure 1. Yajit wrote the exposition to this section and edited closely.}

\section{Changing the first term} \label{sec:firstTerm}

Another variation of the problem arises by changing the first term. For instance, in $A$, we have that $a_1=1$. Setting $a_1=n$ for an arbitrary $n\in \mathbb N$ gives a different sequence.
\begin{definition} \label{def:A_m}
The sequence $A_m$ is defined as the sequence of numbers $a_0=0, a_1=m$, and all subsequent terms $a_{n+1}$ defined as the least term greater than $a_n$ such that no three distinct terms in the sequence so far form an arithmetic progression.
\end{definition}
This definition yields a multitude of new sequences. For instance, 
\[A_2=0, 2, 3, 5, 9, 11, 12, 14, 27, \ldots,\]
\[A_3=0, 3, 4, 7, 9, 12, 13, 16, 27, \ldots,\]
and
\[A_8=0, 8, 9, 11, 12, 17, 19, 20, 33, \ldots\]

Our central result of this section relates to the sequence $A_2$.
\begin{definition} \label{def:b_n}
Define the sequence $\{b_i\}$ as follows: for each positive integer $i$, write the binary representation of $i$ as $(x_kx_{k-1}\ldots x_1x_0)_2$. Then $b_n$ is defined as
\[b_i=(x_k x_{k-1}\ldots x_1x_0)_3\]
when $i$ is even, and
\[b_i=(x_k x_{k-1}\ldots x_{1}(x_0+1))_3\]
when $i$ is odd.
\end{definition}
Note by the closed form expression for $A$ that, for $i$ even, $b_i$ is equal to the $i$th term of $A$, and that for $i$ odd, $b_i$ is one larger than the $i$th term of $A$. Note also that $\{b_i\}$ as a set consists precisely of those positive integers whose base 3 representation has a 0 or 2 in the units digit, and a 0 or 1 for every other digit.

Before we state and prove the main theorem involving this sequence $A_2$, we consider the following lemma.

\begin{lemma} \label{lemma2}
Let $t$ be a positive integer satisfying $b_{n-1}<t<b_n$ for some integer $n>1$. Then there exist elements $r, s\in \{b_i\}$ such that $r, s, t$ form a strictly increasing arithmetic progression with common difference strictly greater than $0$.
\end{lemma}
\begin{proof}
Note that, when $n$ is even, we have $a_{n-1}+1=b_{n-1}$ and $a_n=b_n$, so that $t$ satisfies $a_{n-1}+1<t<a_n$. When $n$ is odd, we instead obtain that $a_{n-1}<t<a_n+1$. We first prove that the claim holds when $t$ satisfies $a_{n-1}<t<a_n$. This covers the case when $n$ is even; to show the case when $n$ is odd, we separately prove the lemma for $t=a_n$.

Suppose $t$ satisfies $a_{n-1}<t<a_n$. Then by Lemma \ref{lemma1} there exist $\ell, m\in A$ such that $\ell, m, t$ is an arithmetic progression. We seek to vary $\ell$ and $m$ to obtain elements $\ell', m'\in \{b_n\}$ such that $\ell', m', t$ is an arithmetic progression.

Write the base 3 representations of $\ell$ and $m$ as $\ell=\ell_k\cdot 3^k+\ldots+\ell_1\cdot 3 + \ell_0$ and $m=m_k\cdot 3^k+\ldots+m_1\cdot 3 + m_0$. We have by our choice of $\ell$ and $m$ that $\ell_i=m_i$ or $\ell_i=0,m_i=1$ for each $0\leq i \leq k$. We divide our analysis into a number of cases, depending on the values of $\ell_0, \ell_1, m_0$, and $m_1$. 
\begin{itemize}
\item $\ell_0=m_0=0$. In this case, $\ell$ and $m$ have base 3 representation consisting of a 0 in the units digit, and a 0 or 1 in every other digit. Hence $\ell, m\in \{b_i\}$, and $\ell, m, t$ is a strictly increasing arithmetic progression.
\item $\ell_0=0, m_0=1$. Since $\ell, m, t$ form a strictly increasing arithmetic progression, so do $\ell+2, m+1, t$. Then $\ell+2$ has base 3 representation $\ell_k\cdot 3^k+\ldots+\ell_1\cdot 3 + 2$, and $m+1$ has base 3 representation $m_k\cdot 3^k+\ldots + m_1\cdot 3 + 2$; since both $\ell+2$ and $m+1$ have base 3 representation with a 0 or 2 in the units digit, and a 0 or 1 for every other digit, both $\ell+2$ and $m+1$ are in $\{b_i\}$. Finally, because $\ell, m, t$ was a strictly increasing arithmetic progression, we must have that $m+1\leq t$. Hence $\ell+2, m+1, t$ must either be a strictly increasing arithmetic progression or a constant one.

Suppose for the sake of contradiction that it were a constant arithmetic progression, so that $t=m+1$. Then $t\in \{b_i\}$, but by assumption, $b_{n-1}<t<b_n$; hence $t\not\in \{b_i\}$, and we arrive at a contradiction. Thus $\ell+2, m+1, t$ is a strictly increasing arithmetic progression, with $\ell+2, m+1\in \{b_i\}$, as desired.
\item $\ell_0=m_0=1, \ell_1=m_1=0$: Because $\ell, m, t$ is a strictly increasing arithmetic progression, so is $\ell+4, m+2, t$. It is clear that $\ell+4$ has base 3 representation given by $\ell_k\cdot 3^k+\ldots + \ell_2\cdot 3^2 + 1\cdot 3 + 2$, and $m+2$ has base 3 representation given by $m_k\cdot 3^k+\ldots + m_2\cdot 3^2+1\cdot 3 + 0$. Hence both $\ell+4$ and $m+2$ are in $\{b_i\}$.

As in the previous case, we need to check that $\ell+4, m+2, t$ is a strictly increasing arithmetic progression. Well, because $\ell\neq m$, there must exist some minimal $i$ such that $\ell_i=0, m_i=1$. By assumption, this $i$ is not 0 or 1; hence it is at least 2, and so $m-\ell\geq 9$. In particular, $\ell+4<m+2$; hence $\ell+4, m+2, t$ is a strictly increasing arithmetic progression with $\ell+4, m+2\in \{b_i\}$, as desired.
\item $\ell_0=m_0=1, \ell_1=m_1=1$. Then $\ell-2, m-1, t$ is certainly an arithmetic progression, and since $m>\ell$ we have that $m-1>\ell-2$, so that it is in fact a strictly increasing arithmetic progression. The base 3 representation of $\ell-2$ is given by $\ell_k\cdot 3^k+\ldots + \ell_2\cdot 3^2 + 0\cdot 3^1 + 2$, and the base 3 representation of $m-1$ is given by $m_k\cdot 3^k+\ldots + m_2\cdot m^2 + 1\cdot 3 + 0$, so that both $\ell-2$ and $m-1$ are in $\{b_i\}$. Hence $\ell-2, m-1, t$ is a strictly increasing arithmetic progression with $\ell-2, m-1\in \{b_i\}$.
\item $\ell_0=m_0=1, \ell_1=0, m_1=1$. Then $\ell+2, m+1, t$ is certainly an arithmetic progression. Also, the base 3 represention of $\ell+2$ is given by $\ell_k\cdot 3^k+\ldots + \ell_2\cdot 3^2+ 1\cdot 3 + 0$, and the base 3 representation of $m+1$ is $m_k\cdot 3^k+\ldots + m_2\cdot 3^2 + 1\cdot 3 + 2$, so that both $\ell+2$ and $m+1$ are in $\{b_i\}$. Finally, the same argument as in the second of these cases shows that $\ell+2, m+1, t$ is in fact a strictly increasing arithmetic progression.
\end{itemize}
Having exhausted all possibilities, the claim is proved for $t$ satisfying $a_{n-1}<t<a_n$. 
As stated previously, this is sufficient to prove the lemma for even $n$, as the condition $b_{n-1}<t<b_n$ becomes $a_{n-1}+1<t<a_n$. When $n$ is odd, our condition is that $a_{n-1}<t<a_n+1$. The above argument covers the case when $a_{n-1}<t<a_n$; we must check the $t=a_n$ case separately. This amounts, of course, to exhibiting elements $r, s \in \{b_i\}$ such that $r,s, a_n$ is a strictly increasing arithmetic progression.

Write the base 3 representation of $t=a_n$ as 
\[t=t_k\cdot 3^k+\ldots + t_2\cdot 3^2 + t_1\cdot 3 + t_0,\] where each $t_i\in \{0, 1\}$. In fact, because $n$ is odd, the closed form expression for $a_n$ tells us that $t_0=1$.

Pick the smallest $i>0$ such that $t_i=1$. (Note that, if no such $i$ exists, then $a_n=1$, so $n=1$, a contradiction of our assumption that $n>1$.) Let 
\[\ell=t_k\cdot 3^k + \ldots + t_{i+1}\cdot 3^{i+1} + 0\cdot 3^i + 0\cdot 3^{i-1} + 0\cdot 3^{i-2} + \ldots + 0 \cdot 3+ 0,\]
and 
\[m=t_k\cdot 3^k + \ldots + t_{i+1}\cdot 3^{i+1} + 0\cdot 3^i + 1\cdot 3^{i-1} + 1\cdot 3^{i-2} + \ldots + 1 \cdot 3+ 2.\]
Clearly $\ell, m$ have base 3 representation with a 0 or 2 in the units digit, and a 0 or 1 for every other digit; hence $\ell, m\in \{b_i\}$. It is also easy to see that
\[m-\ell=1\cdot 3^{i-1}+1\cdot 3^{i-2}+\ldots + 1\cdot 3 + 2,\]
and that
\[t-m=1\cdot 3^{i-1}+1\cdot 3^{i-2}+\ldots + 1\cdot 3 + 2.\]
Hence $\ell, m, t$ is a strictly increasing arithmetic progression with $\ell, m\in \{b_i\}$, as desired.
\end{proof}

Having proven the above lemma, we move on to our main theorem.

\begin{theorem} \label{thm:A_2}
The sequence $A_2$ is precisely the sequence $\{b_2\}$. In particular, for $n$ even, the $n^{th}$ term of $A_2$ equals the $n^{th}$ term of $A$, and for $n$ odd, the $n^{th}$ term of $A_2$ is one larger than the $n^{th}$ term of $A$.
\end{theorem}

\begin{proof}
We prove the theorem by induction on $n$. The base cases, $n=0$ and $n=1$, are easy to verify: $b_0=0$ is defined to be the zeroth term of $A_2$, and $b_1=2$ is defined to be the first term of $A_2$.

Suppose the theorem holds up to some arbitrary $n-1$: that is, the $i$th term of $A_2$ is $b_i$ for all positive integers $i\leq n-1$. We want to show that the $n^{th}$ term of $A_2$ is $b_n$. To do so, for all integers $t$ such that $b_{n-1}<t<b_n$, we will exhibit an increasing three-term arithmetic progression $r, s, t$ with $r,s\in A_2$; this will show that $t\not\in A_2$. We will then show that $b_n$ is in the sequence by supposing such an arithmetic progression existed and deriving a contradiction.

First, pick an arbitrary $t$ such that $b_{n-1}<t<b_n$. In fact, most of the work has already been done: by Lemma \ref{lemma2}, there exist $r, s\in \{b_i\}$ such that $r, s, t$ is an increasing arithmetic progression. By the induction hypothesis, $r,s\in A_2$. Hence $t\not\in A_2$.

Now consider $t=b_n$, and suppose for the sake of contradiction that there exist $r,s\in A_2$ such that $r, s,t$ is an increasing geometric sequence. Write the base 3 representations of $r$ and $s$ as $r=r_k\cdot 3^k+\ldots + r_1\cdot 3 + r_0$ and $s=s_k\cdot 3^k+\ldots + s_1\cdot 3 + s_0$. Note that, by the induction hypothesis, $r,s\in \{b_i\}$, and so that $r_0, s_0\in \{0, 2\}$ and all other $r_i, s_i\in \{0, 1\}$.

We split our analysis into two cases based on the parity of $n$.

\begin{itemize}
\item{\bf Case 1:} $n$ is even.

Then $b_n=a_n$, and we can write the base 3 representation of $t=b_n$ as $t=t_k\cdot 3^k+\ldots + t_1\cdot 3 + t_0$, with $t_0=0$ and all other $t_i\in \{0,1\}$. Note that $r+t=2s$, so that in particular $r_0+t_0\equiv r_0\equiv 2s_0\mod 3$.

Recall that $s_0, r_0\in \{0, 2\}$. Suppose that $r_0=0$: then $2s_0=0+0=0$ as well, so that $s_0=0$ and $r, s\in A$. Then $r,s,t$ would be an arithmetic progression with $r,s,t\in A$, a contradiction; so $r_0\neq 0$. Now suppose that $r_0=2$. Then $2s_0=2+0=2$, so that $s_0=1$, again a contradiction as $s_0\in \{0, 2\}$. Hence $r_0\neq 2$, and so $r_0\not\in \{0, 2\}$; another contradiction. It follows that such a sequence $r,s,t$ cannot exist; hence $t=b_n$ is indeed in $A_2$.
\item {\bf Case 2:} $n$ is odd.

Then $b_n=a_n+1$, and we can write the base 3 representation of $t=b_n$ as $t=t_k\cdot 3^k+\ldots + t_1\cdot 3 + t_0$, with $t_0=2$ and all other $t_i\in \{0,1\}$. Note that $r+t=2s$, so that in particular $r_0+t_0\equiv r_0+2\equiv 2s_0\mod 3$.

As before, we have that $s_0, r_0\in \{0, 2\}$. Suppose that $r_0=0$: then $2s_0=2\mod 3$, and $s_0=1$, a contradiction as $s_0\in \{0, 2\}$. Now suppose that $r_0=2$: then $2s_0\equiv 2+2\equiv 1\mod 3$, so that $s_0=2$. Also, because $r_0=t_0=2$, a one carries over into our sum in the 3 digit, and we must have that $2s_1\equiv r_1+t_1+1\mod 3$. Recall that $r_1, t_1,s_1\in \{0,1\}$: checking all possibilities for $r_1,t_1$ shows that the only possibilities are $s_1=r_1=1, t_1=0$, or $s_1=t_1=1, r_1=0$, or $r_1=t_1=1, s_1=0$.

In the first two cases, no additional one carries over in the sum $r+t$, and so for all $i>1$ we must have $2s_i\equiv r_i+t_i\mod 3$. Then we must $s_i=r_i=t_i$, as $s_i=0$ implies $r_i=t_i=0$ and $s_i=1$ implies $r_i=t_i=1$; hence either $s=r$ (in the first case) or $s=t$ (in the second), and we thus obtain a contradiction. In the last case, another one carries over, and we must have that $2s_2\equiv r_2+t_2+1\mod 3$, and the same argument holds to show that $s_2=0$ and $2s_3\equiv r_3+t_3+1\mod 3$. We can proceed by induction to conclude that $s_i=0$ for all $i>0$: thus $s=2$, and so it must follow that $r=0$ and $t=4$. But $r\not\in \{b_i\}$, as can be easily checked; thus we obtain a contradiction, and it follows again that such a sequence $r, s, t$ cannot exist. Hence $t=b_n$ is indeed in $A_2$.
\end{itemize}\end{proof}
This proof is apparently much more involved than the proof of Theorem \ref{thm:mainThmA}. In some sense, however, we are lucky simply to have the result at all; most other sequences obtained by varying the first term have more complicated behavior than $A_2$, as can be seen in the first graph below.

Based on Figure $2$, it seems plausible that all such sequences might have the same growth order. Indeed, Figure 3 plots $A_5/n^{\log_2 3}$ and $x^{\log_2 3}$ on a $\log$-$\log$ scale, and the two lines do appear to have very similar slopes.

\emph{Leon wrote this section and produced the figures. Yajit edited this section.}

\begin{figure}[!h]
  \centering
    \includegraphics[resolution=140]{plot1.png}
    \caption{Plot of terms in the sequence $A_m$ vs $n$ for different values of $m$}
\end{figure}

\vspace{.5in}

\begin{figure}[!h]
  \centering
    \includegraphics[resolution=140]{plot3.png}
    \caption{Log-log plot of $A_5$ versus $n^{\log_2 3}$}
\end{figure}
\newpage\section{Avoiding solutions to the equation $x + y = z$} \label{sec:diffEquation}

It turns out that the sequence $A$ can be seen as finding terms that avoid a solution to the equation $x + y = 2z$. Interestingly enough, tweaking this equation gives us other sequences with completely different properties.

To begin, we are going to examine the equation $x + y = z$.

We define the modified sequence $A'$ below.

\begin{definition} \label{def:A'}
The sequence $A'$ is defined as the sequence of numbers $a_0, a_1, a_2, \ldots.$ such that $a_0 = 0$, and all subsequent terms $a_{n+1}$ are defined as the least term greater than $a_n$ such that no three distinct terms $x, y, z$ in the sequence $A'$ satisfy the equation $x + y = z$.
\end{definition}

Upon inspection, we see that the first few terms of the sequence are $0, 1, 2, 4, 7, 10, \ldots$. One might therefore predict that every subsequent element $a_n$ for $n \geq 3$ is of the form $3k + 1$, where $k$ is an integer. In fact, we make this statement even stronger.

\begin{theorem}
Every integer of the form $3k + 1$ belongs to the sequence $A'$, and the only integers of the form $3k$ and $3k+2$ that belong to the sequence are $0$ and $2$.
\end{theorem}

\begin{proof}
We will prove this hypothesis by induction.
Our base cases here are $0$, $1$ and $2$ -- clearly they belong to the sequence, because the sequence starts at $0$ and $2 > 0 + 1$.

Now we need to prove the inductive case. Our inductive hypothesis is as follows -- For all numbers $i < k$, if $i \mod 3 = 1$, then $i$ belongs to the sequence $A'$, otherwise $i$ belongs to the sequence only if $i = 0$ or $i = 2$.

Now, we need to prove the inductive hypothesis for $k$. We split this into two cases.

\begin{itemize}

\item \textbf{Case 1:} $k \equiv 1 \mod 3$

We want to prove that when $k \equiv 1 \mod 3$, $k$ belongs to the sequence $A'$. This is easy to see by just looking at remainders modulo 3. By our inductive hypothesis, every number in this sequence is $1$ modulo $3$ except $0$ and $2$. There is no way to obtain a number that is equal to $1$ modulo $3$ by summing two numbers that are also $1$ modulo $3$. In fact, the only way to obtain a number $1$ modulo $3$ given the terms already in the sequence $A'$ is by summing $0$ and a number $1$ modulo $3$, but clearly this does not give a new term in the sequence. Hence we conclude that if $k \equiv 1 \mod 3$, then $k$ must belong to the sequence $A'$.

\item \textbf{Case 2:} $k \not \equiv 1 \mod 3$

Since our base case already contains $0$ and $2$, all we need to do is prove that when $k > 2$ and $k \not \equiv 1 \mod 3$, $k$ cannot belong to the sequence $A'$. This is easy to see since if $k \equiv 0 \mod 3$, we can express $k$ as the sum of $k - 2$ and $2$, which are both part of the sequence ($k - 2$ by the fact that all integers less than $k$ that are congruent to $1$ modulo $3$ belong to the sequence and $2$ by the way the sequence was defined).

Similarly, when $k \equiv 2 \mod 3$ we can express it as the sum of $k -1 $ and $1$, which are both part of the sequence, by our inductive hypothesis.

From this we can conclude that if $k \not \equiv 1 \mod 3$, $k$ does not belong to the sequence $A'$.

\end{itemize}

Since the above two cases hold, we see that the inductive hypothesis holds for $k$ as well, and hence our inductive argument is complete.

\end{proof}

As it turns out, $x+y=z$ and $x+y=2z$ are the only two equations for which we could find a closed form expression for the $n^{th}$ term. We tried several equations of the form $x+y=kz$ where $k>2$, as well as several equations of the form $k_1x+k_2y = (k_1 + k_2)z$ where $k_1 > k_2 \geq 1$. Unfortunately, none of these equations seemed to produce sequences as elegant as the ones presented in this paper.

\emph{Deepak wrote this section. Leon made edits to this section.}

\end{document}

### 2014-11-02 project1/code/sequence_generator.py

@@ -1,0 +1,49 @@
+import sys
+
+
+def get_sequence(first_term, num):
+    terms = [first_term]
+    for i in xrange(num):
+        next_term = get_next_term(terms)
+        terms.append(next_term)
+    return terms
+
+
+def get_differences(terms):
+    differences = list()
+    for i in xrange(1, len(terms)):
+        differences.append(terms[i] - terms[i-1])
+    return differences
+
+
+def get_next_term(terms):
+    possible_term = terms[-1] + 1
+    while not does_term_work(possible_term, terms):
+        possible_term += 1
+    return possible_term
+
+
+def does_term_work(possible_term, terms):
+    term_set = set([2 * term for term in terms])
+    for term in terms:
+        if (term + possible_term) in term_set:
+            print possible_term, list_not_working_pairs(possible_term, terms)
+            return False
+    return True
+
+
+def list_not_working_pairs(possible_term, terms):
+    term_set = set([2 * term for term in terms])
+    not_working_pairs = list()
+    for term in terms:
+        if (term + possible_term) in term_set:
+            not_working_pairs.append((term, (possible_term + term) / 2))
+    return not_working_pairs
+
+
+if __name__ == '__main__':
+    num = int(sys.argv[1])
+    terms = get_sequence(0, num)
+    print ", ".join([str(ele) for ele in terms])
+    differences = get_differences(terms)
+    print ", ".join([str(ele) for ele in differences])

### 2014-11-02 project1/code/sequence_generator_k1x+k2y=(k1+k2)z.py

@@ -1,0 +1,36 @@
+import sys
+
+verbose = False
+
+
+def get_sequence_linear_equation(first_term, num, k1, k2):
+    terms = [first_term]
+    for i in xrange(num):
+        next_term = get_next_term_linear_equation(terms, k1, k2)
+        terms.append(next_term)
+    return terms
+
+
+def get_next_term_linear_equation(terms, k1, k2):
+    possible_term = terms[-1] + 1
+    while not does_term_work_linear_equation(possible_term, terms, k1, k2):
+        possible_term += 1
+    return possible_term
+
+
+def does_term_work_linear_equation(possible_term, terms, k1, k2):
+    term_set_k1 = set([term * k1 for term in terms])
+    term_set_k2 = set([term * k2 for term in terms])
+    for term in terms:
+        if ((((k1 + k2) * term - (k2 * possible_term)) in term_set_k1) or
+            (((k1 + k2) * term - (k1 * possible_term)) in term_set_k2)):
+            return False
+    return True
+
+
+if __name__ == '__main__':
+    k1 = int(sys.argv[1])
+    k2 = int(sys.argv[2])
+    num_terms = int(sys.argv[3])
+    terms = get_sequence_linear_equation(0, num_terms, k1, k2)
+    print ", ".join([str(term) for term in terms])

### 2014-11-02 project1/code/sequence_generator_x+y=kz.py

@@ -1,0 +1,64 @@
+import sys
+
+verbose = False
+
+
+def get_sequence(first_term, num, k):
+    terms = [first_term]
+    for i in xrange(num):
+        if k == 1:
+            next_term = get_next_term_one(terms)
+        else:
+            next_term = get_next_term(terms, k)
+        terms.append(next_term)
+    return terms
+
+
+def get_next_term(terms, k):
+    possible_term = terms[-1] + 1
+    while not does_term_work(possible_term, terms, k):
+        possible_term += 1
+    return possible_term
+
+
+def does_term_work(possible_term, terms, k):
+    term_set = set([k * term for term in terms])
+    for term in terms:
+        if ((term + possible_term) in term_set and
+                (possible_term != (k - 1) * term)):
+            if verbose:
+                print possible_term, list_not_working_pairs(
+                    possible_term, terms, k)
+            return False
+    return True
+
+
+def get_next_term_one(terms):
+    possible_term = terms[-1] + 1
+    while not does_term_work_one(possible_term, terms):
+        possible_term += 1
+    return possible_term
+
+
+def does_term_work_one(possible_term, terms):
+    term_set = set(terms)
+    for term in terms:
+        if (possible_term - term) in term_set and (possible_term != (2 * term)):
+            return False
+    return True
+
+
+def list_not_working_pairs(possible_term, terms, k):
+    term_set = set([k * term for term in terms])
+    not_working_pairs = list()
+    for term in terms:
+        if ((term + possible_term) in term_set):
+            not_working_pairs.append((term, (possible_term + term) / k))
+    return not_working_pairs
+
+
+if __name__ == '__main__':
+    k = int(sys.argv[1])
+    num_terms = int(sys.argv[2])
+    terms = get_sequence(0, num_terms, k)
+    print terms

### 2014-11-05 project2/paper/paper.tex
Cleaned up t_n(312) = t_n(321) proof

--- 2014-11-02 paper/paper.tex
@@
 
-We now construct a bijective mapping between all permutations in $T_n$ that avoid $312$ and all permutations in $T_n$ that avoid $321$.
+Before proving that there exists a bijective mapping between all permutations in $T_n$ that avoid $312$ (call this set $A$) and all permutations in $T_n$ that avoid $321$ (call this set $B$), we establish a chain of transformation moves that can be used to transform an element $a \in A$ to an element $b \in B$.
 
@@
 For each permutation $\pi \in T_n$ that avoids $312$, we want to show that there exists a sequence of transformation moves that transforms $\pi$ to a unique permutation $\pi'$ that avoids $321$.
-
-\textbf{May want to skip the transformation stuff, and directly talk about the bijection being established.}
 
@@
 
-What remains to be proven is that the order of swapping operations in the chain of transformations does not matter -- that is, any valid transformation chain starting with some permutation $\pi \in T_n$ that avoids $312$, will always produce the same permutation $\pi' \in T_n$ that avoids $321$.
-Note that since we're only swapping even elements, the odd elements are in exactly the same position in both $\pi$ and $\pi'$.
-In addition, for every element $i \in \pi$, all even numbers in $\pi$ to the right of $i$ and less than $i$ must be in decreasing order; in $\pi'$, for every element $i \in \pi'$, all even numbers in $\pi'$ to the right of $i$ and less than $i$ must be in increasing order.
+We now prove that there exists a bijection between the sets $A$ and $B$.
+
+Let $a$ be an arbitrary permutation in $A$.
+Clearly, by the definition of $A$, $a$ must avoid $312$.
+Now, we claim that for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in decreasing order; if this were not the case, then we would have a subsequence in the relative ordering $312$, which would be a contradiction.
+It is easy to see that we can make a corresponding claim about permutation $b$, that is for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in increasing order.
+
+The above two claims shows there's actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$ -- to obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order); similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order).
+Since the mapping from $A$ to $B$ is clearly reversible, and hence invertible, we conclude that the mapping we just established between $A$ and $B$ is bijective.
 

### 2014-11-06 project2/paper/paper.tex
new introduction

--- 2014-11-02 paper/paper.tex
@@
 
-In this paper we will investigate pattern avoidance in permutations of finite sets of the first $n$ positive integers for arbitrary $n$.
+In this paper we will investigate pattern avoidance in finite sequences of natural numbers.
+Let us start by consider the ordered sequences 123 and 1324.
+We can describe the relationships between elements of the sequence 123 as follows: the first element in the sequence is the smallest element of the sequence, the second element of the sequence is the second smallest element of the sequence, and the third element of the sequence is the largest element of the sequence.
+Now notice that there is an ordered subsequence of 1324 that follows the same general pattern, namely 124.
+The relationships between elements of the sequence 124 is the same as the relationships we described for 123.
+In this situation we say that the sequence 1324 \emph{does not avoid} 123.
+If we examine the sequence 1432 we see that there are no ordered subsequences of 1432 that have the same relations between elements as 123.
+We say that 1432 \emph{avoids} 123.
 
-Before beginning our discussion, we present the following definitions.
+With an example of a four element sequence that avoids 123 and a four element sequence that does not avoid 123, it is natural to ask the question, how many four element sequences avoid 123?
+As we will see below, the answer to this question is simply 5, but this leads to a more general question, how many $n$-element sequences avoid 123?
+If we define $s_n(123)$ to be the number of $n$ element sequences that avoid 123, then the question above is equivalent to identifying the sequence $(s_n(123))_n$.
+In this paper we consider specific cases of a very general avoidance problem, describing the sequence $(s_n(\sigma))_n$ where $\sigma$ is any permutation of the elements of the set $\{1,2,\cdots, k\}\subset\mathbb{N}$ for some $k$.
+
+In section \ref{S3} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
+In section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$.
+In section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements.
+Before proceeding, we will formally introduce the notation that will be used throughout this paper.
+
+\subsection{Notation}
 
 \begin{definition}
-Two finite sequences $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
-\end{definition}
-
-\begin{definition}
-A finite sequence of numbers $a_1a_2\cdots a_n$ avoids a sequence $b_1\cdots b_k$ with $n\ge k$ if no subsequence $a_{i_1}a_{i_2}\cdots a_{i_k}$ of $a_1a_2\cdots a_n$ has its terms in the same relative order as $b_1\cdots b_k$.
+The set $S_n$ is the collection of permutations of the set $\{1,2,\cdots, n\}$. 
 \end{definition}
@@
 \begin{example}
-The sequence 15432 avoids 123 while the sequence 13425 does not. The latter example fails because 134 appears in 15342 with the same relative order as 123. 
+$S_3=\{123,132,213,231,312,321\}$. 
 \end{example}
@@
 \begin{remark}
-Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$.
+The notation used to describe the elements of $S_n$ above is `one-line' notation.
+Thus far we have ignored the fact that $S_n$ has a group structure.
+When discussing this group structure it is convenient to use a different notation known as `cycle' notation.
+This alternative notation will be used in Section\ref{S4} and an introduction to cycle notation can be found in Appendix [].
+\end{remark}
+
+\begin{definition}
+Two finite permutations $a_1,\cdots a_k$ and $b_1,\cdots b_k$ in $S_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
+\end{definition}
+
+\begin{remark}
+An equivalent definition is that two permutations in $S_k$ have the same relative order if the elements in each position of the sequence satisfy the same set of inequalities. 
+\end{remark}
+
+\begin{definition}
+A permutation $a_1a_2\cdots a_n$ avoids a permutation $b_1\cdots b_k$ with $n\ge k$ if no subsequence $a_{i_1}a_{i_2}\cdots a_{i_k}$ of $a_1a_2\cdots a_n$ has its terms in the same relative order as $b_1\cdots b_k$.
+\end{definition}
+
+%\begin{remark}
+%Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$.
 There are two conventions commonly used to write elements of $S_n$.
@@
 43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.
-
-The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
+%The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
 For example, the cycle $(14)(23)$ sends $12345$ to $43215$.
 Unless otherwise stated, we will refer to permutations in one-line notation.
-\end{remark}
+%\end{remark}
 
@@
 
-In this paper we concern ourselves with the sequence of numbers $s_n(\pi)$ for various $\pi$.
-In section \ref{S3} we will consider sequence $s_n(\pi)$ for $\pi\in S_3$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
-In section \ref{S4} we will consider sequence $s_n(\pi)$ for $\pi\in S_4$.
-In section \ref{Tn} we consider a variation on the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of $S_n$.
+\section{Flipping and Reversing}
 
@@
 We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with the largest element swapped with the smallest element, the second largest element swapped with the second smallest element, etc.
-For example, the flip of 1324 is 4321.
+For example, the flip of 1324 is 4231.
 We will discuss the flip as it becomes useful.
+
+\subsection{Reversing}
 
@@
 \hline
-1234, 4321&4132, 1423&4231,1324\\
+1234, 4321&4132, 1423&4231, 1324\\
 1243, 4312&4213, 1342&\\
-1432, 4123&2431,3124&\\
+1432, 4123&2431, 3124&\\
 2134, 3421&2413, 3142&\\

### 2014-11-06 project2/code/compute_sn.py

@@ -57,7 +57,7 @@ def compute_sn(n, order):
 
 if __name__ == '__main__':
     n = int(sys.argv[1])
-    order = [1, 2, 3, 4]
+    order = [1, 2, 3]
     k = len(order)
     for i in xrange(k, n):
         print compute_sn(i, order)

### 2014-11-06 project2/paper/paper.tex
Added proofs for t_n(213) and t_n(231)

@@
 
-Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above, partially due to the fact that the element corresponding to $2$ and $3$ in the above sub-sequences can be either even or odd, which means we now also have to account for the positions of odd elements too.
-Proving this equivalence remains future work.
+Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above, however we do prove that both the sequences $(t_n(231))$ and $(t_n(213)$ are equal to the sequence $1, 2, 4, 8, \ldots$ below.
+
+\begin{theorem}
+$t_n(231) = 1, 2, 4, 8, \ldots$
+\end{theorem}
+
+\begin{proof}
+Proving the above theorem is equivalent to proving that the $m^{th}$ term of $t_n(231)$ is equal to $2^{m-1}$, or that the number of permutations in $T_{2m}$ that avoid $231$ is equal to $2^{m-1}$.
+
+Consider a permutation $\sigma$ in $T_{2m}$ that avoids $231$.
+Consider an arbitrary even element $2i$ in $\sigma$.
+For this element $2i$, define $A$ to be the set of numbers greater than $2i$ and to the left of $2i$ in the permutation $\sigma$.
+First off, note that the set $A$ cannot contain any even numbers, because if it did, then the permutation $\sigma$ would contain a three-element sub-sequence in the relative ordering $231$ -- if $a$ is an even number in the set $A$, then $(a-1, a, 2i)$ would be a three-element sub-sequence of $\sigma$ in the relative ordering $231$ (Note that $a-1$ must be to the left of $a$ in the permutation $\sigma$ because of the way the set $T_{2m}$ was constructed.
+
+Furthermore, the set $A$ cannot contain any odd elements greater than $2i+1$, because if it did, then again the permutation $\sigma$ would contain a three-element subsequence in the relative ordering $231$ -- if $a > 2i+1$ is an odd number in $A$, then $(2i+1, a, 2i)$ is a three-element sub-sequence of $\sigma$ in the relative ordering $231$ Therefore, we conclude that the set $A$ either only contains the element $2i+1$, or is empty.
+
+From this we can conclude that the element $2i$ can be inserted into exactly two valid positions in a permutation of length $2m$ -- at index $2i$ and at index $2i+1$.
+It follows from the above result that when element $2i$ is at index $2i$, element $2i+1$ is at index $2i+1$ in the permutation, and when element $2i$ is at index $2i+1$, element $2i+1$ is at index $2i$.
+Note that the element $1$ must always be the first element in the permutation, and the element $2m$ must always be the last element.
+
+Combining all these facts together, we see that the number of permutations $\sigma$ in $T_n$ that avoid $231$ must be equal to $2^{(2m-2)/2} = 2^{m-1}$, as desired.
+
+\end{proof}
+
+\begin{theorem}
+$t_n(213) = 1, 2, 4, 8, \ldots$
+\end{theorem}
+
+\begin{proof}
+As before, proving the above theorem is equivalent to proving that the $m^{th}$ term of $t_n(213)$ is equal to $2^{m-1}$, or that the number of permutations in $T_{2m}$ that avoid $213$ is equal to $2^{m-1}$.
+
+First, let us define $A_m$ as the set of all permutations in $T_{2m}$ that avoid $213$. It is easy to see that the $m^{th}$ term of $t_n(213)$ is $|A_m|$.
+
+We now claim that for every permutation $\sigma$ in $A_m$, the numbers $2m-1$ and $2m$ must be consecutive in the permutation $\sigma$.
+We argue that this must be true by contradiction.
+If this were not the case, then it is easy to see that the permutation $\sigma$ would have a sub-sequence in the relative ordering $213$ -- let $j$ be an element that's between $2m-1$ and $2m$ in $\sigma$.
+Clearly, $j < 2m -1 < 2m$ (since every element in $\sigma$ other than $2m$ is less than $2m-1$), which means $(2m - 1, j, 2m)$ are in the relative ordering $213$.
+We conclude that $2m-1$ and $2m$ must be consecutive.
+
+Furthermore, we see that we can construct $A_m$ from $A_{m-1}$ by inserting the consecutive elements $2m-1, 2m$ into a permutation $\sigma'$ in $A_{m-1}$.
+Furthermore, we claim that we get a valid permutation $\sigma$ in $A_m$ from $\sigma'$ only if we insert the consecutive elements $2m-1, 2m$ right after either $2m-3$ or $2m-2$ in $\sigma'$ (which themselves are consecutive in $\sigma'$ from the above claim).
+
+First, note that $2m-1, 2m$ must be inserted to the right of both $2m-3$ and $2m-2$, because of the way $T_{2m}$ is defined.
+Furthermore, if we insert $2m-1, 2m$ right after some $a$ in $\sigma$ not equal to $2m-3$ and $2m-2$, then we see that the produced permutation $\sigma$ has a three-element sub-sequence in the relative ordering $213$ -- $(2m-2, a, 2m-1)$.
+Hence we conclude that $2m-1, 2m$ must be inserted to the right of either $2m-3$ or $2m-2$; since we can do this for every permutation $\sigma'$ in $A_{m-1}$, we conclude that $|A_m| = 2|A_{m-1}|$.
+
+Given that $|A_1|$ is trivially equal to $1$, we conclude that $|A_m|$, or the $m^{th}$ term of $t_n(213)$ is indeed equal to $2^{m-1}$, as desired.
+\end{proof}
+
+Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $1, 2, 4, 8, \ldots$ are vastly different.
 

### 2014-11-06 project2/paper/paper.tex
flipping and reversing, other changes

--- 2014-11-06 project2/paper/paper.tex
@@
 When discussing this group structure it is convenient to use a different notation known as `cycle' notation.
-This alternative notation will be used in Section\ref{S4} and an introduction to cycle notation can be found in Appendix [].
+This alternative notation will be used in Section \ref{S4} and an introduction to cycle notation can be found in Appendix [].
 \end{remark}
@@
 
-%\begin{remark}
-%Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$.
-There are two conventions commonly used to write elements of $S_n$.
-The first is $\emph{one-line}$ notation where permutations in $S_n$ are written explicitly, e.g.
-43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.
-%The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
-For example, the cycle $(14)(23)$ sends $12345$ to $43215$.
-Unless otherwise stated, we will refer to permutations in one-line notation.
-%\end{remark}
-
 \begin{definition} 
@@
 
+In this section we will identify operations that preserve avoidance.
+Explicitly, we say that an operator $\mathcal{O}$ preserves avoidance if a permutation $\sigma$ avoids $\pi$ if and only if $\mathcal{O}(\sigma)$ avoids $\mathcal{O}(\pi)$.
+If $\mathcal{O}$ is an avoidance preserving operator then for any $\sigma\in S_k$, $s_n(\sigma)=s_n(\mathcal{O}(\sigma))$ for all $n$ and $k$.
+Two such operators that we will identify are the `flipping' and `reversing' operators.
+Before discussing these operators, we will introduce generating trees, a tool which we use to study pattern avoidance and particularly to prove the avoidance preserving nature of these operators.
+
 \subsection{Generating Trees}
 
-Before continuing we introduce generating trees, a tool we can use to study pattern avoidance.
 The generating tree for the sequence $s_n(\pi)$ is the infinite tree whose vertices are elements of $S_n$ that avoid $\pi$.
+The first node in the tree is always 1.
 To construct the tree, we define the children of an arbitrary node.
@@
 \subsection{Flipping}
-Another notion that we will use in this paper is the concept of `flipping'.
+In this section we introduce the concept of flipping and use generating trees to prove the `Flipping Lemma.'
+\begin{definition}
 We define the \emph{flip} of a sequence $a$ as the sequence $b$ with the same elements as $a$, but with the largest element swapped with the smallest element, the second largest element swapped with the second smallest element, etc.
-For example, the flip of 1324 is 4231.
-We will discuss the flip as it becomes useful.
+The flipping operator will be denoted by $\mathcal{F}$.
+\end{definition}
 
+\begin{example}
+$\mathcal{F}(1324)=4231$.
+\end{example}
+
+\begin{lemma}[Flipping Lemma]
+The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$. I.e. $\mathcal{F}$ is an avoidance preserving operator. 
+\end{lemma}
+\begin{proof}
+Notice that we only need to proof one direction of this statement.
+The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation.
+We show that if ${\sigma}$ avoids $\pi$ then $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$.
+Let $\pi=b_1\cdots b_k$.
+We proceed by induction.
+Certainly the statement is true for elements of $S_1$.
+Suppose it is true for elements of $S_{n-1}$.
+Let $\sigma\in S_n$.
+If ${\sigma}$ does avoids $\pi$ then $\sigma$ appears in the generating tree of $\pi$.
+Then $\sigma=a_1\cdots a_n$ where $a_i\in\{1,\cdots, n\}$.
+Without loss of generality, let $a_\ell=n$.
+Then the parent node of $\sigma$ is $\sigma_p=a_1\cdots a_{\ell-1} a_{\ell+1}\cdots a_n$.
+The induction hypothesis allows us to conclude that $\mathcal{F}(\sigma_p)$ avoids $\mathcal{F}(\pi)$.
+Now suppose that $\mathcal{F}(\sigma)$ is not a child of $\mathcal{F}(\sigma_p)$.
+Let $\mathcal{F}(\sigma)=a'_1\cdots a'_{\gamma-1} a_{\gamma}a'_{\gamma+1}\cdots a'_n$ and $\mathcal{F}(\sigma_p)=a'_1\cdots a'_{\gamma-1} a'_{\gamma+1}\cdots a'_n$ with $a'_\gamma=n$.
+Then $\mathcal{F}(\sigma_p)$ avoids $\pi$ while $\mathcal{F}(\sigma)$ does not.
+That means that there is a subsequence $\omega$ of $\mathcal{F}(\sigma)$ that has the same relative order as $\mathcal{F}(\pi)$.
+Let $\mathcal{F}(\pi)= b'_1\cdots b'_k$ with $b'_i=1$ and $b'_j=k$.
+So $\omega$ contains $a'_\gamma$, and $a'_\gamma$ occupies the same position as $b'_j$.
+Now if we remove 1 and $n$ from $\mathcal{F}(\sigma)$, then the result does not avoid $\mathcal{F}(\pi)$ with $b'_i=1$ and $b'_j=k$ omitted.
+Then if we apply a flip, by the induction hypothesis we see that the result of flipping $\mathcal{F}(\sigma)$ with $1$ and $n$ omitted does not avoid $\pi$ with $1$ and $k$ omitted.
+If we add in 1 and $n$ in the appropriate places to reconstruct $\mathcal{F}(\mathcal{F}(\sigma))=\sigma$ and if we reconstruct $\sigma$ similarly, then $\sigma$ does not avoid $\pi$, a contradiction.
+\end{proof}
+
+\begin{corollary} 
+For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{F}(\pi))$. 
+\end{corollary}
+\begin{proof}
+This result follows because the flipping lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
+\end{proof}
 \subsection{Reversing}
 
+\begin{lemma}[Reversing Lemma]
+The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. I.e. $\mathcal{R}$ is an avoidance preserving operator. 
+\end{lemma}
+\begin{proof}
+Again we use the generating tree.
+Certainly the statement is true for elements of $S_1$.
+Suppose it is true for elements of $S_{n-1}$.
+Suppose that $\sigma\in S_n$ avoids $\pi$.
+Then if $\sigma_p$ is the parent of $\sigma$, $\mathcal{R}(\sigma_p)$ avoids $\mathcal{R}(\pi)$.
+Now if we add $n$ to the permutation $\mathcal{R}(\sigma_p)$ to construct $\mathcal{F}(\sigma)$, we see that $\mathcal{F}(\sigma)$ must avoid $\mathcal{F}(\pi)$, for if it did not, we could reverse one more time and see obtain notice that $\sigma\in S_n$ does not avoid $\pi$, a contradiction.
+The other direction follows because $\mathcal{R}^2=1$.
+\end{proof}
+
+\begin{corollary} 
+For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{R}(\pi))$. 
+\end{corollary}
+\begin{proof}
+This result follows because the reversing lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
+\end{proof}
 \section{Terms in the sequence $s_n(312)$}
@@
 
+\section{Appendix A}
+
+Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$.
+There are two conventions commonly used to write elements of $S_n$.
+The first is $\emph{one-line}$ notation where permutations in $S_n$ are written explicitly, e.g.
+43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.
+
+The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
+For example, the cycle $(14)(23)$ sends $12345$ to $43215$.
+
 \end{document}

### 2014-11-06 project2/paper/paper.tex
made a bunch of edits + added some todos

--- 2014-11-06 project2/paper/paper.tex
@@
 \section{Introduction}
-
-In this paper we will investigate pattern avoidance in finite sequences of natural numbers.
-Let us start by consider the ordered sequences 123 and 1324.
-We can describe the relationships between elements of the sequence 123 as follows: the first element in the sequence is the smallest element of the sequence, the second element of the sequence is the second smallest element of the sequence, and the third element of the sequence is the largest element of the sequence.
-Now notice that there is an ordered subsequence of 1324 that follows the same general pattern, namely 124.
-The relationships between elements of the sequence 124 is the same as the relationships we described for 123.
-In this situation we say that the sequence 1324 \emph{does not avoid} 123.
-If we examine the sequence 1432 we see that there are no ordered subsequences of 1432 that have the same relations between elements as 123.
+In this paper we investigate pattern avoidance in finite sequences of natural numbers.
+We begin with an example: consider the sequences 123 and 1324.
+Note the order relationships of the elements of 123: we have $1< 2 < 3$.
+Notice in addition that there is an ordered subsequence of 1324 that follows the same general pattern, namely 124.
+In this situation we say that the sequence 1324 \emph{does not avoid the pattern 123}.
+We can also examine the sequence 1432, and observe that none of its subsequences have the same order relationships as 123.
 We say that 1432 \emph{avoids} 123.
 
-With an example of a four element sequence that avoids 123 and a four element sequence that does not avoid 123, it is natural to ask the question, how many four element sequences avoid 123?
-As we will see below, the answer to this question is simply 5, but this leads to a more general question, how many $n$-element sequences avoid 123?
+With examples of a four element sequence that avoids 123 and a four element sequence that does not avoid 123, it is natural to ask the question: how many four element sequences avoid 123?
+And more generally, how many $n$-element sequences avoid 123?
 If we define $s_n(123)$ to be the number of $n$ element sequences that avoid 123, then the question above is equivalent to identifying the sequence $(s_n(123))_n$.
@@
 
-In section \ref{S3} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
-In section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$.
-In section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements.
-Before proceeding, we will formally introduce the notation that will be used throughout this paper.
+In Section \ref{S3} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
+In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$.
+In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements.
+Before proceeding, however, we formally introduce the notation that will be used throughout this paper.
 
@@
 
+\todo{We never define what a pattern is. Either scrub the paper of mentions or define it here}
+
 \begin{remark}
-The notation used to describe the elements of $S_n$ above is `one-line' notation.
-Thus far we have ignored the fact that $S_n$ has a group structure.
-When discussing this group structure it is convenient to use a different notation known as `cycle' notation.
-This alternative notation will be used in Section\ref{S4} and an introduction to cycle notation can be found in Appendix [].
+The notation used to describe the elements of $S_n$ above is `one-line' notation.\todo{Jerison wanted us to define this formally, I think} Thus far we have ignored the fact that $S_n$ has a group structure.
+When discussing this group structure it is convenient to use a different notation known as `cycle' notation.\todo{Also should define this formally} This alternative notation will be used in Section \ref{S4} and an introduction to cycle notation can be found in Appendix [].
 \end{remark}
@@
 \begin{definition}
-Two finite permutations $a_1,\cdots a_k$ and $b_1,\cdots b_k$ in $S_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
-\end{definition}
+Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
+\end{definition}\todo{I changed this definition... take a look and see if you agree with it}
 
 \begin{remark}
-An equivalent definition is that two permutations in $S_k$ have the same relative order if the elements in each position of the sequence satisfy the same set of inequalities. 
+Equivalently, two finite sequences of distinct positive integers have the same relative order if the elements in each position of the sequence satisfy the same set of inequalities.
 \end{remark}
@@
 \begin{definition}
-A permutation $a_1a_2\cdots a_n$ avoids a permutation $b_1\cdots b_k$ with $n\ge k$ if no subsequence $a_{i_1}a_{i_2}\cdots a_{i_k}$ of $a_1a_2\cdots a_n$ has its terms in the same relative order as $b_1\cdots b_k$.
+A finite sequence of distinct positive integers $a_1a_2\cdots a_n$ avoids another finite sequence of distinct positive integers $b_1\cdots b_k$ with $n\ge k$ if no subsequence $a_{i_1}a_{i_2}\cdots a_{i_k}$ of $a_1a_2\cdots a_n$ has its terms in the same relative order as $b_1\cdots b_k$.
 \end{definition}
@@
 
-We claim that the sequence of numbers $s_n(312)$ is in fact the sequence of Catalan numbers. We state this result formally as the following theorem,
+We claim that the sequence of numbers $s_n(312)$ is in fact the sequence of Catalan numbers. Before formally stating and proving the result, however, we prove the following useful lemma.
+\begin{lemma}
+\label{lemma1}
+The permutations of $\{1,2,\dots,k,k+1\}$ ending in $i$ that avoid the pattern $312$ are precisely those of the form,
+$$\pi_1 \pi_2 i$$
+the concatenation of $\pi_1, \pi_2$, and $i$, where $\pi_1$ is a permutation of $\{1,2,\ldots,i-1\}$ that avoids the pattern $312$ and $\pi_2$ is a permutation of $\{i+1,\ldots,k+1\}$ that avoids the pattern $312$.
+\end{lemma}\todo{Explained the notation for concatenation}
+
+\begin{proof}
+
+It is easy to see that such a permutation is sufficient for avoiding $312$ and ending in $i$. We will prove that the condition is necessary as well.
+
+First, we define the sets $A=\{1,2,\ldots,i-1\}$ and $B=\{i+1,i+2,\ldots,k+1\}$.
+For the sake of contradiction, let us assume that there exists some permutation $\pi$ of $\{1,2,...,k,k+1\}$ that ends with value $i$ such that some integer $x < i$ (that is, $x \in A$) is to the right of some integer $y > i$ ($y \in B$).
+Clearly, this permutation is not of the form described above, and the subsequence $(y, x, i)$ is of the same relative order as 312; as a consequence, $\pi$ does not avoid 312.
+
+It follows that for any permutation $\pi\in S_{k+1}$ avoiding $312$ and ending in $i$, all $x < i$ must be to the left of all $y > i$.
+Hence $\pi$ must be the concatenation of three subsequences $\pi_1 \pi_2 i$, where $\pi_1$ is a permutation of $\{1, 2, \ldots, i-1\}$ and $\pi_2$ is a permutation of $\{i+1, i+2, \ldots, k+1\}$.
+Furthermore, it is clear that any subsequences of the permutation $\pi = \pi_1 \pi_2 i$ must avoid $312$ if the entire permutation $\pi$ is to avoid $312$ as well; this implies that the permutations $\pi_1$ and $\pi_2$ must avoid $312$ as well.
+\end{proof}
+
+With this lemma proven, we move on to the proof of our main theorem for this section.
 
 \begin{theorem}
-The total number of permutations of $\{1,2,3,...,n\}$ that avoid the order $312$ as a subsequence is $C_n$ where $C_n$ is the $n^{th}$ Catalan number.
-\end{theorem}
-
-Before proving the theorem, we state and prove the following lemma, that will be used in our proof of the theorem.
-\begin{lemma}
-\label{lemma1}
-All permutations of $\{1,2,\dots,k,k+1\}$ ending in $i$ that avoid the order $312$ as a sub-sequence must be of the form,
-$$\pi_1 \pi_2 i$$
-where $\pi_1$ is a permutation of $\{1,2,\ldots,(i-1)\}$ that avoids the order $312$ as a sub-sequence and $\pi_2$ is a permutation of $\{(i+1),\ldots,(k+1)\}$ that avoids the order $312$ as a sub-sequence.
-\end{lemma}
-
-\begin{proof}
-It is clear that any subsequences of the permutation $\pi = \pi_1 \pi_2 i$ must avoid $312$ if the entire permutation $\pi$ is to avoid $312$ as well; this implies that the permutations $\pi_1$ and $\pi_2$ must avoid $312$ as well.
-
-We proceed with a proof by contradiction.
-
-Before proceeding, we define the sets $A$ and $B$ to be $\{1,2,\ldots,(i-1)\}$ and $\{(i+1),(i+2),\ldots,(k+1)\}$ respectively.
-For the sake of contradiction, let us assume that there exists some permutation $\pi$ of $\{1,2,...,k,k+1\}$ that ends with value $i$ such that some integer $x < i$ (that is, $x \in A$) is to the right of some integer $y > i$ ($y \in B$).
-Clearly, this permutation is not of the form described above.
-It is also easy to see that $\pi$ does \emph{not} avoid the order $312$ since the triple $(y, x, i)$ satisfies the condition $y > x > i$ and is in the order $312$.
-
-From this we conclude that all $x < i$ must be to the left of all $y > i$.
-This information coupled with the fact that we are only considering permutations that end with $i$ complete the claim that only permutations of the form described above can avoid $312$.
-
-\end{proof}
-
-With this lemma proven, we move on to the proof of our theorem.
+$s_n(312)$, the total number of permutations of $\{1,2,3,...,n\}$ that avoid the pattern $312$, is equal to $C_n$, the $n^{th}$ Catalan number.
+\end{theorem}\todo{should we define $C_n$ separately?}
 
@@
 
-Now, we need to prove the inductive case. Let us first assume that for all $i$ from $1$ to $k$, the number of permutations of $\{1,2,...,i\}$ that avoid the order $312$ as a subsequence is $C_i$.
+We need to prove the inductive case. Let us first assume that for all $i$ from $1$ to $k$, the number of permutations of $\{1,2,...,i\}$ that avoid the order $312$ as a subsequence is $C_i$.
 
-Now, we want to prove the inductive hypothesis for $\{1,2,...,k,k+1\}$ as well, that is the number of permutations of $\{1,2,...,k,k+1\}$ that avoid the order $312$ as a subsequence is $C_{k+1}$.
+Now, we want to prove the inductive hypothesis for $\{1,2,...,k,k+1\}$ as well.
+That is, we want to prove that the number of permutations of $\{1,2,...,k,k+1\}$ that avoid the order $312$ as a subsequence is $C_{k+1}$.
 
-We count the number of permutations of $\{1,2,...,k,k+1\}$ that avoid $312$ by enumerating through all possible values of the last term of a valid permutation.
-If the last term of the permutation is $i$ (where $i \in \{1,2,...,k,k+1\}$), then let us define the subsets $A$ and $B$ of the  set $\{1,2,...,k+1\} \setminus \{i\}$ as the set of integers less than $i$ and the set of integers greater than $i$ respectively.
+Using our lemma, we count the number of permutations of $\{1,2,...,k,k+1\}$ that avoid $312$ by enumerating through all possible values of the last term of a valid permutation.
+Consider $\pi$ a permutation of $\{1,2,\ldots, k, k+1\}$ ending in $i$.Let us define the subsets $A$ and $B$ of the  set $\{1,2,...,k+1\} \setminus \{i\}$ as the set of integers less than $i$ and the set of integers greater than $i$ respectively.
 It is clear from the definition of $A$ and $B$ that $A$ and $B$ are disjoint from each other.
 
-Now, from the above lemma, we know that all permutations of $\{1,2,...,k,k+1\}$ ending in $i$ that avoid $312$ must be of the form,
+From the above lemma, we know that all permutations of $\{1,2,...,k,k+1\}$ ending in $i$ that avoid $312$ are precisely those of the form
+$$\pi = \pi_1 \pi_2 i$$
+where $\pi_1$ is a permutation of $A$ that avoids $312$ and $\pi_2$ is a permutation of $B$ that avoids $312$.
 
-$$\pi = \pi_1 \pi_2 i$$
-
-where $\pi_1$ is a permutation of $A$ that avoids the order $312$ as a sub-sequence and $\pi_2$ is a permutation of $B$ that avoids the order $312$ as a sub-sequence.
-It is clear that the above permutation contains all integers between $1$ and $k+1$, from the definitions of the subsets $A$ and $B$, which implies that $\pi_1\pi_2i$ is permutation of the set $\{1,2,...,k,k+1\}$.
-
-Now, the total number of permutations $\pi$ is,
-$$n_{\pi_1} \cdot n_{\pi_2} = C_{i-1} \cdot C_{k-i+1}$$
-since the total number of valid permuations $\pi_1$ is simply going to be $C_{i-1}$ (total number of valid permutations of length $i-1$ that avoid the order $312$ as a sub-sequence is $C_{i-1}$; similarly $n_{\pi_2} = C_{k-1+1}$)
+It follows that the total number of permutations $\pi$ avoiding $312$ and ending in $i$ is
+$$C_{i-1} \cdot C_{k-i+1},$$
+since by our induction hypothesis the total number of valid choices for $\pi_1$ is simply $C_{i-1}$, and similarly the total number of valid choices for $\pi_2$ is $C_{k-1+1}$.
 
 Now, summing over all possible values of $i$, we see that the total number of permutations of $\{1,2,...,k+1\}$ that avoid $312$ is equal to,
-$$\sum_{i=1}^{k+1} C_{i-1} \cdot C_{k-i+1} = \sum_{i=0}^k C_i \cdot C_{k-i}$$ which is in fact $C_{k+1}$, and we are done.
+$$\sum_{i=1}^{k+1} C_{i-1} \cdot C_{k-i+1} = \sum_{i=0}^k C_i \cdot C_{k-i}$$ which is well-known to equal $C_{k+1}$, the $n+1$st Catalan number. The proof  therefore follows by induction.
 
@@
 In fact, we claim that if $\sigma$ and $\pi$ are both permutations in $S_k$ for some arbitrary $k$, then if $\sigma = flip(\pi)$, $s_n(\pi) = s_n(\sigma)$.
-This is easy to see since if $a$ is a permutation that avoids $\sigma$, then $flip(a)$ is a permutation that avoids $\pi$ -- since this can be done both ways, it is bijective and we can conclude that the cardinalities of the two sets are equal, or $s_n(\sigma) = s_n(\pi)$.
+This is easy to see since if $a$ is a permutation that avoids $\sigma$, then $flip(a)$ is a permutation that avoids $\pi$ -- since this can be done both ways, it is bijective and we can conclude that the cardinalities of the two sets are equal, or $s_n(\sigma) = s_n(\pi)$.\todo{we should move this up and make it clear from the beginning that this stuff generalizes}
 
@@
 Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number.
-In order to prove this result, however, we will need the machinery of Young tableaux.
+In order to prove this result, we need the machinery of Young tableaux.
 What follows is a brief exposition of the key results, adapted from Stanley's \emph{Algebraic Combinatorics}.
@@
 There exists an algorithm (called the \emph{RSK algorithm}) which defines a bijection between the set of permutations in $S(n)$ avoiding a decreasing subsequence of length 3 and the set of pairs of size-$n$ SYT of the same shape and at most two rows.
-\end{theorem}
+\end{theorem}\todo{give a citation for this}
 Using all this machinery and all these results, we are finally able to prove the closed-form expression for $s_n(321)$.
@@
 Beginning with a pair of size $n$ Young tableaux of shape $\lambda$ and at most two parts, it is easy to see that this process always yields a valid Young tableau of shape $(n, n)$.
-Indeed, we can also define the inverse: given any Young tableau for the partition $(n, n)$, we can find the shape formed by all numbers greater than $n$: split it off from the original tableau, rotate it, and invert the numbers to get a pair of size $n$ Young tableau of the same shape and at most two parts.
+Indeed, we can define its inverse: given any Young tableau for the partition $(n, n)$, we can find the shape formed by all numbers greater than $n$: split it off from the original tableau, rotate it, and invert the numbers to get a pair of size $n$ Young tableau of the same shape and at most two parts.
 These processes can be easily checked to be inverses: hence the two sets are of the same size, and the proof follows.
@@
 
-In this section we will present our findings and conjectures about sequences $s_n(\pi)$ for $\pi\in S_4$.
-We computed values up to $s_{10}$.
+In this section we present our findings and conjectures about sequences $s_n(\pi)$ for $\pi\in S_4$.
+We computed values up to $s_{9}$.
 It appears that there are three sequences that appear, and we have sorted the permutations according to which sequence they produce and paired them with their flips.
 
-We define the following sequences based on the first 8 terms:
+We define the following sequences based on the first nine terms:
 $$

### 2014-11-06 project2/paper/paper.tex

@@
 When discussing this group structure it is convenient to use a different notation known as `cycle' notation.\todo{Also should define this formally} This alternative notation will be used in Section \ref{S4} and an introduction to cycle notation can be found in Appendix [].
+
+The notation used to describe the elements of $S_n$ above is `one-line' notation.
+Thus far we have ignored the fact that $S_n$ has a group structure.
+When discussing this group structure it is convenient to use a different notation known as `cycle' notation.
+This alternative notation will be used in Section \ref{S4} and an introduction to cycle notation can be found in Appendix [].
 \end{remark}
@@
 
-Before proving that there exists a bijective mapping between all permutations in $T_n$ that avoid $312$ (call this set $A$) and all permutations in $T_n$ that avoid $321$ (call this set $B$), we establish a chain of transformation moves that can be used to transform an element $a \in A$ to an element $b \in B$.
+We now construct a bijective mapping between all permutations in $T_n$ that avoid $312$ and all permutations in $T_n$ that avoid $321$.
 
@@
 For each permutation $\pi \in T_n$ that avoids $312$, we want to show that there exists a sequence of transformation moves that transforms $\pi$ to a unique permutation $\pi'$ that avoids $321$.
+
+\textbf{May want to skip the transformation stuff, and directly talk about the bijection being established.}
 
@@
 
-We now prove that there exists a bijection between the sets $A$ and $B$.
-
-Let $a$ be an arbitrary permutation in $A$.
-Clearly, by the definition of $A$, $a$ must avoid $312$.
-Now, we claim that for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in decreasing order; if this were not the case, then we would have a subsequence in the relative ordering $312$, which would be a contradiction.
-It is easy to see that we can make a corresponding claim about permutation $b$, that is for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in increasing order.
-
-The above two claims shows there's actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$ -- to obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order); similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order).
-Since the mapping from $A$ to $B$ is clearly reversible, and hence invertible, we conclude that the mapping we just established between $A$ and $B$ is bijective.
+What remains to be proven is that the order of swapping operations in the chain of transformations does not matter -- that is, any valid transformation chain starting with some permutation $\pi \in T_n$ that avoids $312$, will always produce the same permutation $\pi' \in T_n$ that avoids $321$.
+Note that since we're only swapping even elements, the odd elements are in exactly the same position in both $\pi$ and $\pi'$.
+In addition, for every element $i \in \pi$, all even numbers in $\pi$ to the right of $i$ and less than $i$ must be in decreasing order; in $\pi'$, for every element $i \in \pi'$, all even numbers in $\pi'$ to the right of $i$ and less than $i$ must be in increasing order.
 

### 2014-11-07 project2/paper/paper.tex
Resolved some TODOs and cleaned up some notation

@@
 
-In Section \ref{S3} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
-In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$.
+In Section \ref{S3} we consider sequences $s_n(\pi)$ when $\pi$ is a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
+In Section \ref{S4} we will consider sequences $s_n(\pi)$ when $\pi$ is a permutation of $\{1,2,3,4\}$.
 In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements.
-Before proceeding, however, we formally introduce the notation that will be used throughout this paper.
+Before proceeding, however, we formally introduce the notation that will be used through the rest of this paper.
 
@@
 Then if $\sigma_p$ is the parent of $\sigma$, $\mathcal{R}(\sigma_p)$ avoids $\mathcal{R}(\pi)$.
-Now if we add $n$ to the permutation $\mathcal{R}(\sigma_p)$ to construct $\mathcal{F}(\sigma)$, we see that $\mathcal{F}(\sigma)$ must avoid $\mathcal{F}(\pi)$, for if it did not, we could reverse one more time and see obtain notice that $\sigma\in S_n$ does not avoid $\pi$, a contradiction.
+Now if we add $n$ to the permutation $\mathcal{R}(\sigma_p)$ to construct $\mathcal{F}(\sigma)$, we see that $\mathcal{F}(\sigma)$ must avoid $\mathcal{F}(\pi)$, for if it did not, we could reverse one more time and obtain the result that $\sigma\in S_n$ does not avoid $\pi$, a contradiction.
 The other direction follows because $\mathcal{R}^2=1$.
@@
 the concatenation of $\pi_1, \pi_2$, and $i$, where $\pi_1$ is a permutation of $\{1,2,\ldots,i-1\}$ that avoids the pattern $312$ and $\pi_2$ is a permutation of $\{i+1,\ldots,k+1\}$ that avoids the pattern $312$.
-\end{lemma}\todo{Explained the notation for concatenation}
+\end{lemma}
 
@@
 
-With this lemma proven, we move on to the proof of our main theorem for this section.
+Before we move on to the proof of our main theorem for this section, we look at the sequence of Catalan numbers.
+
+\begin{definition}
+The Catalan numbers are the sequence of positive integers $C_i$ defined as follows,
+$$C_0=1, \, C_{n+1}=\sum_{i=1}^n C_iC_{n-i} \, \text{for} \; n \geq 0$$
+\end{definition}
 
@@
 $s_n(312)$, the total number of permutations of $\{1,2,3,...,n\}$ that avoid the pattern $312$, is equal to $C_n$, the $n^{th}$ Catalan number.
-\end{theorem}\todo{should we define $C_n$ separately?}
+\end{theorem}
 
@@
 Using our lemma, we count the number of permutations of $\{1,2,...,k,k+1\}$ that avoid $312$ by enumerating through all possible values of the last term of a valid permutation.
-Consider $\pi$ a permutation of $\{1,2,\ldots, k, k+1\}$ ending in $i$.Let us define the subsets $A$ and $B$ of the  set $\{1,2,...,k+1\} \setminus \{i\}$ as the set of integers less than $i$ and the set of integers greater than $i$ respectively.
+Consider $\pi$ a permutation of $\{1,2,\ldots, k, k+1\}$ ending in $i$.
+Let us define the subsets $A$ and $B$ of the  set $\{1,2,...,k+1\} \setminus \{i\}$ as the set of integers less than $i$ and the set of integers greater than $i$ respectively.
 It is clear from the definition of $A$ and $B$ that $A$ and $B$ are disjoint from each other.
@@
 
-Before proving that there exists a bijective mapping between all permutations in $T_n$ that avoid $312$ (call this set $A$) and all permutations in $T_n$ that avoid $321$ (call this set $B$), we establish a chain of transformation moves that can be used to transform an element $a \in A$ to an element $b \in B$.
+Before proving that there exists a bijective mapping between the set of permutations in $T_n$ that avoid $312$ (call this set $A$) and the set of permutations in $T_n$ that avoid $321$ (call this set $B$), we establish a chain of transformation moves that can be used to transform an element $a \in A$ to an element $b \in B$.
 
@@
 
-Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above, however we do prove that both the sequences $(t_n(231))$ and $(t_n(213)$ are equal to the sequence $1, 2, 4, 8, \ldots$ below.
+Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above, however we do prove that both the sequences $(t_n(231))$ and $(t_n(213)$ are equal to the sequence $1, 2, 4, 8, \ldots$, and hence equivalent, below.
 
@@
 Consider an arbitrary even element $2i$ in $\sigma$.
-For this element $2i$, define $A$ to be the set of numbers greater than $2i$ and to the left of $2i$ in the permutation $\sigma$.
-First off, note that the set $A$ cannot contain any even numbers, because if it did, then the permutation $\sigma$ would contain a three-element sub-sequence in the relative ordering $231$ -- if $a$ is an even number in the set $A$, then $(a-1, a, 2i)$ would be a three-element sub-sequence of $\sigma$ in the relative ordering $231$ (Note that $a-1$ must be to the left of $a$ in the permutation $\sigma$ because of the way the set $T_{2m}$ was constructed.
+For this element $2i$, define $A$ to be the set of numbers greater than $2i$ and to its left in the permutation $\sigma$.
+First off, note that the set $A$ cannot contain any even numbers, because if it did, then the permutation $\sigma$ would contain a three-element sub-sequence in the relative ordering $231$ -- if $a$ is an even number in the set $A$, then $(a-1, a, 2i)$ would be a three-element sub-sequence of $\sigma$ in the relative ordering $231$ (Note that $a-1$ must be to the left of $a$ in the permutation $\sigma$ because of the way the set $T_{2m}$ was constructed)
 
-Furthermore, the set $A$ cannot contain any odd elements greater than $2i+1$, because if it did, then again the permutation $\sigma$ would contain a three-element subsequence in the relative ordering $231$ -- if $a > 2i+1$ is an odd number in $A$, then $(2i+1, a, 2i)$ is a three-element sub-sequence of $\sigma$ in the relative ordering $231$ Therefore, we conclude that the set $A$ either only contains the element $2i+1$, or is empty.
+Furthermore, the set $A$ cannot contain any odd elements greater than $2i+1$, because if it did, then again the permutation $\sigma$ would contain a three-element subsequence in the relative ordering $231$ -- if $2j+1 > 2i+1$ is an odd number in $A$, then $(2i+1, 2j+1, 2i)$ is a three-element sub-sequence of $\sigma$ in the relative ordering $231$ Therefore, we conclude that the set $A$ either only contains the element $2i+1$, or is empty.
 
@@
 
-Combining all these facts together, we see that the number of permutations $\sigma$ in $T_n$ that avoid $231$ must be equal to $2^{(2m-2)/2} = 2^{m-1}$, as desired.
+Combining all these facts together, we see that the number of permutations $\sigma$ in $T_{2m}$ that avoid $231$ must be equal to $2 \cdot 2 \cdot \ldots \cdot 2 = 2^{m-1}$, since every even element can be inserted in exactly two positions in $\sigma$.
 
@@
 Clearly, $j < 2m -1 < 2m$ (since every element in $\sigma$ other than $2m$ is less than $2m-1$), which means $(2m - 1, j, 2m)$ are in the relative ordering $213$.
-We conclude that $2m-1$ and $2m$ must be consecutive.
+We conclude that $2m-1$ and $2m$ must be consecutive in $\sigma$.
 
 Furthermore, we see that we can construct $A_m$ from $A_{m-1}$ by inserting the consecutive elements $2m-1, 2m$ into a permutation $\sigma'$ in $A_{m-1}$.
-Furthermore, we claim that we get a valid permutation $\sigma$ in $A_m$ from $\sigma'$ only if we insert the consecutive elements $2m-1, 2m$ right after either $2m-3$ or $2m-2$ in $\sigma'$ (which themselves are consecutive in $\sigma'$ from the above claim).
+We claim that we get a valid permutation $\sigma$ in $A_m$ from $\sigma'$ only if we insert the consecutive elements $2m-1, 2m$ right after either $2m-3$ or $2m-2$ in $\sigma'$ (which themselves are consecutive in $\sigma'$ from the above claim).
 
 First, note that $2m-1, 2m$ must be inserted to the right of both $2m-3$ and $2m-2$, because of the way $T_{2m}$ is defined.
-Furthermore, if we insert $2m-1, 2m$ right after some $a$ in $\sigma$ not equal to $2m-3$ and $2m-2$, then we see that the produced permutation $\sigma$ has a three-element sub-sequence in the relative ordering $213$ -- $(2m-2, a, 2m-1)$.
+Furthermore, if we insert $2m-1, 2m$ right after some $a$ in $\sigma$ not equal to $2m-3$ or $2m-2$, then we see that the produced permutation $\sigma$ has a three-element sub-sequence in the relative ordering $213$ -- $(2m-2, a, 2m-1)$.
 Hence we conclude that $2m-1, 2m$ must be inserted to the right of either $2m-3$ or $2m-2$; since we can do this for every permutation $\sigma'$ in $A_{m-1}$, we conclude that $|A_m| = 2|A_{m-1}|$.
@@
 
-Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $1, 2, 4, 8, \ldots$ are vastly different.
+Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $1, 2, 4, 8, \ldots$ are vastly different -- this is an interesting topic of research that we would like to pursue as future work.
 

### 2014-11-07 project2/paper/paper.tex
Resolved TODO, and corrected typo

@@
 
-With examples of a four element sequence that avoids 123 and a four element sequence that does not avoid 123, it is natural to ask the question: how many four element sequences avoid 123?
+With examples of a four element sequence that avoids 123 and a four element sequence that does not avoid 123, it is natural to ask the question: how many $4$-element sequences avoid 123?
 And more generally, how many $n$-element sequences avoid 123?
@@
 Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
-\end{definition}\todo{I changed this definition... take a look and see if you agree with it}
+\end{definition}
 
@@
 \begin{proof}
-Notice that we only need to proof one direction of this statement.
+Notice that we only need to prove one direction of this statement.
 The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation.
@@
 We proceed by induction.
+
 Certainly the statement is true for elements of $S_1$.

### 2014-11-07 project2/paper/paper.tex
edits to introduction

--- 2014-11-06 project2/paper/paper.tex
@@
 And more generally, how many $n$-element sequences avoid 123?
+For a sequence $\pi$ that is a permutation of the elements of $\{1,\cdots,k\}$, we define
+$$
+s_n(\pi)=\#\{\text{the set of permutations of $\{1,\cdots, n\}$ that avoid $\pi$}\}.
+$$
+
 If we define $s_n(123)$ to be the number of $n$ element sequences that avoid 123, then the question above is equivalent to identifying the sequence $(s_n(123))_n$.
@@
 
-In Section \ref{S3} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
+In Sections \ref{S3} and \ref{S4} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
 In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$.
@@
 \subsection{Notation}
+
+\begin{definition}
+A permutation of a set $D$ is a bijective map from $D$ to itself. 
+\end{definition}
+
+\begin{example}
+Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=2$, $\pi(3)=4$, $\pi(4)=1$, and $\pi(5)=3$. Then $\pi$ is a permutation of the set $D=\{1,2,3,4,5\}$. 
+\end{example}
+
+\begin{remark}
+There are two notations for permutations, `one-line' notation and `cycle' notation. We give definitions of both below. Unless stated otherwise, we will use one-line notation in this paper.
+\end{remark}
+
+\begin{definition}
+Let $\pi$ be the permutation on elements of $\{1,2,\cdots, n\}$ so that $\pi(1)=a_1, \pi(2)=a_2,\cdots, \pi(n)=a_n$.
+Then in one-line notation $\pi$ is written as $\pi(1)\pi(2)\cdots\pi(n)$, or equivalently, $a_1a_2,\cdots a_n$.
+\end{definition}
+
+\begin{definition}
+In cycle notation we identify cycles in the permutation and write them as such.
+Let $\pi$ be the permutation on elements of $\{1,2,\cdots, n\}$ so that $\pi(1)=a_1, \pi(2)=a_2,\cdots, \pi(n)=a_n$.
+Then the cycle notation for $\pi$ is $$(1 \ \pi(1) \ \pi(\pi(1)) \cdots \pi^{k-1}(1))(\alpha_1 \ \pi(\alpha_1) \ \pi(\pi(\alpha_1)) \cdots)\cdots(\alpha_\ell \ \pi(\alpha_\ell) \ \pi(\pi(\alpha_\ell)) \cdots).$$ Let $k$ be the smallest natural number for which $\pi^k(1)=1$.
+Then $\pi^{k-1}(1)$ is the last entry in the cycle before it is closed.
+If there are elements of $\{1,2,\cdots, n\}$ that are not in the cycle then let $\alpha_1$ be the smallest such element and continue the process until all elements are accounted for.
+\end{definition}
+
+\begin{remark}
+Note, cycle notation is not unique. 
+\end{remark}
+
+\begin{example}
+Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=4$, $\pi(3)=1$, $\pi(4)=2$, and $\pi(5)=3$. Then in one-line notation $\pi=54123$. In cycle notation $\pi=(153)(24)$. 
+\end{example}
 
@@
 
-\begin{remark}
-The notation used to describe the elements of $S_n$ above is `one-line' notation.\todo{Jerison wanted us to define this formally, I think} Thus far we have ignored the fact that $S_n$ has a group structure.
-When discussing this group structure it is convenient to use a different notation known as `cycle' notation.\todo{Also should define this formally} This alternative notation will be used in Section \ref{S4} and an introduction to cycle notation can be found in Appendix [].
+%Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$.
+There are two conventions commonly used to write elements of $S_n$.
+The first is $\emph{one-line}$ notation where permutations in $S_n$ are written explicitly, e.g.
+43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.
 
-The notation used to describe the elements of $S_n$ above is `one-line' notation.
-Thus far we have ignored the fact that $S_n$ has a group structure.
-When discussing this group structure it is convenient to use a different notation known as `cycle' notation.
-This alternative notation will be used in Section \ref{S4} and an introduction to cycle notation can be found in Appendix [].
-\end{remark}
+%The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
+For example, the cycle $(14)(23)$ sends $12345$ to $43215$.
 
@@
 
-\section{Appendix A}
-
-Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$.
-There are two conventions commonly used to write elements of $S_n$.
-The first is $\emph{one-line}$ notation where permutations in $S_n$ are written explicitly, e.g.
-43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.
-
-The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
-For example, the cycle $(14)(23)$ sends $12345$ to $43215$.
-
 \end{document}

### 2014-11-07 project2/paper/paper.tex

--- 2014-11-07 project2/paper/paper.tex
@@
 
-In Section \ref{S3} we consider sequences $s_n(\pi)$ when $\pi$ is a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
-In Section \ref{S4} we will consider sequences $s_n(\pi)$ when $\pi$ is a permutation of $\{1,2,3,4\}$.
+In Sections \ref{S3} and \ref{S4} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
+In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$.
 In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements.
-Before proceeding, however, we formally introduce the notation that will be used through the rest of this paper.
+Before proceeding, however, we formally introduce the notation that will be used throughout this paper.
 
 \subsection{Notation}
+
+\begin{definition}
+A permutation of a set $D$ is a bijective map from $D$ to itself. 
+\end{definition}
+
+\begin{example}
+Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=2$, $\pi(3)=4$, $\pi(4)=1$, and $\pi(5)=3$. Then $\pi$ is a permutation of the set $D=\{1,2,3,4,5\}$. 
+\end{example}
+
+\begin{remark}
+There are two notations for permutations, `one-line' notation and `cycle' notation. We give definitions of both below. Unless stated otherwise, we will use one-line notation in this paper.
+\end{remark}
+
+\begin{definition}
+Let $\pi$ be the permutation on elements of $\{1,2,\cdots, n\}$ so that $\pi(1)=a_1, \pi(2)=a_2,\cdots, \pi(n)=a_n$.
+Then in one-line notation $\pi$ is written as $\pi(1)\pi(2)\cdots\pi(n)$, or equivalently, $a_1a_2,\cdots a_n$.
+\end{definition}
+
+\begin{definition}
+In cycle notation we identify cycles in the permutation and write them as such.
+Let $\pi$ be the permutation on elements of $\{1,2,\cdots, n\}$ so that $\pi(1)=a_1, \pi(2)=a_2,\cdots, \pi(n)=a_n$.
+Then the cycle notation for $\pi$ is $$(1 \ \pi(1) \ \pi(\pi(1)) \cdots \pi^{k-1}(1))(\alpha_1 \ \pi(\alpha_1) \ \pi(\pi(\alpha_1)) \cdots)\cdots(\alpha_\ell \ \pi(\alpha_\ell) \ \pi(\pi(\alpha_\ell)) \cdots).$$ Let $k$ be the smallest natural number for which $\pi^k(1)=1$.
+Then $\pi^{k-1}(1)$ is the last entry in the cycle before it is closed.
+If there are elements of $\{1,2,\cdots, n\}$ that are not in the cycle then let $\alpha_1$ be the smallest such element and continue the process until all elements are accounted for.
+\end{definition}
+
+\begin{remark}
+Note, cycle notation is not unique. 
+\end{remark}
+
+\begin{example}
+Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=4$, $\pi(3)=1$, $\pi(4)=2$, and $\pi(5)=3$. Then in one-line notation $\pi=54123$. In cycle notation $\pi=(153)(24)$. 
+\end{example}
 

### 2014-11-07 project2/paper/paper.tex
fixed things in introduction

@@
 
-In Sections \ref{S3} and \ref{S4} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
-In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$.
+In Sections \ref{312} and \ref{321} we consider sequences $s_n(\pi)$ for $\pi$ an element of  \\ $\{123,132,213,231,312,321\}$, and prove the following theorem.
+
+\begin{theorem*}
+For $\pi$ an element of $\{123,132,213,231,312,321\}$, the sequence $(s_n(\pi))$ is the Catalan numbers.
+\end{theorem*}
+
+Of particular interest is the fact that two entirely distinct methods of proof had to be used to prove this theorem for all six permutations.
+In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$ and state a conjecture concerning $(s_n(\pi))$.
 In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements.
@@
 
+\begin{remark}
+There are two notations for permutations, `one-line' notation and `cycle' notation. We define both below. Unless stated otherwise, we will use one-line notation in this paper.
+\end{remark}
+
 \begin{example}
-Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=2$, $\pi(3)=4$, $\pi(4)=1$, and $\pi(5)=3$. Then $\pi$ is a permutation of the set $D=\{1,2,3,4,5\}$. 
+Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=4$, $\pi(3)=1$, $\pi(4)=2$, $\pi(5)=3$, and $\pi(6)=6$. Then $\pi$ is a permutation of the set $D=\{1,2,3,4,5\}$. 
 \end{example}
-
-\begin{remark}
-There are two notations for permutations, `one-line' notation and `cycle' notation. We give definitions of both below. Unless stated otherwise, we will use one-line notation in this paper.
-\end{remark}
 
@@
 
+\begin{example}
+Using $\pi$ from the previous example, in one-line notation $\pi=541236$.
+\end{example}
+
 \begin{definition}
-In cycle notation we identify cycles in the permutation and write them as such.
-Let $\pi$ be the permutation on elements of $\{1,2,\cdots, n\}$ so that $\pi(1)=a_1, \pi(2)=a_2,\cdots, \pi(n)=a_n$.
-Then the cycle notation for $\pi$ is $$(1 \ \pi(1) \ \pi(\pi(1)) \cdots \pi^{k-1}(1))(\alpha_1 \ \pi(\alpha_1) \ \pi(\pi(\alpha_1)) \cdots)\cdots(\alpha_\ell \ \pi(\alpha_\ell) \ \pi(\pi(\alpha_\ell)) \cdots).$$ Let $k$ be the smallest natural number for which $\pi^k(1)=1$.
-Then $\pi^{k-1}(1)$ is the last entry in the cycle before it is closed.
-If there are elements of $\{1,2,\cdots, n\}$ that are not in the cycle then let $\alpha_1$ be the smallest such element and continue the process until all elements are accounted for.
+A cycle $(a_1\cdots a_n)$ with $1\le a_i\le n$ acts as a permutation on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
 \end{definition}
 
-\begin{remark}
-Note, cycle notation is not unique. 
-\end{remark}
-
-\begin{example}
-Let $\pi$ be a function so that $\pi(1)=5$, $\pi(2)=4$, $\pi(3)=1$, $\pi(4)=2$, and $\pi(5)=3$. Then in one-line notation $\pi=54123$. In cycle notation $\pi=(153)(24)$. 
-\end{example}
+A more formal definition of cycle notation is especially cumbersome, so we proceed with the example above, a permutation $\pi$ with $\pi(1)=5$, $\pi(2)=4$, $\pi(3)=1$, $\pi(4)=2$, $\pi(5)=3$, and $\pi(6)=6$.
+To write $\pi$ in cycle notation we start with 1 and identify the cycle $1\mapsto 5\mapsto 3\mapsto 1$.
+Once we come back to 1 we close the cycle and write $(135)$.
+We keep identifying cycles until every number is accounted for.
+So $2\mapsto 4\mapsto 2$ gives us $(24)$ and $6\mapsto 6$ gives us $(6)$.
+Finally we can write $\pi$ as the composition of functions $(135)(24)(6)$.
+Note, cycle notation is not necessarily unique.
 
@@
 \todo{We never define what a pattern is. Either scrub the paper of mentions or define it here}
-
-%Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$.
-There are two conventions commonly used to write elements of $S_n$.
-The first is $\emph{one-line}$ notation where permutations in $S_n$ are written explicitly, e.g.
-43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.
-
-%The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
-For example, the cycle $(14)(23)$ sends $12345$ to $43215$.
 
@@
 \end{definition}
-
-%\begin{remark}
-%Let $S_n$ be the set of permutations of $\{1,\cdots, n\}$.
-There are two conventions commonly used to write elements of $S_n$.
-The first is $\emph{one-line}$ notation where permutations in $S_n$ are written explicitly, e.g.
-43215 is a permutation of $\{1,2,3,4,5\}$ in $S_5$.
-%The second is \emph{cycle notation}, where a cycle $(a_1\cdots a_n)\in S_n$ with $1\le a_i\le n$ acts on a sequence by sending $a_1\mapsto a_2$, $a_2\mapsto a_3$, $\cdots$, $a_n\mapsto a_1$.
-For example, the cycle $(14)(23)$ sends $12345$ to $43215$.
-Unless otherwise stated, we will refer to permutations in one-line notation.
-%\end{remark}
 
@@
 \section{Terms in the sequence $s_n(312)$}
-\label{S3}
+\label{312}
 
@@
 \section{Terms in the sequence $s_n(321)$}
-
+\label{321}
 Unfortunately, there is no clear way to reformulate Lemma \ref{lemma1} for the case of avoiding the permutation 321 or 123.

### 2014-11-07 project2/paper/paper.tex
FIxed label

--- 2014-11-07 project2/paper/paper.tex
@@
 
-In Sections \ref{S3} and \ref{S4} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
+In Sections \ref{S3} and \ref{S32} we consider sequences $s_n(\pi)$ for $\pi$ a permutation on the set $\{1,2,3\}$, and prove that the sequence $s_n(\pi)$ for all such $\pi$ is in fact the sequence of Catalan numbers.
 In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$.
@@
 \section{Terms in the sequence $s_n(321)$}
+\label{S32}
 

### 2014-11-07 project2/paper/paper.tex
added flipping and reversing to section5

--- 2014-11-07 project2/paper/paper.tex
@@
 \section{Terms in the sequence $s_n(321)$}
+
 \label{321}
@@
 In this section we present our findings and conjectures about sequences $s_n(\pi)$ for $\pi\in S_4$.
-We computed values up to $s_{9}$.
-It appears that there are three sequences that appear, and we have sorted the permutations according to which sequence they produce and paired them with their flips.
+Using the flipping and reversing operators we can group elements of $S_4$ by their orbits.
+The following groups appear:
+$$
+\{1234,4321\},
+\{1243,4312,2134,3412\},
+\{1432,4123,2341,3214\},
+\{2143,3412\},
+$$
+$$
+\{4132,1423,2314,3241\},
+\{4213,1342,3124,2431\},
+\{2413,3142\}
+$$
+$$
+\{4231,1324\}
+$$
+So all permutations that belong to a single group in the list above must produce the same sequence. 
+We computed values up to $s_{9}$. It appears that there are three sequences that appear, and we have sorted the permutations according to which sequence they produce and paired them with their flips.
 
@@
 
+Comparing this table to the orbit decomposition above we notice that all of the groups in the first line appear to produce sequence $B$, all of the groups in the second line appear to produce sequence $A$, and the single orbit in line three produces sequence $C$.
 There is no guarantee that these sequences will not diverge and produce more than three distinct sequences, however for now we hypothesize that $A$, $B$, and $C$ are the only sequences that arise for $s_n(\pi)$ with $\pi\in S_4$.
@@
 \hline
-(1),(14)(23)&(243),(142)&(23),(14)\\
+(1)(2)(3)(4),(14)(23)&(243),(142)&(23),(14)\\
 (34),(1423)&(234),(143)&\\
@@
 So we conjecture that sequence $A$ is characterized by elements of $S_4$ that are three cycles or four cycles when written in cycle notation.
-Sequence $B$ seems to be characterized by either a pair of disjoint two cycles paired with a flip that is a 2-cycle, or a single 4-cycle paired with a flip that is a 2-cycle.
+Sequence $B$ seems to be characterized by either a pair of disjoint two cycles paired with a flip that is a pair of disjoint 2-cycles, or a single 4-cycle paired with a flip that is a 2-cycle.
 Sequence $C$ appears to be characterized by a 2-cycle paired with a 2-cycle as a flip.

### 2014-11-07 project2/paper/paper.tex
some edits to section 6

--- 2014-11-07 project2/paper/paper.tex
@@
 \label{Tn}
-We now impose further restrictions on the set of permutations $S_n$. This new set of permutations, $T_n$ is defined follows.
+We now impose further restrictions on the set of permutations $S_n$. This new set of permutations, $T_n$ is defined as follows.
 
 \begin{definition}
-For even $n$, $T_n$ is defined as the set of all permutations $\sigma \in S_n$ for which $1,3,5,\ldots,2n-1$ appear in increasing order, and $2i$ always appears to the right of $2i-1$.
+Let $n$ be an even integer.
+The set $T_n$ is defined as the set of all permutations $\sigma \in S_n$ in which $1,3,5,\ldots,2n-1$ appear in increasing order, and $2i$ always appears to the right of $2i-1$.
 \end{definition}
 
-Since $T_n$ is only defined for even $n$, henceforth we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $0$.
+Since $T_n$ is only defined for even $n$, henceforth we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $0$.\todo{0 or 1 here?}
 
@@
 \begin{proof}
-Observe that since $1,3,5,\ldots,2m-1$ must appear in increasing order in $T_{2m}$, our problem is now reduced to determining the relative order of $2,4,6,\ldots,2m$ with respect to each other, as well as with respect to $1,3,5,\ldots,2m-1$.
+Because $1,3,5,\ldots,2m-1$ must appear in increasing order in $T_{2m}$, the size of $T_{2m}$ is given by the number of ways in which $2, 4, 6, \ldots, 2m$ can be ordered relative to each other and to $1, 3, \ldots, 2m-1$.
 
-Let us first try to insert $2m$ into the sequence $1,3,5,\ldots,2m-1$.
-Clearly, the way $T_{2m}$ is defined, $2m$ can be inserted into only $1$ slot, the one following $2m-1$.
-(Here we define a slot as a gap between two existing elements of the sequence, or the gap that follows the last element of the sequence or the gap that precedes the first element of the sequence)
+Let us first insert $2m$ into the sequence $1,3,5,\ldots,2m-1$.
+Because $2m$ must appear to the right of $2m-1$, it can be inserted into only a single slot, the one following $2m-1$.
+(By ``slot'' we mean a gap between two existing elements of the sequence, or the gap that follows the last element of the sequence, or the gap that precedes the first element of the sequence.)
 
-Now, let us try to insert $(2m-2)$ into the sequence.
-$(2m-2)$ can be inserted into $3$ slots in the sequence -- the one between $2m-3$ and $2m-1$, the one between $2m-1$ and $2m$ or the one following $2m$.
+Next let us insert $(2m-2)$ into the sequence. $(2m-2)$ can be inserted into $3$ slots in the sequence -- the one between $2m-3$ and $2m-1$, the one between $2m-1$ and $2m$ or the one following $2m$.
 
-Now, if we try to insert $(2m-4)$ into this incompletely formed sequence, we will see that there are $5$ possible locations into which this number can be inserted.
-Continuing this for all even numbers upto $2$, we observe that the total number of ways such a sequence can  be created is equal to $1 \cdot 3 \cdot 5 \cdot \ldots \cdot (2m-1)$, as desired.
+We want to next insert $(2m-4)$ into this incompletely formed sequence; we see that there are $5$ possible locations.
+Continuing this for all even numbers up to $2$, we observe that the total number of ways such a sequence can  be created is equal to $1 \cdot 3 \cdot 5 \cdot \ldots \cdot (2m-1)$, as desired.
 \end{proof}
 
-Studying avoidance of permutations in $T_n$ seems an interesting exercise.
-In the following section, we consider the sequence of numbers $t_n(\pi)$, defined as the number of permutations in $T_n$ that avoid the permutation $\pi$.
-In particular, we consider permutations $\pi \in S_3$.
+Studying avoidance of permutations in $T_n$ seems an interesting exercise. In this section, we consider the sequence of numbers $t_n(\pi)$, defined as 
+\[t_n(\pi)=\#\{\sigma\in T_n|\sigma \text{ avoids } \pi\},\]
+with $\pi\in S_3$.
 
@@
 
-More interesting is the equivalences between $213$ and $231$, and $312$ and $321$.
-We will prove the equivalence between $312$ and $321$ below, but before doing so, we first state and prove the following lemmas which will be useful in our proof.
+More interesting are the equivalences between $213$ and $231$, and $312$ and $321$.
+We  prove the equivalence between $312$ and $321$ below, but before doing so, we first state and prove the following useful lemmas.
 
@@
 \label{tn_avoidance}
-If $a$ and $b$ are two elements that appear in a permutation in $T_n$ such that $a$ is to the left of $b$, then if $a > b$, $b$ must be even.
+Let $a$ and $b$ be two elements that appear in a permutation in $T_n$ such that $a$ is to the left of $b$. If $a > b$, then $b$ must be even.
 \end{lemma}
@@
 \begin{proof}
-For the sake of contradiction, let us assume that $b$ is odd. Then there exist two possibilities that need to be considered -- one, that $a$ is odd, and two, that $a$ is even.
+For the sake of contradiction, let us assume that $b$ is odd. Then there exist two possibilities that need to be considered -- either $a$ is odd or $a$ is even.
 
@@
 So if $a$ appears to the left of $b$, and both $a$ and $b$ are odd, then $a$ must be less than $b$ -- a contradiction.
-Hence, we conclude that $a$ can't be odd.
+Hence $a$ cannot be odd.
 
@@
 For all permutations $\sigma$ in $T_n$, every even element $j$ must be to the right of all odd elements less than $j$.
-From this, we conclude that since $a$ is even, it must be to the right of all odd numbers less than it.
-But $b$ is an odd number less than $a$ -- we conclude that $a$ can't be even either.
+Since $a$ is even, therefore, it must be to the right of all odd numbers less than it.
+But $b$ is an odd number less than $a$, and $a$ is to the left of $b$ -- another contradiction.
+Hence $a$ cannot be even, either.
 \end{itemize}
 
-Since $a$ can't be odd or even if $b$ is odd, we conclude that $b$ must be even, as desired.
+We conclude that $b$ must be even, as desired.
 \end{proof}
@@
 \label{swapping_lemma}
-If $x, y, z$ is a three-element subsequence of a permutation $\pi$ in $T_n$ such that $x > y > z$ or $x > z > y$, then swapping $y$ and $z$ in the permutation $\pi$ yields a permutation $\pi'$ that also belongs to $T_n$.
+If $x, y, z$ is a three-element subsequence of a permutation $\pi$ in $T_n$ such that $x > y > z$ or $x > z > y$, then swapping $y$ and $z$ in the permutation $\pi$ yields a permutation that also belongs to $T_n$.
 \end{lemma}
@@
 \begin{proof}
-From Lemma \ref{tn_avoidance} we see that both $y$ and $z$ must be even. Then there exist two cases that need to be covered.
+Since $x$ is to the left of $y$ and $z$, and $x>y$ and $x>z$, it follows by Lemma \ref{tn_avoidance} that both $y$ and $z$ must be even.
+Let $\pi'$ be the permutation obtained by swapping $y$ and $z$ in the permutation $\pi$.
+We divide our analysis into two cases:
 \begin{itemize}
@@
 
-If $x$ is odd, then it is clear that the permutation $\pi'$ obtained from swapping $y$ and $z$ is still valid, since both $y$ and $z$ are to the right of the odd numbers $y-1$ and $z-1$ (since $y-1$ and $z-1$ are themselves to the left of $x$ since they are strictly less than $x$)
+Then it suffices to check that $y$ is to the right of $y-1$ and that $z$ is still to the right of $z-1$.
+This is easy: since $y-1$ and $z-1$ are odd numbers less than $x$ another odd number, they must be to the left of $x$, and hence to the left of $y$ and $z$.
 
@@
 
-If $x$ is even, then $x-1$ which is an odd number greater than both $y-1$ and $z-1$, must be to the left of $x$.
-Since $y-1$ and $z-1$ are already to the left of $x-1$, we conclude that swapping $y$ and $z$ still ensures that $y$ and $z$ are both to the right of $y-1$ and $z-1$.
+Since $\pi\in T_n$, $x-1$ must be to the left of $x$.
+Because $x>y$ and $x>z$, we must have that $x-1>y-1$ and $x-1>z-1$.
+It follows that $y-1$ and $z-1$ are to the left of $x-1$, and hence $x$, in $\pi$.
+We conclude that $y$ and $z$ are to the right of $y-1$ and $z-1$ in $\pi'$, since $y-1$ and $z-1$ are to the left of $x$ and $y$ and $z$ remain to the right of $x$.
 \end{itemize}
-Hence, we see that regardless of whether $x$ is even or odd, swapping $y$ and $z$ in a permutation $\pi \in T_n$ if $x > y > z$ or $x > z > y$ yields a permutation $\pi'$ that also belongs to $T_n$.
+Hence, regardless of the parity of $x$, swapping $y$ and $z$ yields a permutation $\pi'$ that also belongs to $T_n$, as desired.
 \end{proof}
 
-Given these lemmas, we now present a proof for $t_n(312) = t_n(321)$.
+With these lemmas in hand, we now present a proof that $t_n(312) = t_n(321)$.
 
 \begin{theorem}
-$t_n(312) = t_n(321)$
+$t_n(312)$, the set of permutations in $T_n$ avoiding 312, is precisely the same size as $t_n(321)$, the set of permutations in $T_n$ avoiding 321.
 \end{theorem}
@@
 \begin{proof}
-First observe from Lemma \ref{tn_avoidance} that if $x, y, z$ is a subsequence of three elements of a permutation in $T_n$ in the relative order $312$ or $321$, then both $y$ and $z$ must be even.
+Let $A$ be the set of permutations in $T_n$ avoiding 312, and $B$ the set of permutations in $T_n$ avoiding 321. We will prove the theorem by constructing a bijection between $A$ and $B$.
 
-Before proving that there exists a bijective mapping between the set of permutations in $T_n$ that avoid $312$ (call this set $A$) and the set of permutations in $T_n$ that avoid $321$ (call this set $B$), we establish a chain of transformation moves that can be used to transform an element $a \in A$ to an element $b \in B$.
+Note by Lemma \ref{tn_avoidance} that if $x, y, z$ is a subsequence of three elements of a permutation in $T_n$ in the relative order $312$ or $321$, then both $y$ and $z$ must be even.
+
+Before we do so, however, we establish a process to transform an element $a \in A$ to an element $b \in B$.
 

### 2014-11-07 project2/paper/paper.tex
some edits to section 6 + todo

@@
 \begin{theorem}
+\label{catalan_2}
 The number of SYT of shape $(n, n)$ is given by $C_n$, the $n^{th}$ Catalan number.
@@
 \begin{theorem}
+\label{rsk}
 There exists an algorithm (called the \emph{RSK algorithm}) which defines a bijection between the set of permutations in $S(n)$ avoiding a decreasing subsequence of length 3 and the set of pairs of size-$n$ SYT of the same shape and at most two rows.
@@
 \begin{proof}
-By the first part of Theorem 12, there are $s_n(321)$ elements in the set $A$ of pairs of size-$n$ SYT of the same shape and at most two rows.
-By Theorem 11, the set $B_n$ of SYT of shape $(n, n)$ has its size given by $C_n$.
+By Theorem \ref{rsk}, there are $s_n(321)$ elements in the set $A$ of pairs of size-$n$ SYT of the same shape and at most two rows.
+By Theorem \ref{catalan_2}, the set $B_n$ of SYT of shape $(n, n)$ has its size given by $C_n$.
 We will construct a bijection between the sets $A_n$ and $B_n$: because $|A_n|=s_n(321)$ and $|B_n|=C_n$, it will therefore follow that $s_n(321)=C_n$, as desired.
@@
 \begin{theorem}
-$t_n(312)$, the set of permutations in $T_n$ avoiding 312, is precisely the same size as $t_n(321)$, the set of permutations in $T_n$ avoiding 321.
+There exists a bijection between the set of permutations in $T_n$ avoiding 312 and the set of permutations in $T_n$ avoiding 321. As a consequence, $t_n(312)=t_n(321)$.
 \end{theorem}
@@
 $$\pi \rightarrow \pi_1 \rightarrow \pi_2 \rightarrow \ldots \rightarrow \pi_k$$
-Where the last permutation in the chain $\pi_k$ does not have a single three-element sub-sequence in the relative ordering $321$, that is $\pi' = \pi_k$ is a permutation in $T_n$ that avoids $321$.
+Where the last permutation in the chain $\pi_k$ does not have a single three-element sub-sequence in the relative ordering $321$, that is $\pi' = \pi_k$ is a permutation in $T_n$ that avoids $321$.\todo{why does this terminate?
+also, why is this necessary?}
 
@@
 The above two claims shows there's actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$ -- to obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order); similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order).
+\todo{should have some justification here that reversing reordering these elements doesn't 'step on each other'...
+that is, you can always do it for some $i_1$ and then some $i_2$ afterward}
 Since the mapping from $A$ to $B$ is clearly reversible, and hence invertible, we conclude that the mapping we just established between $A$ and $B$ is bijective.

### 2014-11-07 project2/paper/paper.tex
comments and changes in section 6

@@
 
-Studying avoidance of permutations in $T_n$ seems an interesting exercise. In this section, we consider the sequence of numbers $t_n(\pi)$, defined as 
+We will now study avoidance of permutations in $T_n$. In this section, we consider the sequence of numbers $t_n(\pi)$, defined as 
 \[t_n(\pi)=\#\{\sigma\in T_n|\sigma \text{ avoids } \pi\},\]
@@
 
-The above two claims shows there's actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$ -- to obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order); similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order).
+\todo{this is a really long sentence}The above two claims show that there is actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$ -- to obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order); similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order).
 \todo{should have some justification here that reversing reordering these elements doesn't 'step on each other'...
@@
 \begin{proof}
-As before, proving the above theorem is equivalent to proving that the $m^{th}$ term of $t_n(213)$ is equal to $2^{m-1}$, or that the number of permutations in $T_{2m}$ that avoid $213$ is equal to $2^{m-1}$.
+As before, proving the theorem is equivalent to proving that the $m^{th}$ term of $t_n(213)$ is equal to $2^{m-1}$, or that the number of permutations in $T_{2m}$ that avoid $213$ is equal to $2^{m-1}$.
 
@@
 
-Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $1, 2, 4, 8, \ldots$ are vastly different -- this is an interesting topic of research that we would like to pursue as future work.
+Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $1, 2, 4, 8, \ldots$ are vastly different -- this is an interesting topic of research that we would like to pursue as future work.\todo{is this necessary?
+it seems frantic.
+Maybe if we're going to do this, we should also go back to the point about the $s_n$ proofs being different and say a little more}
 

### 2014-11-07 project2/paper/paper.tex
Addressed TODOs, removed redundant chain explanation

@@
 
-Since $T_n$ is only defined for even $n$, henceforth we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $0$.\todo{0 or 1 here?}
+Since $T_n$ is only defined for even $n$, henceforth we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $1$.
 
@@
 Then it suffices to check that $y$ is to the right of $y-1$ and that $z$ is still to the right of $z-1$.
-This is easy: since $y-1$ and $z-1$ are odd numbers less than $x$ another odd number, they must be to the left of $x$, and hence to the left of $y$ and $z$.
+This is easy: since $y-1$ and $z-1$ are odd numbers less than $x$ (another odd number, by assumption), they must be to the left of $x$, and hence to the left of $y$ and $z$.
 
@@
 
-Before we do so, however, we establish a process to transform an element $a \in A$ to an element $b \in B$.
-
-Consider a permutation $\pi$ in $T_n$ that avoids $312$.
-For each permutation $\pi \in T_n$ that avoids $312$, we want to show that there exists a sequence of transformation moves that transforms $\pi$ to a unique permutation $\pi'$ that avoids $321$.
-
-If the original permutation $\pi$ avoids $321$ already, then clearly $\pi' = \pi$ avoids $321$.
-
-Now if $\pi$ does not avoid $321$, there must exist some sub-sequence $x, y, z$ in $\pi$ in the relative ordering $321$.
-From Lemma \ref{swapping_lemma}, we see that swapping $y$ and $z$ in the permutation $\pi$ yields another valid permutation $\pi_1$ in $T_n$ -- note that the permutation produced after this swap has $x, y, z$ in the relative ordering $312$ instead of $321$.
-Note that $\pi_1$ may still have sub-sequences in the relative ordering $321$, so we may have to perform the above swapping operation again to get some permutation $\pi_2 \in T_n$ such that a sub-sequence $x_1, y_1, z_1$ in $\pi_1$ that has a relative ordering $321$ is in the relative ordering $312$ in $\pi_2$.
-Continuing this process produces the following chain of permutations,
-$$\pi \rightarrow \pi_1 \rightarrow \pi_2 \rightarrow \ldots \rightarrow \pi_k$$
-Where the last permutation in the chain $\pi_k$ does not have a single three-element sub-sequence in the relative ordering $321$, that is $\pi' = \pi_k$ is a permutation in $T_n$ that avoids $321$.\todo{why does this terminate?
-also, why is this necessary?}
-
-This process is completely invertible, since we can take a permutation $\pi'$ in $T_n$ that avoids $321$, and traverse through the above chain in reverse order to obtain a permutation that avoids $312$.
-
 We now prove that there exists a bijection between the sets $A$ and $B$.
@@
 Clearly, by the definition of $A$, $a$ must avoid $312$.
-Now, we claim that for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in decreasing order; if this were not the case, then we would have a subsequence in the relative ordering $312$, which would be a contradiction.
+Now, we claim that for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in decreasing order; if this were not the case, then we would have a subsequence in the relative ordering $312$ (the element $i$ followed  by the two increasing even numbers less than $i$), which would be a contradiction.
 It is easy to see that we can make a corresponding claim about permutation $b$, that is for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in increasing order.
 
-\todo{this is a really long sentence}The above two claims show that there is actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$ -- to obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order); similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order).
+The above two claims show that there is actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$.
+To obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order).
+Similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order).
+Note that these transformations can be expressed as multiple swaps (as described in Lemma \ref{swapping_lemma}).
 \todo{should have some justification here that reversing reordering these elements doesn't 'step on each other'...

### 2014-11-07 project2/paper/paper.tex
more changes

@@
 
-\subsection{Generating Trees}
-
-The generating tree for the sequence $s_n(\pi)$ is the infinite tree whose vertices are elements of $S_n$ that avoid $\pi$.
+%\subsection{Generating Trees}
+%
+%The generating tree for the sequence $s_n(\pi)$ is the infinite tree whose vertices are elements of $S_n$ that avoid $\pi$.
 The first node in the tree is always 1.
@@
 Figure \ref{fig:M1} is an example of the generating tree of $s_n(312)$.
+%
+%\begin{figure}[h!]
+%\begin{tikzpicture}[level distance=0.5cm, growth parent anchor={south}, nodes={anchor=north}]
+%\tikzstyle{hollow node}=[circle,draw,inner sep=0.1mm, text width=1.36cm,align=center]
+%\tikzstyle{level 1}=[sibling distance=6cm] 
+%\tikzstyle{level 2}=[sibling distance=3cm] 
+%\tikzstyle{level 3}=[sibling distance=1.8cm]
+%\tikzstyle{level 4}=[sibling distance=0.9cm]
+%\node [hollow node]{1}
+%	child {
+%		node[hollow node] {12}
+%		child{
+%			node[hollow node]{123}
+%			child{node[hollow node]{1234}}
+%			child{node[hollow node]{1243}}
+%		}
+%		child{
+%			node[hollow node]{132}
+%			child{node{\vdots}}
+%		}
+%	}
+%    	child {
+%		node[hollow node] {21}
+%    		child {
+%			node[hollow node] {213}
+%			child{node{\vdots}}
+%		}
+%      		child {
+%			node[hollow node] {231}
+%			child{node{\vdots}}
+%		}
+%		child {
+%			node[hollow node] {321}
+%			child{node{\vdots}}
+%		}
+%    	};
+%
+%\end{tikzpicture}
+%\caption{Generating tree of the sequence $s_n(312)$} \label{fig:M1}
+%\end{figure}
+\subsection{Reversing}
 
-\begin{figure}[h!]
-\begin{tikzpicture}[level distance=0.5cm, growth parent anchor={south}, nodes={anchor=north}]
-\tikzstyle{hollow node}=[circle,draw,inner sep=0.1mm, text width=1.36cm,align=center]
-\tikzstyle{level 1}=[sibling distance=6cm] 
-\tikzstyle{level 2}=[sibling distance=3cm] 
-\tikzstyle{level 3}=[sibling distance=1.8cm]
-\tikzstyle{level 4}=[sibling distance=0.9cm]
-\node [hollow node]{1}
-	child {
-		node[hollow node] {12}
-		child{
-			node[hollow node]{123}
-			child{node[hollow node]{1234}}
-			child{node[hollow node]{1243}}
-		}
-		child{
-			node[hollow node]{132}
-			child{node{\vdots}}
-		}
-	}
-    	child {
-		node[hollow node] {21}
-    		child {
-			node[hollow node] {213}
-			child{node{\vdots}}
-		}
-      		child {
-			node[hollow node] {231}
-			child{node{\vdots}}
-		}
-		child {
-			node[hollow node] {321}
-			child{node{\vdots}}
-		}
-    	};
+\begin{lemma}[Reversing Lemma]
+The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. I.e. $\mathcal{R}$ is an avoidance preserving operator. 
+\end{lemma}
+\begin{proof}
+Certainly the statement is true for elements of $S_1$.
+Suppose it is true for elements of $S_{n-1}$.
+Suppose that $\sigma\in S_n$ avoids $\pi$.
+Then if $\sigma_p$ is the parent of $\sigma$, $\mathcal{R}(\sigma_p)$ avoids $\mathcal{R}(\pi)$.
+Now if we add $n$ to the permutation $\mathcal{R}(\sigma_p)$ to construct $\mathcal{F}(\sigma)$, we see that $\mathcal{F}(\sigma)$ must avoid $\mathcal{F}(\pi)$, for if it did not, we could reverse one more time and obtain the result that $\sigma\in S_n$ does not avoid $\pi$, a contradiction.
+The other direction follows because $\mathcal{R}^2=1$.
+\end{proof}
 
-\end{tikzpicture}
-\caption{Generating tree of the sequence $s_n(312)$} \label{fig:M1}
-\end{figure}
+\begin{corollary} 
+For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{R}(\pi))$. 
+\end{corollary}
+\begin{proof}
+This result follows because the reversing lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
+\end{proof}
 
@@
 The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation.
-We show that if ${\sigma}$ avoids $\pi$ then $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$.
+We show that if ${\sigma}\in S_n$ does not avoid $\pi$ then $\mathcal{F}(\sigma)$ does not avoid $\mathcal{F}(\pi)$.
+Well, if ${\sigma}$ does not avoid $\pi$ then there is some subsequence of $\sigma$ that obeys the same set of inequalities as $\pi$.
+Now since $\sigma\in S_n$, $\mathcal{F}(\sigma)$ is the same as the sequence obtained by subtracting every entry of $\sigma$ from $n+1$.
+If we take the collection of inequalities that that the
+
 Let $\pi=b_1\cdots b_k$.
 We proceed by induction.
+Certainly the statement is true for elements of $S_1$.
+Suppose it is true for elements of $S_{n-1}$.
+If ${\sigma}$ does avoids $\pi$ then $\sigma$ appears in the generating tree of $\pi$.
+Let $\sigma_P$ be the parent of $\sigma$ in the generating tree of $\pi$.
+Then by the induction hypothesis, $\mathcal{F}(\sigma_P)$ appears in the generating tree of $\mathcal{F}(\pi)$.
+Now suppose that $\mathcal{F}(\sigma)$ is not a child of $\mathcal{F}(\sigma_P)$  in the generating tree of $\mathcal{F}(\pi)$
 
-Certainly the statement is true for elements of $S_1$.
+%Certainly the statement is true for elements of $S_1$.
 Suppose it is true for elements of $S_{n-1}$.
@@
 This result follows because the flipping lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
-\end{proof}
-\subsection{Reversing}
-
-\begin{lemma}[Reversing Lemma]
-The permutation $\sigma$ avoids the permutation $\pi$ if and only if $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$. I.e. $\mathcal{R}$ is an avoidance preserving operator. 
-\end{lemma}
-\begin{proof}
-Again we use the generating tree.
-Certainly the statement is true for elements of $S_1$.
-Suppose it is true for elements of $S_{n-1}$.
-Suppose that $\sigma\in S_n$ avoids $\pi$.
-Then if $\sigma_p$ is the parent of $\sigma$, $\mathcal{R}(\sigma_p)$ avoids $\mathcal{R}(\pi)$.
-Now if we add $n$ to the permutation $\mathcal{R}(\sigma_p)$ to construct $\mathcal{F}(\sigma)$, we see that $\mathcal{F}(\sigma)$ must avoid $\mathcal{F}(\pi)$, for if it did not, we could reverse one more time and obtain the result that $\sigma\in S_n$ does not avoid $\pi$, a contradiction.
-The other direction follows because $\mathcal{R}^2=1$.
-\end{proof}
-
-\begin{corollary} 
-For a permutation $\pi$, $s_n(\pi)=s_n(\mathcal{R}(\pi))$. 
-\end{corollary}
-\begin{proof}
-This result follows because the reversing lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
 \end{proof}
@@
 
-Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $1, 2, 4, 8, \ldots$ are vastly different -- this is an interesting topic of research that we would like to pursue as future work.\todo{is this necessary?
-it seems frantic.
-Maybe if we're going to do this, we should also go back to the point about the $s_n$ proofs being different and say a little more}
+Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $(2^n)$ are vastly different -- this is an interesting topic of research that we would like to pursue as future work.
+
+This observation parallels the fact that two different methods of proof were needed for $(s_n(312))$ and $(s_n(123))$, even though both sequences were identical to the Catalan numbers.
+This begs the question, is there are proof technique that works across the board for these types of permutation avoidance problems?
 

### 2014-11-07 project2/paper/paper.tex
more changes

@@
 
-\title{Avoidance in permutations}
+\title{Pattern Avoidance}
 
@@
 
-\todo{We never define what a pattern is. Either scrub the paper of mentions or define it here}
-
 \begin{definition}
-Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same relative order if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
+Two finite sequences of distinct positive integers $a_1,\cdots a_k$ and $b_1,\cdots b_k$ have the same pattern (or relative order) if the $i^\text{th}$ largest entry of each unordered set $\{a_1,\cdots,a_k\}$ and $\{b_1,\cdots, b_k\}$ appears in the same position in $a_1\cdots a_k$ and $b_1\cdots b_k$ for each $1\le i\le k$.
 \end{definition}
@@
 Two such operators that we will identify are the `flipping' and `reversing' operators.
-Before discussing these operators, we will introduce generating trees, a tool which we use to study pattern avoidance and particularly to prove the avoidance preserving nature of these operators.
 
@@
 \begin{proof}
-Certainly the statement is true for elements of $S_1$.
-Suppose it is true for elements of $S_{n-1}$.
-Suppose that $\sigma\in S_n$ avoids $\pi$.
-Then if $\sigma_p$ is the parent of $\sigma$, $\mathcal{R}(\sigma_p)$ avoids $\mathcal{R}(\pi)$.
-Now if we add $n$ to the permutation $\mathcal{R}(\sigma_p)$ to construct $\mathcal{F}(\sigma)$, we see that $\mathcal{F}(\sigma)$ must avoid $\mathcal{F}(\pi)$, for if it did not, we could reverse one more time and obtain the result that $\sigma\in S_n$ does not avoid $\pi$, a contradiction.
-The other direction follows because $\mathcal{R}^2=1$.
+We need to only prove one direction of this theorem, then the other direction follows because $\mathcal{R}^2=1$.
+We show that if ${\sigma}\in S_n$ avoids $\pi$ then $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$.
+Suppose that $\mathcal{R}(\sigma)$ does not avoid $\mathcal{R}(\pi)$.
+Then If we reverse $\mathcal{R}(\sigma)$ again, the subsequence of $\mathcal{R}(\sigma)$ that follows the same pattern as $\mathcal{R}(\pi)$ will reverse to a subsequence that follows the same pattern as $\pi$, giving us a contradiction.
+Therefore $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$.
+
 \end{proof}
@@
 The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation.
-We show that if ${\sigma}\in S_n$ does not avoid $\pi$ then $\mathcal{F}(\sigma)$ does not avoid $\mathcal{F}(\pi)$.
-Well, if ${\sigma}$ does not avoid $\pi$ then there is some subsequence of $\sigma$ that obeys the same set of inequalities as $\pi$.
-Now since $\sigma\in S_n$, $\mathcal{F}(\sigma)$ is the same as the sequence obtained by subtracting every entry of $\sigma$ from $n+1$.
-If we take the collection of inequalities that that the
-
-Let $\pi=b_1\cdots b_k$.
-We proceed by induction.
-Certainly the statement is true for elements of $S_1$.
-Suppose it is true for elements of $S_{n-1}$.
-If ${\sigma}$ does avoids $\pi$ then $\sigma$ appears in the generating tree of $\pi$.
-Let $\sigma_P$ be the parent of $\sigma$ in the generating tree of $\pi$.
-Then by the induction hypothesis, $\mathcal{F}(\sigma_P)$ appears in the generating tree of $\mathcal{F}(\pi)$.
-Now suppose that $\mathcal{F}(\sigma)$ is not a child of $\mathcal{F}(\sigma_P)$  in the generating tree of $\mathcal{F}(\pi)$
-
-%Certainly the statement is true for elements of $S_1$.
-Suppose it is true for elements of $S_{n-1}$.
-Let $\sigma\in S_n$.
-If ${\sigma}$ does avoids $\pi$ then $\sigma$ appears in the generating tree of $\pi$.
-Then $\sigma=a_1\cdots a_n$ where $a_i\in\{1,\cdots, n\}$.
-Without loss of generality, let $a_\ell=n$.
-Then the parent node of $\sigma$ is $\sigma_p=a_1\cdots a_{\ell-1} a_{\ell+1}\cdots a_n$.
-The induction hypothesis allows us to conclude that $\mathcal{F}(\sigma_p)$ avoids $\mathcal{F}(\pi)$.
-Now suppose that $\mathcal{F}(\sigma)$ is not a child of $\mathcal{F}(\sigma_p)$.
-Let $\mathcal{F}(\sigma)=a'_1\cdots a'_{\gamma-1} a_{\gamma}a'_{\gamma+1}\cdots a'_n$ and $\mathcal{F}(\sigma_p)=a'_1\cdots a'_{\gamma-1} a'_{\gamma+1}\cdots a'_n$ with $a'_\gamma=n$.
-Then $\mathcal{F}(\sigma_p)$ avoids $\pi$ while $\mathcal{F}(\sigma)$ does not.
-That means that there is a subsequence $\omega$ of $\mathcal{F}(\sigma)$ that has the same relative order as $\mathcal{F}(\pi)$.
-Let $\mathcal{F}(\pi)= b'_1\cdots b'_k$ with $b'_i=1$ and $b'_j=k$.
-So $\omega$ contains $a'_\gamma$, and $a'_\gamma$ occupies the same position as $b'_j$.
-Now if we remove 1 and $n$ from $\mathcal{F}(\sigma)$, then the result does not avoid $\mathcal{F}(\pi)$ with $b'_i=1$ and $b'_j=k$ omitted.
-Then if we apply a flip, by the induction hypothesis we see that the result of flipping $\mathcal{F}(\sigma)$ with $1$ and $n$ omitted does not avoid $\pi$ with $1$ and $k$ omitted.
-If we add in 1 and $n$ in the appropriate places to reconstruct $\mathcal{F}(\mathcal{F}(\sigma))=\sigma$ and if we reconstruct $\sigma$ similarly, then $\sigma$ does not avoid $\pi$, a contradiction.
+We show that if ${\sigma}\in S_n$ avoids $\pi$ then $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$.
+Suppose that $\mathcal{F}(\sigma)$ does not avoid $\mathcal{F}(\pi)$.
+Let $\mathcal{F}(\sigma)=a_1\cdots a_n$ and let $\mathcal{F}(\pi)=b_1\cdots  b_k$.
+Then there is some subsequence $a_{i_1}\cdots a_{i_k}$ so that $a_{i_c}<a_{i_d}$ if and only if $b_c<b_d$.
+Now flipping $\mathcal{F}(\sigma)$ gives us the sequence $(n+1-a_1)(n+1-a_2)\cdots(n+1-a_n)$ and flipping $\mathcal{F}(\pi)$ gives the sequence $(k+1-b_1)(k+1-b_2)\cdots(k+1-b_k)$.
+Thus $n+1-a_{i_c}>n+1-a_{i_d}$ if and only if $k+1-b_c>k+1-b_d$.
+So  this implies that $\sigma$ does not avoid $\pi$, a contradiction.
 \end{proof}
@@
 
-It is easy to see that the above proof can be modified slightly to prove that the sequences $s_n(132)$, $s_n(213)$ and $s_n(231)$ are in fact the sequence of catalan numbers as well.
+By the flipping lemma and reversing lemmas we see that  $s_n(132)$, $s_n(213)$ and $s_n(231)$ are the sequence of Catalan numbers as well.
+
+%It is easy to see that the above proof can be modified slightly to prove that the sequences $s_n(132)$, $s_n(213)$ and $s_n(231)$ are in fact the sequence of catalan numbers as well.
 In fact, we claim that if $\sigma$ and $\pi$ are both permutations in $S_k$ for some arbitrary $k$, then if $\sigma = flip(\pi)$, $s_n(\pi) = s_n(\sigma)$.

### 2014-11-07 project2/paper/paper.tex
small edits to Theorem 28

--- 2014-11-07 project2/paper/paper.tex
@@
 
-Note by Lemma \ref{tn_avoidance} that if $x, y, z$ is a subsequence of three elements of a permutation in $T_n$ in the relative order $312$ or $321$, then both $y$ and $z$ must be even.
-
-We now prove that there exists a bijection between the sets $A$ and $B$.
+Let $a$ be an arbitrary permutation in $A$, and $b$ an arbitrary permutation in $B$.
+By the definition of $A$, $a$ avoids 312.
+Pick any positive integer $i\leq n$.
+We note that all even integers to the right of $i$ and less than $i$ in $a$ must be in decreasing order; otherwise, we would obtain a subsequence with pattern 312.
+Similarly, all even integers to the right of $i$ and less than $i$ in $b$ must be in increasing order.
 

### 2014-11-07 project2/paper/paper.tex
added more rigor to the proof to Theorem 28

@@
 \begin{proof}
-Let $A$ be the set of permutations in $T_n$ avoiding 312, and $B$ the set of permutations in $T_n$ avoiding 321. We will prove the theorem by constructing a bijection between $A$ and $B$.
-
+Let $A$ be the set of permutations in $T_n$ avoiding 312, and $B$ the set of permutations in $T_n$ avoiding 321.
 Let $a$ be an arbitrary permutation in $A$, and $b$ an arbitrary permutation in $B$.
@@
 
-Let $a$ be an arbitrary permutation in $A$.
-Clearly, by the definition of $A$, $a$ must avoid $312$.
-Now, we claim that for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in decreasing order; if this were not the case, then we would have a subsequence in the relative ordering $312$ (the element $i$ followed  by the two increasing even numbers less than $i$), which would be a contradiction.
-It is easy to see that we can make a corresponding claim about permutation $b$, that is for every element $i$ in the permutation $a$, all even numbers in $a$ to the right of $i$ and less than $i$ must be in increasing order.
+In fact, note that any permutation $\pi\in T_n$ which satisfies this condition - for all $i\leq n$, all the even integers to the right of $i$ and less than $i$ are in decreasing order - is in fact a permutation in $A$.
+Suppose for the sake of contradiction $\pi$ had a subsequence $x, y, z$ following the pattern 312 (so in particular $x>z>y$).
+Since it is in $T_n$, by Lemma  \ref{tn_avoidance} $y$ and $z$ must be even.
+Then by the assumed property of $\pi$ we must have $y>z$, a contradiction; hence such a subsequence cannot exist, and so $\pi$ avoids 312.
+We have a similar result for $B$: $B$ is precisely the set of permutations $\pi\in T_n$ satisfying the condition that, for all $i\leq n$, all even integers to the right of $i$ and less than $i$ are in decreasing order.
 
-The above two claims show that there is actually an easy way to transform an arbitrary permutation $a \in A$ to a $b \in B$, and similarly, to transform an arbitrary permutation $b \in B$ to a $a \in A$.
-To obtain a permutation in $B$ from a permutation $a \in A$, for every element $i$ in the permutation $a$, all even elements in $a$ to the right of $i$ less than $i$ should be reversed (that is reordered to in increasing order).
-Similarly to obtain a permutation in $A$ from a permutation $b \in B$, for every element $i$ in the permutation $b$, all even elements in $b$ to the right of $i$, less than $i$ should be reversed (that is reordered to in decreasing order).
-Note that these transformations can be expressed as multiple swaps (as described in Lemma \ref{swapping_lemma}).
-\todo{should have some justification here that reversing reordering these elements doesn't 'step on each other'...
-that is, you can always do it for some $i_1$ and then some $i_2$ afterward}
-Since the mapping from $A$ to $B$ is clearly reversible, and hence invertible, we conclude that the mapping we just established between $A$ and $B$ is bijective.
+Then the bijection between $A$ and $B$ is as follows: given a permutation $a\in A$, consider each of its elements $i$ beginning with its rightmost element and moving leftward.
+By the argument above, the collection of even elements to the right of $i$ and less than $i$ will be in decreasing order; for each $i$, we sort this collection to make the elements increasing instead.
+Note that this process can be obtained by transposing elements, so that by Lemma \ref{swapping_lemma} the result at each stage will be a permutation in $T_n$.
+The end result will therefore be a permutation in $T_n$ with the property that for each $i$, all even integers to the right of $i$ and less than $i$ must be in increasing order.
+As we argued above, therefore, the end result is a permutation in $B$.
 
-Since there exists a bijective mapping between permutations in $T_n$ that avoid $312$ and permutations in $T_n$ that avoid $321$, we can conclude that $t_n(312) = t_n(321)$.
-
+Indeed, it is easy to see that this process is invertible: for each element $j$, beginning with the leftmost element and moving right, we reverse sort the elements less than $j$ and to the right of $j$.
+Since the process is invertible, this map in fact defines a bijection, as desired.
 \end{proof}

### 2014-11-07 project2/paper/paper.tex
in fact

@@
 
-In fact, note that any permutation $\pi\in T_n$ which satisfies this condition - for all $i\leq n$, all the even integers to the right of $i$ and less than $i$ are in decreasing order - is in fact a permutation in $A$.
+In fact, note that any permutation $\pi\in T_n$ which satisfies this condition - for all $i\leq n$, all the even integers to the right of $i$ and less than $i$ are in decreasing order - is a permutation in $A$.
 Suppose for the sake of contradiction $\pi$ had a subsequence $x, y, z$ following the pattern 312 (so in particular $x>z>y$).

### 2014-11-07 project2/code/compute_tn.py

@@ -80,7 +80,7 @@ def compute_tn(n, order):
 
 if __name__ == '__main__':
     n = int(sys.argv[1])
-    order = [2, 1, 3]
+    order = [3,2,1]
     for i in xrange(n):
         if (2*i < len(order)):
             continue

### 2014-11-07 project2/paper/paper.tex
changed theorems in sections 3,4

@@
 \end{proof}
-\section{Terms in the sequence $s_n(312)$}
+\section{Avoiding Permutations of Length Three Part I}
 \label{312}
@@
 \begin{theorem}
-$s_n(312)$, the total number of permutations of $\{1,2,3,...,n\}$ that avoid the pattern $312$, is equal to $C_n$, the $n^{th}$ Catalan number.
+$s_n(312), s_n(132), s_n(213)$, and $s_n(231)$ are equal to $C_n$, the $n^{th}$ Catalan number.
 \end{theorem}
@@
 \begin{proof}
+We prove this result for $(s_n(312))$ by induction and then use the flipping and reversing lemmas for the other three sequences.
 The inductive hypothesis holds for our base case of $\{1\}$, since the only permutation of $\{1\}$ trivially avoids $312$.
 
-We need to prove the inductive case. Let us first assume that for all $i$ from $1$ to $k$, the number of permutations of $\{1,2,...,i\}$ that avoid the order $312$ as a subsequence is $C_i$.
+Let us assume that for all $i$ from $1$ to $k$, the number of permutations of $\{1,2,...,i\}$ that avoid the order $312$ as a subsequence is $C_i$.
 
@@
 
+By the flipping lemma and reversing lemma we see that  $s_n(132)$, $s_n(213)$ and $s_n(231)$ are the sequence of Catalan numbers as well. 
 \end{proof}
-
-By the flipping lemma and reversing lemmas we see that  $s_n(132)$, $s_n(213)$ and $s_n(231)$ are the sequence of Catalan numbers as well.
 
@@
 
-\section{Terms in the sequence $s_n(321)$}
+\section{Avoiding Permutations of Length Three Part II}
 
@@
 As a consequence, the proof in the previous section simply does not hold.
-Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping principle).
+Fortunately, however, we still have a closed form expression for $s_n(321)$ (recall that $s_n(321)=s_n(123)$ by our flipping lemma).
 Surprisingly, in fact, the sequence remains the same: $s_n(321)=C_n$, the $n^{th}$ Catalan number.
@@
 \begin{theorem}
-The total number of permutations of $\{1,2,3,\ldots, n\}$ that avoid the order 321 as a subsequence is $C_n$, where $C_n$ is the $n^{th}$ Catalan number.
+$s_n(321)$ and $s_n(123)$ are equal to $C_n$, the $n^{th}$ Catalan number.
 \end{theorem}
 \begin{proof}
+By the reversing lemma we only need to prove this result for $s_n(321)$.
 By Theorem \ref{rsk}, there are $s_n(321)$ elements in the set $A$ of pairs of size-$n$ SYT of the same shape and at most two rows.

### 2014-11-07 project2/paper/paper.tex
changed theorem numbering
capitalization
attributions
missing comma
typo in section 1

@@
 \theoremstyle{theorem}
-\newtheorem{theorem}{Theorem}
+\newtheorem{theorem}{Theorem}[section]
 \newtheorem*{theorem*}{Theorem}
@@
 \theoremstyle{remark}
-\newtheorem{remark}{Remark}
+\newtheorem{remark}[theorem]{Remark}
 \newtheorem*{remark*}{Remark}
@@
 \newtheorem{question}{Question}
-\newtheorem{example}{Example}
+\newtheorem{example}[theorem]{Example}
 
@@
 
+\emph{Yajit wrote this section. }
+
 \section{Flipping and Reversing}
@@
 \begin{proof}
-This result follows because the reversing lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
+This result follows because the reversing lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{R}(\pi)$. 
 \end{proof}
@@
 \end{proof}
+
+\emph{Yajit wrote this section.}
+
 \section{Avoiding Permutations of Length Three Part I}
@@
 \end{proof}
+
+\emph{Deepak wrote this section.}
 
@@
 
+\emph{Leon wrote this section. Yajit edited this section.}
+
 \section{Conjectures on $s_n(\pi)$ for $\pi\in S_4$}
@@
 \{4213,1342,3124,2431\},
-\{2413,3142\}
+\{2413,3142\},
 $$
@@
 
-\section{Avoidance of permutations of $T_n$}
+\
+
+\emph{Yajit wrote this section.}
+
+\section{Avoidance of Permutations of $T_n$}
 \label{Tn}
@@
 
+\
+
+\emph{Deepak wrote this section. Leon and Yajit closely edited this section.}
+
 \end{document}

### 2014-11-07 project2/paper/paper.tex
added more rigor to Theorems 29 and 30

--- 2014-11-07 project2/paper/paper.tex
@@
 
-In fact, note that any permutation $\pi\in T_n$ which satisfies this condition - for all $i\leq n$, all the even integers to the right of $i$ and less than $i$ are in decreasing order - is a permutation in $A$.
+In fact, note that any permutation $\pi\in T_n$ which satisfies this condition - for all $i\leq n$, all the even integers to the right of $i$ and less than $i$ are in decreasing order - is in fact a permutation in $A$.
 Suppose for the sake of contradiction $\pi$ had a subsequence $x, y, z$ following the pattern 312 (so in particular $x>z>y$).
@@
 
-Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above, however we do prove that both the sequences $(t_n(231))$ and $(t_n(213)$ are equal to the sequence $1, 2, 4, 8, \ldots$, and hence equivalent, below.
+Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above.
+In fact, however, we can prove a stronger result: both $t_n(231)$ and $t_n(213)$ are equal to $2^{n-1}$, and hence the sequences we obtain are equivalent.
 
 \begin{theorem}
-$t_n(231) = 1, 2, 4, 8, \ldots$
+$t_n(231)$, the number of permutations in $T_n$ avoiding the pattern 231, is equal to $2^{n-1}$.
 \end{theorem}
@@
 \begin{proof}
-Proving the above theorem is equivalent to proving that the $m^{th}$ term of $t_n(231)$ is equal to $2^{m-1}$, or that the number of permutations in $T_{2m}$ that avoid $231$ is equal to $2^{m-1}$.
+Let $n=2m$.
+Consider a permutation $\sigma$ in $T_{n}$ that avoids $231$, and consider an arbitrary even element $2i$ in $\sigma$.
+For this element $2i$, define $A$ to be the set of numbers greater than $2i$ and to its left in the permutation $\sigma$.
+First off, we claim that the set $A$ cannot contain any even numbers.
+To see why, suppose for the sake of contradiction that there existed such an even number $a$.
+Then $a-1$ would be to the left of $a$, because $\sigma\in T_{2m}$, and $a-1>2i$, so that $a-1, a, 2i$ is a subsequence with pattern 231: a contradiction.
 
-Consider a permutation $\sigma$ in $T_{2m}$ that avoids $231$.
-Consider an arbitrary even element $2i$ in $\sigma$.
-For this element $2i$, define $A$ to be the set of numbers greater than $2i$ and to its left in the permutation $\sigma$.
-First off, note that the set $A$ cannot contain any even numbers, because if it did, then the permutation $\sigma$ would contain a three-element sub-sequence in the relative ordering $231$ -- if $a$ is an even number in the set $A$, then $(a-1, a, 2i)$ would be a three-element sub-sequence of $\sigma$ in the relative ordering $231$ (Note that $a-1$ must be to the left of $a$ in the permutation $\sigma$ because of the way the set $T_{2m}$ was constructed)
+Furthermore, the set $A$ cannot contain any odd elements greater than $2i+1$.
+To see why, again suppose such an element $2j+1$ existed.
+Then $2i+1$ would appear to the left of $2j+1$, since our permutation $\sigma\in T_{2m}$, and so $2i+1, 2j+1, 2i$ would be a subsequence with pattern 231: again a contradiction.
 
-Furthermore, the set $A$ cannot contain any odd elements greater than $2i+1$, because if it did, then again the permutation $\sigma$ would contain a three-element subsequence in the relative ordering $231$ -- if $2j+1 > 2i+1$ is an odd number in $A$, then $(2i+1, 2j+1, 2i)$ is a three-element sub-sequence of $\sigma$ in the relative ordering $231$ Therefore, we conclude that the set $A$ either only contains the element $2i+1$, or is empty.
+It follows from these two results that the only elements to the left of $2i$ must be either less than $2i$ or $2i+1$.
+We claim that each even $2i$ can appear only in the $2i$ or $2i+1$st index, and that $2i+1$ appears in the other index.
+To prove this, we induct on $i$.
+First, we address the base case: $i=1$.
+Since the first entry must be 1, and the only elements which can appear to the left of 2 are 1 and 3, 2 must either be in the second or third entry; furthermore, it is clear that 3 must appear in the other entry.
+The base case therefore follows.
 
-From this we can conclude that the element $2i$ can be inserted into exactly two valid positions in a permutation of length $2m$ -- at index $2i$ and at index $2i+1$.
-It follows from the above result that when element $2i$ is at index $2i$, element $2i+1$ is at index $2i+1$ in the permutation, and when element $2i$ is at index $2i+1$, element $2i+1$ is at index $2i$.
-Note that the element $1$ must always be the first element in the permutation, and the element $2m$ must always be the last element.
+Now suppose the claim holds up to some general $i-1$.
+We want to show that it holds for $i$ as well.
+By the argument at the beginning of the proof, there are only $2i$ elements which can appear to the left of $2i$, so that the largest index in which $2i$ can appear is $2i+1$.
+By the induction hypothesis, however, the earliest index in which $2i$ can appear is $2i$: all the earlier indices have already been occupied.
+As in the base case, it is easy to see that if $2i$ is in the $2i$th position, then $2i+1$ must be in the $2i+1$st position, and if $2i$ is in the $2i+1$st position, then $2i+1$ must be in the $2i$th index.
+The claim follows by induction
 
-Combining all these facts together, we see that the number of permutations $\sigma$ in $T_{2m}$ that avoid $231$ must be equal to $2 \cdot 2 \cdot \ldots \cdot 2 = 2^{m-1}$, since every even element can be inserted in exactly two positions in $\sigma$.
+Combining all these facts together, we see that the number of permutations $\sigma$ in $T_{2m}$ that avoid $231$ must be equal to $ 2^{m-1}$, since every even element can be inserted in exactly two positions in $\sigma$.
 
@@
 \begin{theorem}
-$t_n(213) = 1, 2, 4, 8, \ldots$
+$t_n(213)$, the number of permutations in $T_n$ avoiding the pattern 213, is equal to $2^{n-1}$.
 \end{theorem}
@@
 \begin{proof}
-As before, proving the theorem is equivalent to proving that the $m^{th}$ term of $t_n(213)$ is equal to $2^{m-1}$, or that the number of permutations in $T_{2m}$ that avoid $213$ is equal to $2^{m-1}$.
+Let $n=2m$. We first define $A_m$ as the set of all permutations in $T_{2m}$ that avoid $213$. It is easy to see that $t_m(213)$ is simply the size of $A_m$.
 
-First, let us define $A_m$ as the set of all permutations in $T_{2m}$ that avoid $213$. It is easy to see that the $m^{th}$ term of $t_n(213)$ is $|A_m|$.
+We claim that for every permutation $\sigma$ in $A_m$, the numbers $2m-1$ and $2m$ must be consecutive in the permutation $\sigma$.
+We prove the claim by contradiction: suppose otherwise, i.e.
+that $2m-1$ and $2m$ are not consecutive in our arbitrary choice of $\sigma$.
+Then there must exist some $j$ between them, and $2m-1, j, 2m$ is clearly a subsequence following the pattern 213, a contradiction.
+Hence $2m-1$ and $2m$ must be consecutive.
 
-We now claim that for every permutation $\sigma$ in $A_m$, the numbers $2m-1$ and $2m$ must be consecutive in the permutation $\sigma$.
-We argue that this must be true by contradiction.
-If this were not the case, then it is easy to see that the permutation $\sigma$ would have a sub-sequence in the relative ordering $213$ -- let $j$ be an element that's between $2m-1$ and $2m$ in $\sigma$.
-Clearly, $j < 2m -1 < 2m$ (since every element in $\sigma$ other than $2m$ is less than $2m-1$), which means $(2m - 1, j, 2m)$ are in the relative ordering $213$.
-We conclude that $2m-1$ and $2m$ must be consecutive in $\sigma$.
+We prove the theorem by induction on $m$. That is, we prove that the size of $A_m$ is precisely $2^{m-1}$. First, we easily see that $A_1$ has only one element, so that the base case holds.
 
-Furthermore, we see that we can construct $A_m$ from $A_{m-1}$ by inserting the consecutive elements $2m-1, 2m$ into a permutation $\sigma'$ in $A_{m-1}$.
+Suppose the induction hypothesis holds up to some general $m-1$, i.e.
+that $A_{m-1}$ has size $2^{m-2}$.
+We see that we can construct $A_m$ from $A_{m-1}$ by inserting the consecutive elements $2m-1, 2m$ into a permutation $\sigma'$ in $A_{m-1}$.
 We claim that we get a valid permutation $\sigma$ in $A_m$ from $\sigma'$ only if we insert the consecutive elements $2m-1, 2m$ right after either $2m-3$ or $2m-2$ in $\sigma'$ (which themselves are consecutive in $\sigma'$ from the above claim).
 
-First, note that $2m-1, 2m$ must be inserted to the right of both $2m-3$ and $2m-2$, because of the way $T_{2m}$ is defined.
+First, note that $2m-1, 2m$ must be inserted to the right of  $2m-3$, because of the way $T_{2m}$ is defined.
 Furthermore, if we insert $2m-1, 2m$ right after some $a$ in $\sigma$ not equal to $2m-3$ or $2m-2$, then we see that the produced permutation $\sigma$ has a three-element sub-sequence in the relative ordering $213$ -- $(2m-2, a, 2m-1)$.
 Hence we conclude that $2m-1, 2m$ must be inserted to the right of either $2m-3$ or $2m-2$; since we can do this for every permutation $\sigma'$ in $A_{m-1}$, we conclude that $|A_m| = 2|A_{m-1}|$.
-
-Given that $|A_1|$ is trivially equal to $1$, we conclude that $|A_m|$, or the $m^{th}$ term of $t_n(213)$ is indeed equal to $2^{m-1}$, as desired.
+By the induction hypothesis, we have that $|A_{m-1}|=2^{m-2}$: so $A_m$ has size $2^{m-1}$, as desired, and the result follows by induction.
 \end{proof}
 
-Note that even though we proved that the sequences $t_n(231)$ and $t_n(213)$ are both equivalent, the methods used to prove that each sequence is equal to the sequence $(2^n)$ are vastly different -- this is an interesting topic of research that we would like to pursue as future work.
-
-This observation parallels the fact that two different methods of proof were needed for $(s_n(312))$ and $(s_n(123))$, even though both sequences were identical to the Catalan numbers.
-This begs the question, is there are proof technique that works across the board for these types of permutation avoidance problems?
-
 \end{document}

### 2014-11-07 project2/paper/paper.tex

@@
 \end{proof}
-\section{Terms in the sequence $s_n(312)$}
+
+\emph{Yajit wrote this section.}
+
+\section{Avoiding Permutations of Length Three Part I}
 \label{312}
@@
 
+By the flipping lemma and reversing lemma we see that  $s_n(132)$, $s_n(213)$ and $s_n(231)$ are the sequence of Catalan numbers as well. 
 \end{proof}
 
-By the flipping lemma and reversing lemmas we see that  $s_n(132)$, $s_n(213)$ and $s_n(231)$ are the sequence of Catalan numbers as well.
+\emph{Deepak wrote this section.}
 

### 2014-11-07 project2/paper/paper.tex
more attributions
claimed stuff and edited Section 4 some

@@
 
-\emph{Yajit wrote this section. }
+\emph{Yajit wrote this section. Leon closely edited this section.}
 
@@
 \subsection{Flipping}
-In this section we introduce the concept of flipping and use generating trees to prove the `Flipping Lemma.'
+In this section we introduce the concept of flipping to prove the `Flipping Lemma.'
 \begin{definition}
@@
 
-\emph{Yajit wrote this section.}
+\emph{Yajit wrote this section. Leon edited this section.}
 
@@
 
-\emph{Deepak wrote this section.}
+\emph{Deepak wrote this section. Leon closely edited this section.}
 
@@
 
-\begin{theorem}
+\begin{theorem}\footnote{A more detailed exposition of this result can be found at http://math.mit.edu/~jnovak/d.18.312/lecture16.pdf. The result itself appears on page 9.}
 \label{rsk}
 There exists an algorithm (called the \emph{RSK algorithm}) which defines a bijection between the set of permutations in $S(n)$ avoiding a decreasing subsequence of length 3 and the set of pairs of size-$n$ SYT of the same shape and at most two rows.
-\end{theorem}\todo{give a citation for this}
+\end{theorem}
 Using all this machinery and all these results, we are finally able to prove the closed-form expression for $s_n(321)$.
 \begin{theorem}
-$s_n(321)$ and $s_n(123)$ are equal to $C_n$, the $n^{th}$ Catalan number.
+$s_n(321)$ and $s_n(123)$, the number of permutations on $n$ elements which avoid 321 and 123 respectively, are equal to $C_n$, the $n^{th}$ Catalan number.
 \end{theorem}
@@
 
-\emph{Deepak wrote this section. Leon and Yajit closely edited this section.}
+\emph{Deepak wrote this section. Leon rewrote parts of it. Yajit closely edited this section.}
 

### 2014-11-08 project2/paper/paper.tex
changed attributions
another edit on Section 4
Cleaned up some notation

@@
 
-In Sections \ref{312} and \ref{321} we consider sequences $s_n(\pi)$ for $\pi$ an element of  \\ $\{123,132,213,231,312,321\}$, and prove the following theorem.
+In Sections \ref{312} and \ref{321} we consider sequences $s_n(\pi)$ where $\pi$ is an element of  \\ $\{123,132,213,231,312,321\}$, and prove the following theorem.
 
 \begin{theorem*}
-For $\pi$ an element of $\{123,132,213,231,312,321\}$, the sequence $(s_n(\pi))$ is the Catalan numbers.
+For $\pi$ in $\{123,132,213,231,312,321\}$, the sequence $(s_n(\pi))$ is the Catalan numbers.
 \end{theorem*}
@@
 Of particular interest is the fact that two entirely distinct methods of proof had to be used to prove this theorem for all six permutations.
-In Section \ref{S4} we will consider sequences $s_n(\pi)$ for $\pi$ a permutation of $\{1,2,3,4\}$ and state a conjecture concerning $(s_n(\pi))$.
+In Section \ref{S4} we will consider sequences $s_n(\pi)$ where $\pi$ is a permutation of $\{1,2,3,4\}$ and state a conjecture concerning $(s_n(\pi))$.
 In Section \ref{Tn} we consider a variation of the avoidance problem that considers the problem of avoiding sequences in a subset $T_n$ of the permutations of $n$ elements.
@@
 
-\emph{Yajit wrote this section. Leon closely edited this section.}
+\emph{Yajit wrote this section. Leon closely edited this section. Deepak made minor edits to this section.}
 
@@
 
-%\subsection{Generating Trees}
-%
-%The generating tree for the sequence $s_n(\pi)$ is the infinite tree whose vertices are elements of $S_n$ that avoid $\pi$.
-The first node in the tree is always 1.
-To construct the tree, we define the children of an arbitrary node.
-Let $\omega\in S_k$ be a node of the generating tree of $s_n(\pi)$.
-Then the children of $\omega$ are elements of $S_{k+1}$ that avoid $\pi$ and that can be obtained by inserting the integer $k+1$ between integers in $\omega$.
-Figure \ref{fig:M1} is an example of the generating tree of $s_n(312)$.
-%
-%\begin{figure}[h!]
-%\begin{tikzpicture}[level distance=0.5cm, growth parent anchor={south}, nodes={anchor=north}]
-%\tikzstyle{hollow node}=[circle,draw,inner sep=0.1mm, text width=1.36cm,align=center]
-%\tikzstyle{level 1}=[sibling distance=6cm] 
-%\tikzstyle{level 2}=[sibling distance=3cm] 
-%\tikzstyle{level 3}=[sibling distance=1.8cm]
-%\tikzstyle{level 4}=[sibling distance=0.9cm]
-%\node [hollow node]{1}
-%	child {
-%		node[hollow node] {12}
-%		child{
-%			node[hollow node]{123}
-%			child{node[hollow node]{1234}}
-%			child{node[hollow node]{1243}}
-%		}
-%		child{
-%			node[hollow node]{132}
-%			child{node{\vdots}}
-%		}
-%	}
-%    	child {
-%		node[hollow node] {21}
-%    		child {
-%			node[hollow node] {213}
-%			child{node{\vdots}}
-%		}
-%      		child {
-%			node[hollow node] {231}
-%			child{node{\vdots}}
-%		}
-%		child {
-%			node[hollow node] {321}
-%			child{node{\vdots}}
-%		}
-%    	};
-%
-%\end{tikzpicture}
-%\caption{Generating tree of the sequence $s_n(312)$} \label{fig:M1}
-%\end{figure}
 \subsection{Reversing}
+
+We notate the reversing operator as $\mathcal{R}$. Formally, if $a_1 a_2 \ldots a_n$ is a permutation of $n$ elements, then $\mathcal{R}(a_1 a_2 \ldots a_{n-1} a_n) = a_n a_{n-1} \dots a_2 a_1$.
 
@@
 Suppose that $\mathcal{R}(\sigma)$ does not avoid $\mathcal{R}(\pi)$.
-Then If we reverse $\mathcal{R}(\sigma)$ again, the subsequence of $\mathcal{R}(\sigma)$ that follows the same pattern as $\mathcal{R}(\pi)$ will reverse to a subsequence that follows the same pattern as $\pi$, giving us a contradiction.
+Then if we reverse $\mathcal{R}(\sigma)$ again, the subsequence of $\mathcal{R}(\sigma)$ that follows the same pattern as $\mathcal{R}(\pi)$ will reverse to a subsequence that follows the same pattern as $\pi$, giving us a contradiction.
 Therefore $\mathcal{R}(\sigma)$ avoids $\mathcal{R}(\pi)$.
@@
 \begin{proof}
-This result follows because the reversing lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{R}(\pi)$. 
+This result follows trivially because the reversing lemma describes a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{R}(\pi)$. 
 \end{proof}
@@
 \subsection{Flipping}
-In this section we introduce the concept of flipping to prove the `Flipping Lemma.'
+In this section we introduce the concept of flipping to prove the `Flipping Lemma'.
 \begin{definition}
@@
 Notice that we only need to prove one direction of this statement.
-The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation.
+The opposite direction will follow since applying $\mathcal{F}$ twice returns the original permutation, or $\mathcal{F}^2 = 1$.
 We show that if ${\sigma}\in S_n$ avoids $\pi$ then $\mathcal{F}(\sigma)$ avoids $\mathcal{F}(\pi)$.
@@
 \begin{proof}
-This result follows because the flipping lemma gives a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
+This result follows because the flipping lemma describes a bijection between permutations that avoid $\pi$ and permutations that avoid $\mathcal{F}(\pi)$. 
 \end{proof}
 
-\emph{Yajit wrote this section. Leon edited this section.}
+\emph{Yajit wrote this section. Leon edited this section. Deepak made minor edits.}
 
@@
 
-\emph{Deepak wrote this section. Leon closely edited this section.}
-
-%It is easy to see that the above proof can be modified slightly to prove that the sequences $s_n(132)$, $s_n(213)$ and $s_n(231)$ are in fact the sequence of catalan numbers as well.
-In fact, we claim that if $\sigma$ and $\pi$ are both permutations in $S_k$ for some arbitrary $k$, then if $\sigma = flip(\pi)$, $s_n(\pi) = s_n(\sigma)$.
-This is easy to see since if $a$ is a permutation that avoids $\sigma$, then $flip(a)$ is a permutation that avoids $\pi$ -- since this can be done both ways, it is bijective and we can conclude that the cardinalities of the two sets are equal, or $s_n(\sigma) = s_n(\pi)$.\todo{we should move this up and make it clear from the beginning that this stuff generalizes}
+\emph{Deepak wrote this section. Leon closely edited this section. Yajit made minor revisions.}
 
@@
 In order to prove this result, we need the machinery of Young tableaux.
-What follows is a brief exposition of the key results, adapted from Stanley's \emph{Algebraic Combinatorics}.
+What follows is a brief exposition of the key results, adapted from Chapter 8 of Stanley's \emph{Algebraic Combinatorics}.
 
@@
 
-\emph{Leon wrote this section. Yajit edited this section.}
+\emph{Leon wrote this section. Yajit made minor revisions.}
 
@@
 
-Since $T_n$ is only defined for even $n$, henceforth we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $1$.
+Since $T_n$ is only defined for even $n$, wherever convenient, we shall refer to $T_n$ as $T_{2m}$ where $m$ is any integer greater than or equal to $1$.
 
@@
 Unfortunately, it turns out that proving the equivalence between $231$ and $213$ is not as straightforward as above.
-In fact, however, we can prove a stronger result: both $t_n(231)$ and $t_n(213)$ are equal to $2^{n-1}$, and hence the sequences we obtain are equivalent.
+In fact, however, we can prove a stronger result: both $t_n(231)$ and $t_n(213)$ are equal to $2^{n/2-1}$, and hence the sequences we obtain are equivalent.
 
 \begin{theorem}
-$t_n(231)$, the number of permutations in $T_n$ avoiding the pattern 231, is equal to $2^{n-1}$.
+$t_n(231)$, the number of permutations in $T_n$ avoiding the pattern 231, where $n$ is even, is equal to $2^{n/2-1}$.
 \end{theorem}
@@
 
-Combining all these facts together, we see that the number of permutations $\sigma$ in $T_{2m}$ that avoid $231$ must be equal to $ 2^{m-1}$, since every even element can be inserted in exactly two positions in $\sigma$.
+Combining all these facts together, we see that the number of permutations $\sigma$ in $T_{2m}$ that avoid $231$ must be equal to $ 2^{m-1} = 2^{n/2-1}$, since every even element except $2m$ can be inserted in exactly two positions in $\sigma$.
 
@@
 \begin{theorem}
-$t_n(213)$, the number of permutations in $T_n$ avoiding the pattern 213, is equal to $2^{n-1}$.
+$t_n(213)$, the number of permutations in $T_n$ avoiding the pattern 213, where $n$ is even, is equal to $2^{n/2-1}$.
 \end{theorem}
@@
 Hence we conclude that $2m-1, 2m$ must be inserted to the right of either $2m-3$ or $2m-2$; since we can do this for every permutation $\sigma'$ in $A_{m-1}$, we conclude that $|A_m| = 2|A_{m-1}|$.
-By the induction hypothesis, we have that $|A_{m-1}|=2^{m-2}$: so $A_m$ has size $2^{m-1}$, as desired, and the result follows by induction.
+By the induction hypothesis, we have that $|A_{m-1}|=2^{m-2}$: so $A_m$ has size $2^{m-1} = 2^{n/2 - 1}$, as desired, and the result follows by induction.
 \end{proof}

### 2014-11-16 project2/code/compute_sn.py
new intro slides

@@ -51,13 +51,14 @@ def compute_sn(n, order):
     tot_count = 0
     for permutation in all_permutations:
         if not is_order_in_permutation(permutation, order):
+            print permutation
             tot_count += 1
     return tot_count
 
 
 if __name__ == '__main__':
     n = int(sys.argv[1])
-    order = [1, 2, 3]
+    order = [5,4,1,2,3]
     k = len(order)
     for i in xrange(k, n):
         print compute_sn(i, order)

### 2014-11-19 question sheet/questions.tex
questions

\documentclass{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{fancyhdr}
\usepackage[margin=2.5cm]{geometry}
\setlength{\parindent}{0in}

\begin{document}

\begin{center}
{\bf\Large {Questions for Pattern Avoidance}}
\end{center}

\

\

\emph{Question 1}

\

Give two permutations in $S_3$ that 51234 avoids.

\vspace{4cm}

\emph{Question 2}

\

blah blah blah

\vspace{4cm}

\emph{Question 3}

\

Give two permutations of $S_6$ that are in $T_6$, and one permutation that is not.
\end{document}

### 2014-11-24 project3/code/compute_epsilon_estimate.py
New code files

@@ -1,0 +1,48 @@
+import random
+
+
+def get_random_vectors(n, N):
+    random_vectors = list()
+    for i in xrange(N):
+        random_vector = list()
+        norm = 0.0
+        for i in xrange(n):
+            random_vector.append(random.random() - 0.5)
+            norm += (random_vector[-1] ** 2)
+        norm = norm ** 0.5
+        for i in xrange(n):
+            random_vector[i] = random_vector[i] / norm
+        random_vectors.append(random_vector)
+    return random_vectors
+
+
+def compute_dp(vector1, vector2):
+    dp = 0.0
+    for i in xrange(len(vector1)):
+        dp += (vector1[i] * vector2[i])
+    return dp
+
+
+def compute_epsilon_random(n, N):
+     random_vectors = get_random_vectors(n, N)
+     max_dp = 0.0
+     for i in xrange(len(random_vectors)):
+         for j in xrange(i+1, len(random_vectors)):
+             dp = compute_dp(random_vectors[i], random_vectors[j])
+             if max_dp < abs(dp):
+                 max_dp = abs(dp)
+     return max_dp
+
+
+def get_min_epsilon(n, N, num_iter):
+    min_epsilon = 1.0
+    for i in xrange(num_iter):
+        epsilon = compute_epsilon_random(n, N)
+        if epsilon < min_epsilon:
+            min_epsilon = epsilon
+    print min_epsilon ** 2
+
+
+if __name__ == '__main__':
+    get_min_epsilon(4, 5, 1000000)
+        

### 2014-11-26 project3/paper/paper.tex
Project 3 paper.tex

\documentclass[11pt,letterpaper,twoside,english]{article}

\usepackage[margin=1.4in]{geometry} % controls the size of the margins

% Special symbols, etc.
\usepackage{amssymb,amsbsy,latexsym,ytableau}
\usepackage{amsmath}
\usepackage{graphics, subfigure, float} 
\usepackage{cancel}
%\usepackage{todonotes}
% Encoding settings
\usepackage[latin1]{inputenc}
\usepackage[american]{babel}
\usepackage[T1]{fontenc} 
\usepackage{tikz}

\usepackage{titling} % allows posttitle command

% AMS Math packages

\usepackage{amscd,amsthm}

\usepackage{verbatim, comment} % can comment out text 
\usepackage{mdwlist}

% Graphics
%\usepackage[dvips]{graphicx,epsfig,color}
%\usepackage{subfigure}
%\usepackage{pst-all}
%\usepackage{pstricks-add}
\usepackage{hyperref}  % can only be used with pdflatex - gives hyperlinks
\usepackage{bm} % bold math font
\usepackage{bbm}

\usepackage{todonotes}

\newtheoremstyle{theorem}{1em}{1em}{\slshape}{0pt}{\bfseries}{.}{ }{}
\theoremstyle{theorem}
\newtheorem{theorem}{Theorem}[section]
\newtheorem*{theorem*}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{definition}[theorem]{Definition}
\newtheorem*{claim*}{Claim}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem*{remark*}{Remark}
\newtheorem{algorithm}{Algorithm}
\newtheorem*{question*}{Question}
\newtheorem{question}{Question}
\newtheorem{example}[theorem]{Example}

\providecommand{\setN}{\mathbb{N}}
\providecommand{\setZ}{\mathbb{Z}}
\providecommand{\setQ}{\mathbb{Q}}
\providecommand{\setR}{\mathbb{R}}
\providecommand{\E}{\mathrm{E}}
\providecommand{\Pr}{\mathrm{Pr}}
\providecommand{\Var}{\mathrm{Var}}

\tikzset
{
    treenode/.style = {circle, draw=black, align=center, minimum size=1cm},
}

\makeatother

\title{Almost Orthogonal vectors}

\author{Yajit Jain, Deepak Narayanan, Leon Zhang}

\begin{document}

\maketitle

\section{Introduction}
Given a dimension $n$ and a number $N > n$ of vectors, we want to find vectors $v_1, v_2, \ldots, v_N \in \mathcal{R}^n$ of unit length, such that
$$\epsilon = \max_{i \neq j} \{ |v_i \cdot v_j|^2 \}$$
is as small as possible.

\section{$n=2$, generic $N$}
In this section, we consider the problem of almost orthogonal vectors of dimensionality $2$.

It is obvious that for $N=2$, we can find an $\epsilon$ equal to $0$, since it easy to pick two unit vectors $v_1$ and $v_2$ of unit length that are orthogonal to each other -- a simple example of such vectors is $[1, 0]^T$ and $[0, 1]^T$.

However, we see that the problem gets harder for larger $N$. Let us first consider the specific case of $N=3$; once we build some intuition for the problem, we will try to generalize to generic $N$.
\\

It's clear that when $N=3$, the value of $\epsilon$ must be greater than $0$, since there is no way three dimensionality-$2$ vectors can be orthogonal to each other.

In addition, we see that for $N=3$, the quantity $\epsilon$ for three unit-length vectors $v_1, v_2, v_3 \in \mathcal{R}^2$ is given by $$\max \{ |v_1 \cdot v_2|^2, |v_1 \cdot v_3|^2, |v_2 \cdot v_3|^2 \}$$

Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathcal{R}^2$ to be the reverses of $v_1, v_2, v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because if the vector $\mathcal{R}(v)$ is the reverse of the vector $v$, then $|v \cdot v'| = |\mathcal{R}(v) \cdot v'|$ for some arbitrary vector $v'$.

Without loss of generality, let $x_1 = [1, 0]^T$; then since $x_4$ is the reverse of $x_1$, we see that $x_4 = [-1, 0]^T$. Also, without loss of generality assume that $x_2$ and $x_3$ are above the $x$-axis, and that $x_2$ is to the right of $x_3$.

\begin{figure}
    \centering
    \begin{tikzpicture}[yscale=-1] 
        % x-axis
        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
        % y-axis
        \draw [thick,->] (0, 4.5) -- (0, -4.5);
        % origin label
        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
        % x-axis label
        \node at (4.5, 0.5) {\text{$x$}};
        % y-axis label
        \node at (0, -5) {\text{$y$}};
        % circle
        \draw (0,0) circle (3cm);
        \draw (3,0)[blue,fill=blue] circle (0.1cm);
        \draw (1.5, -2.59)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);

        \draw [thick,-,blue] (3, 0) -- (0, 0);
        \draw [thick,-,blue] (1.5, -2.59) -- (0, 0);
        \draw [thick,-,blue] (-1.5, -2.59) -- (0, 0);

        \node at (4, -0.3) {\text{$x_1 = [1,0]$}};
        \node at (2.8, -3.1) {\text{$x_2 = [\cos \pi/3,\sin \pi/3]$}};
        \node at (-2.8, -3.1) {\text{$x_3 = [\cos 2\pi/3,\sin 2\pi/3]$}};
    \end{tikzpicture}
    \caption{A configuration of three unit vectors that produce the optimum $\epsilon$ for $n=2, N=3$}
\end{figure}

Let $\theta_1$ be the angle between $x_1$ and $x_2$, $\theta_2$ be the angle between $x_2$ and $x_3$ and $\theta_3$ be the angle between $x_3$ and $x_4$. Now, it is easy to see that $\theta_1 + \theta_2 + \theta_3 = \pi$ and that $$\epsilon = \max_{i \in \{1,2,3\}} \cos ^2 \theta_i$$

Before proceeding, we state and prove the following lemma.
\begin{lemma}
Consider $n$ angles $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi]$ s.t. $\theta_1 + \theta_2 + \ldots + \theta_n = \pi$. Then $\epsilon = \max_i \cos^2 \theta_i$ must equal $\cos^2 \theta_j$ where $j = \text{argmin }\theta_i$.
\end{lemma}

\begin{proof}

Our arguments will hinge on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing. Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \theta_n$. Hence we want to prove that $\epsilon$ is equal to $\cos^2 \theta_n$.

We split our proof into two cases,
\begin{itemize}
\item \textbf{$\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$:} This is easy to see from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.

\item One of $\theta_1, \theta_2, \dots, \theta_n$ is greater than $\pi/2$:

Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$. Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$ (since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$).

\end{itemize}
\end{proof}

Figure $1$ shows the optimal configuration of vectors $v_1$, $v_2$ and $v_3$.
\begin{figure}
    \centering
    \begin{tikzpicture}[yscale=-1] 
        % x-axis
        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
        % y-axis
        \draw [thick,->] (0, 4.5) -- (0, -4.5);
        % origin label
        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
        % x-axis label
        \node at (4.5, 0.5) {\text{$x$}};
        % y-axis label
        \node at (0, -5) {\text{$y$}};
        % circle
        \draw (0,0) circle (3cm);
        \draw (3,0)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, 2.59)[blue,fill=blue] circle (0.1cm);
        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);

        \draw [thick,-,red,dashed] (3, 0) -- (-1.5, 2.59);
        \draw [thick,-,red,dashed] (3, 0) -- (-1.5, -2.59);
        \draw [thick,-,red,dashed] (-1.5, -2.59) -- (-1.5, 2.59);

        % \draw [thick,-,blue, dotted] (-3, 0) -- (0, 0);
        % \draw [thick,-,blue, dotted] (1.5, -2.59) -- (0, 0);
        % \draw [thick,-,blue, dotted] (1.5, 2.59) -- (0, 0);
        \draw [thick,-,blue] (3, 0) -- (0, 0);
        \draw [thick,-,blue] (-1.5, 2.59) -- (0, 0);
        \draw [thick,-,blue] (-1.5, -2.59) -- (0, 0);
    \end{tikzpicture}
    \caption{A configuration of three unit vectors that produce the optimum $\epsilon$ for $n=2, N=3$}
\end{figure}

\end{document}

### 2014-11-26 project3/paper/paper.tex
proof of (n,n+1)

@@
 \newtheorem{example}[theorem]{Example}
+
+\providecommand{\R}{\mathbb{R}}
 
@@
 
+\section{The Parallelogram Law}
+
+First we give the definition of the norm on vectors in $\R^n$ in terms of the dot product:
+
+\begin{definition}
+For a vector $v\in\R^n$ the norm of $v$ is the positive value of $||v||$ given by the equation
+$$
+||v||^2=v\cdot v.
+$$
+\end{definition}
+\begin{theorem}[Parallelogram Law]
+For vectors $v_1,\cdots,v_n\in\R^n$, the following identity holds
+$$
+||v_1+\cdots+v_n||^2=\displaystyle\sum_{i=1}^n||v_i||^2+2\displaystyle\sum_{1\le i<j\le n}v_i\cdot v_j.
+$$
+\proof
+First consider the following identity from the definition of the norm for vectors $u,v\in\R^n$:
+$$
+||u+v||^2=(u+v)\cdot (u+v)=u\cdot u+u\cdot v+ v\cdot u + v\cdot v=||u||^2+||v||^2+2u\cdot v.
+$$
+Now if we let $u=v_1$ and $v= v_2+\cdots+v_n$ we get 
+$$
+||v_1+v+2+\cdots+v_n||^2=||v_1||^2+||v_2+\cdots +v_n||^2+2\sum_{i=2}^n v_1\cdot v_i.
+$$
+If we recurse on $||v_2+\cdots +v_n||^2$ using induction we get the desired result. 
+\end{theorem}
+
+\section{$(n,n+1)$}
+Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space.
+First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$ in this case.
+
+\begin{definition}
+An $n$-dimensional simplex is given by the convex hull of the $n+1$ points in $\R^{n+1}$ described as having a 1 in a single coordinate and zeros in every other coordinate. 
+\end{definition}
+
+Here are diagrams of a 2-dimensional simplex:
+
+\begin{figure}[h!]
+    \centering
+    \begin{tikzpicture}[yscale=-1] 
+        % x-axis
+        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
+        % y-axis
+        \draw [thick,->] (0, 4.5) -- (0, -4.5);
+        % origin label
+        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
+        % x-axis label
+        \node at (4.5, 0.5) {\text{$x$}};
+        % y-axis label
+        \node at (0, -5) {\text{$y$}};
+        % circle
+        \draw (0,0) circle (3cm);
+        \draw (3,0)[blue,fill=blue] circle (0.1cm);
+        \draw (-1.5, 2.59)[blue,fill=blue] circle (0.1cm);
+        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);
+
+        \draw [thick,-,red,dashed] (3, 0) -- (-1.5, 2.59);
+        \draw [thick,-,red,dashed] (3, 0) -- (-1.5, -2.59);
+        \draw [thick,-,red,dashed] (-1.5, -2.59) -- (-1.5, 2.59);
+
+        % \draw [thick,-,blue, dotted] (-3, 0) -- (0, 0);
+        % \draw [thick,-,blue, dotted] (1.5, -2.59) -- (0, 0);
+        % \draw [thick,-,blue, dotted] (1.5, 2.59) -- (0, 0);
+        \draw [thick,-,blue] (3, 0) -- (0, 0);
+        \draw [thick,-,blue] (-1.5, 2.59) -- (0, 0);
+        \draw [thick,-,blue] (-1.5, -2.59) -- (0, 0);
+    \end{tikzpicture}
+    \caption{A 2-dimensional simplex drawn in $\R^3$ on the left, and then centered at the origin in $\R^2$ on the right.}
+\end{figure}
+\todo{make left diagram}
+
+Before proceeding with the theorem, we notice that when $n=2$, the configuration of $N=3$ vectors that produced the minimum value of $\epsilon$ made up a simplex centered at the origin.
+
+\begin{theorem}
+If $N=n+1$ and the vectors in question are denoted $v_1,\cdots, v_N$, then over all configurations of these vectors the maximum value of $v_i\cdot v_j$ for $i\neq j$ is $\frac{-1}{n}$.
+This is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin, and in this configuration $\epsilon=\frac{1}{n^2}$.
+\end{theorem}
+
+\proof
+Let $\delta=\max_{i\neq j}\{v_i\cdot v_j\}$.
+We first show that $\delta\le-\frac{1}{n}$ by showing that the origin-centered simplex configuration of the vectors $v_1,\cdots v_N$ produces a value $\delta=-\frac{1}{n}$.
+
+We must first describe the vectors that make up the origin centered simplex.
+To center our simplex at the origin we take each vector in the original simplex, and subtract from it the centroid of the original simplex.
+The centroid of the original simplex takes the form $(\frac{1}{n},\frac{1}{n},\cdots,\frac{1}{n})$, so the points of the new simplex take the form
+$$
+\left(-\frac{1}{n},\cdots,-\frac{1}{n},\frac{n-1}{n},-\frac{1}{n},\cdots,-\frac{1}{n}\right).
+$$
+If we take the dot product of any two such vectors we get
+$$
+-2\cdot\frac{n-1}{n^2}+(n-2)\cdot\frac{1}{n^2}=-\frac{1}{n}.
+$$
+
+To see that $\delta\ge-\frac{1}{n}$ we use the parallelogram law. Recall that the parallelogram law states that
+
+$$
+||v_1+\cdots+v_{n+1}||^2=\displaystyle\sum_{i=1}^{n+1}||v_i||^2+2\displaystyle\sum_{1\le i<j\le n+1}v_i\cdot v_j.
+$$
+In our case we know three things. First, since the vectors are arranged in a simplex, $v_1+\cdots+v_{n+1}=0$. Second, $v_i\cdot v_j\le\delta$ for any $i<j$. Third, $||v_i||=1$. Therefore 
+$$
+0\le n+1+n(n+1)\delta.
+$$
+This implies that $\delta\ge -\frac{1}{n}$ as desired. Now we see that $\delta=-\frac{1}{n}$, so $|\delta|=\frac{1}{n}$, and therefore $\epsilon=\frac{1}{n^2}$.
+
 \end{document}

### 2014-11-26 project3/code/compute_epsilon_estimate.py

@@ -37,6 +37,7 @@ def compute_epsilon_random(n, N):
 def get_min_epsilon(n, N, num_iter):
     min_epsilon = 1.0
     for i in xrange(num_iter):
+        #if i%10000==0: print i
         epsilon = compute_epsilon_random(n, N)
         if epsilon < min_epsilon:
             min_epsilon = epsilon
@@ -44,5 +45,7 @@ def get_min_epsilon(n, N, num_iter):
 
 
 if __name__ == '__main__':
-    get_min_epsilon(4, 5, 1000000)
+    for n in xrange(4,5): 
+        print n
+        get_min_epsilon(n, n+1, 100000000)
         

### 2014-11-26 project3/paper/paper.tex
More changes to paper

--- 2014-11-26 project3/paper/paper.tex
@@
 \section{Introduction}
-Given a dimension $n$ and a number $N > n$ of vectors, we want to find vectors $v_1, v_2, \ldots, v_N \in \mathcal{R}^n$ of unit length, such that
+Given a dimension $n$ and a number $N > n$ of vectors, we want to find vectors $v_1, v_2, \ldots, v_N \in \mathbb{R}^n$ of unit length, such that
 $$\epsilon = \max_{i \neq j} \{ |v_i \cdot v_j|^2 \}$$
@@
 
-In addition, we see that for $N=3$, the quantity $\epsilon$ for three unit-length vectors $v_1, v_2, v_3 \in \mathcal{R}^2$ is given by $$\max \{ |v_1 \cdot v_2|^2, |v_1 \cdot v_3|^2, |v_2 \cdot v_3|^2 \}$$
+In addition, we see that for $N=3$, the quantity $\epsilon$ for three unit-length vectors $v_1, v_2, v_3 \in \mathbb{R}^2$ is given by $$\max \{ |v_1 \cdot v_2|^2, |v_1 \cdot v_3|^2, |v_2 \cdot v_3|^2 \}$$
 
-Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathcal{R}^2$ to be the reverses of $v_1, v_2, v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
+Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathbb{R}^2$ to be the reverses of $v_1, v_2, v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
 where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because if the vector $\mathcal{R}(v)$ is the reverse of the vector $v$, then $|v \cdot v'| = |\mathcal{R}(v) \cdot v'|$ for some arbitrary vector $v'$.
@@
 \begin{itemize}
-\item \textbf{$\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$:} This is easy to see from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.
+\item $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$: This is easy to see from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.
 
 \item One of $\theta_1, \theta_2, \dots, \theta_n$ is greater than $\pi/2$:
+\begin{figure}
+    \centering
+    \begin{tikzpicture}[yscale=-1] 
+        % x-axis
+        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
+        % y-axis
+        \draw [thick,->] (0, 4.5) -- (0, -4.5);
+        % origin label
+        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
+        % x-axis label
+        \node at (4.5, 0.5) {\text{$x$}};
+        % y-axis label
+        \node at (0, -5) {\text{$y$}};
+        % circle
+        \draw (0,0) circle (3cm);
+        \draw (3,0)[blue,fill=blue] circle (0.1cm);
+        \draw (2.42, -1.76)[blue,fill=blue] circle (0.1cm);
+        \draw (0.93, -2.85)[blue,fill=blue] circle (0.1cm);
+        \draw (-2.42, -1.76)[blue,fill=blue] circle (0.1cm);
+        \draw (-0.93, -2.85)[blue,fill=blue] circle (0.1cm);
+
+        \draw [thick,-,blue] (3, 0) -- (0, 0);
+        \draw [thick,-,blue] (2.42, -1.76) -- (0, 0);
+        \draw [thick,-,blue] (0.93, -2.85) -- (0, 0);
+        \draw [thick,-,blue] (-2.42, -1.76) -- (0, 0);
+        \draw [thick,-,blue] (-0.93, -2.85) -- (0, 0);
+
+        \node at (4, -0.3) {\text{$x_1 = [1,0]$}};
+        \node at (4.5, -1.76) {\text{$x_2 = [\cos \pi/5,\sin \pi/5]$}};
+        \node at (3.43, -3) {\text{$x_3 = [\cos 2\pi/5,\sin 2\pi/5]$}};
+         \node at (-4.5, -1.76) {\text{$x_4 = [\cos 4\pi/5,\sin 4\pi/5]$}};
+        \node at (-3.43, -3) {\text{$x_5 = [\cos 3\pi/5,\sin 3\pi/5]$}};
+    \end{tikzpicture}
+    \caption{A configuration of five unit vectors that produce the optimum $\epsilon$ for $n=2, N=5$}
+\end{figure}
 
 Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$.
-Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$ (since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$).
+Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$.
+(since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$)
+
+It is easy to see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\epsilon$ is equal to $\cos^2 \theta_n$.
 
@@
 
-Figure $1$ shows the optimal configuration of vectors $v_1$, $v_2$ and $v_3$.
+From the above lemma, we can conclude that if $j = \text{argmin } \theta_i$ and $\theta_1 + \theta_2 + \theta_3 = \pi$, then $\epsilon = \cos^2 \theta_j$.
+
+Furthermore, we can obtain an upper bound on $\theta_j$ by observing that $\theta_ 1 + \theta_ 2 + \theta_3 \geq \theta_ j + \theta_j + \theta_j = 3 \theta_j \Rightarrow \theta_j \leq \pi/3$.
+Since we're interested in the smallest such $\epsilon$ and since the cosine function is a decreasing function in $\theta$ between $0$ and $\pi/2$, we conclude that the optimum value of $\epsilon$ for $n=2$ and $N=3$ is $\cos^2 \pi / 3$.
+Figure $1$ shows this optimum configuration.
+
+\begin{theorem}
+For $n=2$ and arbitrary $N$, the optimum value of $\epsilon$ is given by $\cos^2 \pi/N$.
+\end{theorem}
+
+\begin{proof}
+The proof is easy to see given the above lemma.
+
+Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$.
+
+Let us define $\theta_i$ as the angle between the vectors $x_i$ and $x_{i+1}$.
+Then, we quickly see that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$.
+If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$.
+
+Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$.
+\end{proof}
+
+Figure $2$ shows an optimum configuration for $n=2$ and $N=5$.
 \begin{figure}

### 2014-11-26 project3/code/compute_epsilon_estimate.py

@@ -44,5 +44,6 @@ def get_min_epsilon(n, N, num_iter):
 
 
 if __name__ == '__main__':
-    get_min_epsilon(4, 5, 1000000)
+    for i in xrange(10):
+        get_min_epsilon(4, 5, 1000000)
         

### 2014-11-26 project3/paper/paper.tex
Removed unnecessary figure
More changes to paper

@@
 \usepackage{bbm}
+\usepackage{pgfplots}
 
@@
 
-\begin{figure}
+\begin{figure}[!h]
     \centering
@@
 
-Our arguments will hinge on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing.
-Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \theta_n$.
+\begin{figure}[!h]
+	\centering
+	\begin{tikzpicture}
+		\begin{axis}[%
+			axis x line=center, axis y line=center,
+			width=10cm,
+			height=4cm,
+			scale only axis,
+			xmin=-5,
+			xmax=5,
+			xtick={1.57,  3.14, 4.71},
+			xticklabels={, $\pi$, },
+			extra x ticks={-4.71, -3.14, -1.57},
+			extra x tick labels={, $-\pi$,},
+			extra x tick style={
+			    xticklabel style={yshift=0.5ex, anchor=south}
+			},
+			ymin=-1.4,
+			ymax=1.4,
+			ytick={-1,  0,  1}]]
+			\addplot[domain=-2*pi:2*pi,smooth] (\x,{cos(\x r)});
+		\end{axis}
+	\end{tikzpicture}
+	\caption{Plot of $\cos \theta$ versus $\theta$}
+\end{figure}
+
+Our arguments will hinge on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing -- this is easy to see from Figure $2$.
+Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \geq\theta_n$.
 Hence we want to prove that $\epsilon$ is equal to $\cos^2 \theta_n$.
@@
 \begin{itemize}
-\item $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$: This is easy to see from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.
+\item $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$: In this case, it is easy to see that $\epsilon = \cos^2 \theta_n$ from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.
 
 \item One of $\theta_1, \theta_2, \dots, \theta_n$ is greater than $\pi/2$:
-\begin{figure}
+
+Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$.
+Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$.
+(since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$)
+
+In addition, we see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\epsilon$ is equal to $\cos^2 \theta_n$.
+
+\end{itemize}
+\end{proof}
+
+From the above lemma, we can conclude that if $j = \text{argmin } \theta_i$ and $\theta_1 + \theta_2 + \theta_3 = \pi$, then $\epsilon = \cos^2 \theta_j$.
+
+Furthermore, we can obtain an upper bound on $\theta_j$ by observing that $\theta_ 1 + \theta_ 2 + \theta_3 \geq \theta_ j + \theta_j + \theta_j = 3 \theta_j \Rightarrow \theta_j \leq \pi/3$.
+Since we're interested in the smallest such $\epsilon$ and since the cosine function is a decreasing function in $\theta$ between $0$ and $\pi/2$, we conclude that the optimum value of $\epsilon$ for $n=2$ and $N=3$ is $\cos^2 \pi / 3$.
+Figure $1$ shows this optimum configuration.
+
+\begin{theorem}
+For $n=2$ and arbitrary $N$, the optimum value of $\epsilon$ is given by $\cos^2 \pi/N$.
+\end{theorem}
+
+\begin{proof}
+The proof of this theorem is similar to the proof for the specific case of $N=3$.
+
+Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$ -- if any vector $x_i$ were not above the $x$-axis, then we could just consider its reverse.
+
+Let us define $\theta_i$ as the angle between the vectors $x_i$ and $x_{i+1}$.
+Then, we quickly see that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$.
+If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$.
+
+Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$.
+\end{proof}
+
+\begin{figure}[!h]
     \centering
@@
         \node at (3.43, -3) {\text{$x_3 = [\cos 2\pi/5,\sin 2\pi/5]$}};
-         \node at (-4.5, -1.76) {\text{$x_4 = [\cos 4\pi/5,\sin 4\pi/5]$}};
-        \node at (-3.43, -3) {\text{$x_5 = [\cos 3\pi/5,\sin 3\pi/5]$}};
+         \node at (-4.5, -1.76) {\text{$x_5 = [\cos 4\pi/5,\sin 4\pi/5]$}};
+        \node at (-3.43, -3) {\text{$x_4 = [\cos 3\pi/5,\sin 3\pi/5]$}};
     \end{tikzpicture}
@@
 
-Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$.
-Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$.
-(since $\theta_n < \theta_2 + \theta_3 + \ldots + \theta_n$)
-
-It is easy to see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\epsilon$ is equal to $\cos^2 \theta_n$.
-
-\end{itemize}
-\end{proof}
-
-From the above lemma, we can conclude that if $j = \text{argmin } \theta_i$ and $\theta_1 + \theta_2 + \theta_3 = \pi$, then $\epsilon = \cos^2 \theta_j$.
-
-Furthermore, we can obtain an upper bound on $\theta_j$ by observing that $\theta_ 1 + \theta_ 2 + \theta_3 \geq \theta_ j + \theta_j + \theta_j = 3 \theta_j \Rightarrow \theta_j \leq \pi/3$.
-Since we're interested in the smallest such $\epsilon$ and since the cosine function is a decreasing function in $\theta$ between $0$ and $\pi/2$, we conclude that the optimum value of $\epsilon$ for $n=2$ and $N=3$ is $\cos^2 \pi / 3$.
-Figure $1$ shows this optimum configuration.
-
-\begin{theorem}
-For $n=2$ and arbitrary $N$, the optimum value of $\epsilon$ is given by $\cos^2 \pi/N$.
-\end{theorem}
-
-\begin{proof}
-The proof is easy to see given the above lemma.
-
-Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$.
-
-Let us define $\theta_i$ as the angle between the vectors $x_i$ and $x_{i+1}$.
-Then, we quickly see that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$.
-If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$.
-
-Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$.
-\end{proof}
-
-Figure $2$ shows an optimum configuration for $n=2$ and $N=5$.
-\begin{figure}
+\begin{figure}[!h]
     \centering
@@
         \draw (3,0)[blue,fill=blue] circle (0.1cm);
-        \draw (-1.5, 2.59)[blue,fill=blue] circle (0.1cm);
-        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);
+        \draw (-2.42, 1.76)[blue,fill=blue] circle (0.1cm);
+        \draw (0.93, -2.85)[blue,fill=blue] circle (0.1cm);
+        \draw (-2.42, -1.76)[blue,fill=blue] circle (0.1cm);
+        \draw (0.93, 2.85)[blue,fill=blue] circle (0.1cm);
 
-        \draw [thick,-,red,dashed] (3, 0) -- (-1.5, 2.59);
-        \draw [thick,-,red,dashed] (3, 0) -- (-1.5, -2.59);
-        \draw [thick,-,red,dashed] (-1.5, -2.59) -- (-1.5, 2.59);
+        \draw [thick,-,blue] (3, 0) -- (0, 0);
+        \draw [thick,-,blue] (-2.42, 1.76) -- (0, 0);
+        \draw [thick,-,blue] (0.93, -2.85) -- (0, 0);
+        \draw [thick,-,blue] (-2.42, -1.76) -- (0, 0);
+        \draw [thick,-,blue] (0.93, 2.85) -- (0, 0);
 
-        % \draw [thick,-,blue, dotted] (-3, 0) -- (0, 0);
-        % \draw [thick,-,blue, dotted] (1.5, -2.59) -- (0, 0);
-        % \draw [thick,-,blue, dotted] (1.5, 2.59) -- (0, 0);
-        \draw [thick,-,blue] (3, 0) -- (0, 0);
-        \draw [thick,-,blue] (-1.5, 2.59) -- (0, 0);
-        \draw [thick,-,blue] (-1.5, -2.59) -- (0, 0);
+        \node at (4, -0.3) {\text{$x_1 = [1,0]$}};
+        \node at (-4.5, 1.76) {\text{$x'_2 = [\cos \pi/5,\sin \pi/5]$}};
+        \node at (3.43, -3) {\text{$x_3 = [\cos 2\pi/5,\sin 2\pi/5]$}};
+         \node at (-4.5, -1.76) {\text{$x_5 = [\cos 4\pi/5,\sin 4\pi/5]$}};
+        \node at (3.43, 3) {\text{$x'_4 = [\cos 3\pi/5,\sin 3\pi/5]$}};
+
+        \draw [thick,-,red,dashed] (3, 0) -- (0.93, 2.85);
+        \draw [thick,-,red,dashed] (3, 0) -- (0.93, -2.85);
+        \draw [thick,-,red,dashed] (0.93, 2.85) -- (-2.42, 1.76);
+        \draw [thick,-,red,dashed] (0.93, -2.85) -- (-2.42, -1.76);
+        \draw [thick,-,red,dashed] (-2.42, 1.76) -- (-2.42, -1.76);
     \end{tikzpicture}
-    \caption{A configuration of three unit vectors that produce the optimum $\epsilon$ for $n=2, N=3$}
+    \caption{A configuration of five unit vectors that produce the optimum $\epsilon$ for $n=2, N=5$ with $x_2$ and $x_4$ from the previous figure reversed}
 \end{figure}
+
+Figure $3$ shows an optimum configuration for $n=2$ and $N=5$.
+Note by taking reflections of $x_2$ and $x_4$ instead, we get five vectors that form a regular $5$-gon -- this can be seen in Figure $4$.
+This can be done for all odd $N$; unfortunately, even $N$ do not produce such nice polygons.
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[!h]
     \centering

### 2014-11-26 project3/paper/paper.tex
fixed proof

@@
 \end{figure}
-\todo{make left diagram}
+\missingfigure{Make a sketch of a 2d simplex in $\R^3$}
 
@@
 To center our simplex at the origin we take each vector in the original simplex, and subtract from it the centroid of the original simplex.
-The centroid of the original simplex takes the form $(\frac{1}{n},\frac{1}{n},\cdots,\frac{1}{n})$, so the points of the new simplex take the form
+The centroid of the original simplex takes the form $(\frac{1}{n},\frac{1}{n},\cdots,\frac{1}{N})$, so the points of the new simplex take the form
 $$
-\left(-\frac{1}{n},\cdots,-\frac{1}{n},\frac{n-1}{n},-\frac{1}{n},\cdots,-\frac{1}{n}\right).
+\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
+$$
+If we renormalize these vectors we get 
+$$
+\sqrt{\frac{N}{N-1}}\left(-\frac{1}{N},\cdots,-\frac{1}{N},\frac{N-1}{N},-\frac{1}{N},\cdots,-\frac{1}{N}\right).
 $$
@@
 $$
--2\cdot\frac{n-1}{n^2}+(n-2)\cdot\frac{1}{n^2}=-\frac{1}{n}.
+\frac{N}{N-1}\left(-2\cdot\frac{N-1}{N^2}+(N-2)\cdot\frac{1}{N^2}\right)=\frac{N}{N-1}\cdot-\frac{1}{N}=-\frac{1}{N-1}=-\frac{1}{n}.
 $$
+So an upper bound on the value of $\delta$ for all configurations of $n+1$ vectors in $n+1$ dimensional space is $\delta\le\frac{-1}{n}$.
 
@@
 
+\section{N>n+1}
+
+We conjecture that in cases where $N>n+1$ we can never have an arrangement of the vectors where all values of $|v_i\cdot v_j|^2$ are the same.
+
+\section{Applications to sphere packing}
+We should be able to prove something about the radius of spheres you can pack in the case of $N=n+1$, taking as intuition square and hexagonal tilings of the plane and cubic and tetrahedral tilings of 3-space.
+
 \end{document}

### 2014-11-26 project3/code/compute_epsilon_estimate.py
Added new figures

@@ -45,7 +45,5 @@ def get_min_epsilon(n, N, num_iter):
 
 
 if __name__ == '__main__':
-    for n in xrange(4,5): 
-        print n
-        get_min_epsilon(n, n+1, 100000000)
+    get_min_epsilon(3, 6, 1000000)
         

### 2014-11-26 project3/paper/paper.tex
Added new figures
added an intro

@@
 
-\title{Almost Orthogonal vectors}
+\title{Almost orthogonal vectors}
 
@@
 \section{Introduction}
-Given a dimension $n$ and a number $N > n$ of vectors, we want to find vectors $v_1, v_2, \ldots, v_N \in \mathbb{R}^n$ of unit length, such that
-$$\epsilon = \max_{i \neq j} \{ |v_i \cdot v_j|^2 \}$$
-is as small as possible.
+Consider a collection of $N$ unit vectors $v_1, \ldots, v_N$ in $\mathbb R^n$. Define $\epsilon$ as
+\[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
+For various $n$ and $N$, how do we minimize $\epsilon$?
+
+For $n=N$, it is clear that the minimum value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis.
+For fixed $n$, it is also easy to see that the minimum value of $\epsilon$ is nondecreasing as $N$ increases: that is, if $\epsilon_1$ is the minimum value of $\epsilon$ for $N$ unit vectors in $\mathbb R^n$, and $\epsilon_2$ is the minimum value of $\epsilon$ for $N+1$ unit vectors in $\mathbb R^n$, we must have $\epsilon_1\leq \epsilon_2$.
+Otherwise, given a configuration of $N+1$ vectors in $\mathbb R^n$ with $\epsilon=\epsilon_2$, we could simply remove one vector and obtain a configuration of $N$ vectors whose $\epsilon$ would be less than or equal to $\epsilon_1$.
+
+Furthermore, given a configuration of $N$ vectors in $n$-space with corresponding $\epsilon$, we can obtain a collection of $N+1$ vectors in $\mathbb R^{n+1}$ with the same $\epsilon$: simply embed the first $N$ vectors in $\mathbb R^{n+1}$, then add any vector perpendicular to the hyperplane in which the $N$ vectors lie.
+Hence the minimum $\epsilon$ for collections of $N+1$ vectors in $\mathbb R^{n+1}$ is upper-bounded by the corresponding $\epsilon$ for $N$ vectors in $\mathbb R^n$.
+
+In this paper, we shall see that, for $n=2$, the minimum value of $\epsilon$ will be given by 
+\[\epsilon=\cos^2\pi/N.\] 
+We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ will be 
+\[\epsilon=\frac{1}{n^2},\] 
+with the $n+1$ vectors pointing to the vertices of the $n$-simplex.
 

### 2014-11-27 project3/paper/paper.tex
More changes

@@
 \[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
-For various $n$ and $N$, how do we minimize $\epsilon$?
+In this paper, we consider the problem of minimizing $\epsilon$ for various $n$ and $N$.
 
@@
 \[\epsilon=\cos^2\pi/N.\]
+
 We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ will be 
@@
 
-Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathbb{R}^2$ to be the reverses of $v_1, v_2, v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
-where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because if the vector $\mathcal{R}(v)$ is the reverse of the vector $v$, then $|v \cdot v'| = |\mathcal{R}(v) \cdot v'|$ for some arbitrary vector $v'$.
+Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathbb{R}^2$ to be $-v_1, -v_2, -v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
+where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because $|v \cdot v'| = |(-v) \cdot v'|$ for some arbitrary vector $v'$.
 
-Without loss of generality, let $x_1 = [1, 0]^T$; then since $x_4$ is the reverse of $x_1$, we see that $x_4 = [-1, 0]^T$.
-Also, without loss of generality assume that $x_2$ and $x_3$ are above the $x$-axis, and that $x_2$ is to the right of $x_3$.
+Now, without loss of generality, let $x_1 = [1, 0]^T$; then since $x_4$ is $-x_1$, we see that $x_4 = [-1, 0]^T$.
+Also, without loss of generality let us assume that $x_2$ and $x_3$ are above the $x$-axis, and that $x_2$ is to the right of $x_3$.
 
@@
 
+Given the above result for $N=3$, we attempt to generalize to any integer $N$ in the following theorem.
+
 \begin{theorem}
@@
 
-Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$ -- if any vector $x_i$ were not above the $x$-axis, then we could just consider its reverse.
+Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$ -- if any vector $x_i$ were not above the $x$-axis, then we could just consider $-x_i$ instead.
 
@@
 Then, we quickly see that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$.
-If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$.
+If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$ by the above Lemma.
 
-Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$.
+Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$ (again making use of the fact that the cosine function is decreasing).
 \end{proof}
@@
         \node at (4, -0.3) {\text{$x_1 = [1,0]$}};
-        \node at (-4.5, 1.76) {\text{$x'_2 = [\cos \pi/5,\sin \pi/5]$}};
+        \node at (-4.5, 1.76) {\text{$x'_2 = [\cos 6\pi/5,\sin 6\pi/5]$}};
         \node at (3.43, -3) {\text{$x_3 = [\cos 2\pi/5,\sin 2\pi/5]$}};
          \node at (-4.5, -1.76) {\text{$x_5 = [\cos 4\pi/5,\sin 4\pi/5]$}};
-        \node at (3.43, 3) {\text{$x'_4 = [\cos 3\pi/5,\sin 3\pi/5]$}};
+        \node at (3.43, 3) {\text{$x'_4 = [\cos 8\pi/5,\sin 8\pi/5]$}};
 
@@
 Figure $3$ shows an optimum configuration for $n=2$ and $N=5$.
-Note by taking reflections of $x_2$ and $x_4$ instead, we get five vectors that form a regular $5$-gon -- this can be seen in Figure $4$.
-This can be done for all odd $N$; unfortunately, even $N$ do not produce such nice polygons.
+Note by taking reflections of $x_2$ and $x_4$ in Figure $3$, we get five vectors that form a regular $5$-gon -- this can be seen in Figure $4$.
+A similar transformation can be done for all odd $N$; unfortunately, even $N$ cannot produce regular polygons.
 
@@
 $$
-||v_1+v+2+\cdots+v_n||^2=||v_1||^2+||v_2+\cdots +v_n||^2+2\sum_{i=2}^n v_1\cdot v_i.
+||v_1+v_2+\cdots+v_n||^2=||v_1||^2+||v_2+\cdots +v_n||^2+2\sum_{i=2}^n v_1\cdot v_i.
 $$
@@
 
-\section{$(n,n+1)$}
+\section{$N = n+1$}
 Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space.
@@
     \centering
-    \begin{tikzpicture}[yscale=-1] 
+        \begin{tikzpicture}[yscale=-1, scale=0.75]
+        % x-axis
+        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
+        % y-axis
+        \draw [thick,->] (0, 4.5) -- (0, -4.5);
+        % z-axis
+        \draw [thick, ->] (3.0, -2.0) -- (-3.0, 2.0);
+        % origin label
+        \node at (-1, -0.3) {\text{$(0, 0, 0)$}};
+        % x-axis label
+        \node at (4.5, 0.5) {\text{$x$}};
+        % y-axis label
+        \node at (0, -5) {\text{$y$}};
+         % z-axis label
+        \node at (-3.4, 1.9) {\text{$z$}};
+        \node at (3.5, -0.4) {\text{$(1, 0, 0)$}};
+        \node at (1, -3.4) {\text{$(0, 1, 0)$}};
+        \node at (-2.8, 1.1) {\text{$(0, 0, 1)$}};
+        % circle
+        \draw (3,0)[blue,fill=blue] circle (0.1cm);
+        \draw (0, -3)[blue,fill=blue] circle (0.1cm);
+        \draw (-1.8, 1.2)[blue,fill=blue] circle (0.1cm);
+
+        \draw [thick,-,red,dashed] (3, 0) -- (0, -3);
+        \draw [thick,-,red,dashed] (0, -3) -- (-1.8, 1.2);
+        \draw [thick,-,red,dashed] (3, 0) -- (-1.8, 1.2);
+
+        \draw [thick,-,blue] (3, 0) -- (0, 0);
+        \draw [thick,-,blue] (0, -3) -- (0, 0);
+        \draw [thick,-,blue] (-1.8, 1.2) -- (0, 0);
+    \end{tikzpicture}
+    \begin{tikzpicture}[yscale=-1, scale=0.75]
         % x-axis
@@
         % origin label
-        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
+        \node at (-0.6, 0.3) {\text{$(0, 0)$}};
         % x-axis label
@@
 
-        % \draw [thick,-,blue, dotted] (-3, 0) -- (0, 0);
-        % \draw [thick,-,blue, dotted] (1.5, -2.59) -- (0, 0);
-        % \draw [thick,-,blue, dotted] (1.5, 2.59) -- (0, 0);
         \draw [thick,-,blue] (3, 0) -- (0, 0);
@@
     \end{tikzpicture}
+
     \caption{A 2-dimensional simplex drawn in $\R^3$ on the left, and then centered at the origin in $\R^2$ on the right.}
 \end{figure}
-\missingfigure{Make a sketch of a 2d simplex in $\R^3$}
 
@@
 $$
-If we renormalize these vectors we get 
+If we renormalize these vectors so that each vector has unit length, we get 
 $$
@@
 $$
-In our case we know three things. First, since the vectors are arranged in a simplex, $v_1+\cdots+v_{n+1}=0$. Second, $v_i\cdot v_j\le\delta$ for any $i<j$. Third, $||v_i||=1$. Therefore 
+In our case we know three things.
+First, since the vectors are arranged in a simplex with centroid as the origin, $v_1+\cdots+v_{n+1}=0$.
+Second, $v_i\cdot v_j\le\delta$ for any $i<j$.
+Third, $||v_i||=1$.
+Therefore
 $$
@@
 
-\section{N>n+1}
+\section{$N>n+1$}
 
-We conjecture that in cases where $N>n+1$ we can never have an arrangement of the vectors where all values of $|v_i\cdot v_j|^2$ are the same.
+We conjecture that in cases where $N>n+1$ we can never have an arrangement of the vectors where all values of $|v_i\cdot v_j|^2$ are the same. Proving this remains an object of future study.
 

### 2014-12-05 project3/paper/paper.tex
Changes to paper

@@
 
-Furthermore, given a configuration of $N$ vectors in $n$-space with corresponding $\epsilon$, we can obtain a collection of $N+1$ vectors in $\mathbb R^{n+1}$ with the same $\epsilon$: simply embed the first $N$ vectors in $\mathbb R^{n+1}$, then add any vector perpendicular to the hyperplane in which the $N$ vectors lie.
+Furthermore, given a configuration of $N$ vectors in $n$-space with corresponding $\epsilon$, we can obtain a collection of $N+1$ vectors in $\mathbb R^{n+1}$ with the same $\epsilon$: simply embed the first $N$ vectors in $\mathbb R^{n}$, then add any vector perpendicular to the hyperplane in which the $N$ vectors lie.
 Hence the minimum $\epsilon$ for collections of $N+1$ vectors in $\mathbb R^{n+1}$ is upper-bounded by the corresponding $\epsilon$ for $N$ vectors in $\mathbb R^n$.
@@
 
-We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ will be 
-\[\epsilon=\frac{1}{n^2},\] 
-with the $n+1$ vectors pointing to the vertices of the $n$-simplex.
+We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ will be upper bounded by
+\[\frac{1}{n^2},\] 
+and this configuration is produced when the $n+1$ vectors point to the vertices of the $n$-simplex.
 
@@
 Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathbb{R}^2$ to be $-v_1, -v_2, -v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
-where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because $|v \cdot v'| = |(-v) \cdot v'|$ for some arbitrary vector $v'$.
+where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because $$|v \cdot v'| = |(-v) \cdot v'| = |(-v) \cdot (-v')| = |v \cdot (-v')|$$ for some arbitrary vector $v'$.
+Furthermore, we see that there now exists an axis of symmetry -- one of the diagonals passing through either $v_1$ and $v_4 = -v_1$ or $v_2$ and $v_5 = -v_2$ or $v_3$ and $v_6 = -v_3$ -- so we don't need to look at the entire unit circle anymore, and instead can focus on the four vectors contained in a semi-circle resting on one of these three diameters.
 
-Now, without loss of generality, let $x_1 = [1, 0]^T$; then since $x_4$ is $-x_1$, we see that $x_4 = [-1, 0]^T$.
+Without loss of generality, let $x_1 = [1, 0]^T$; then since $x_4$ is $-x_1$, we see that $x_4 = [-1, 0]^T$.
 Also, without loss of generality let us assume that $x_2$ and $x_3$ are above the $x$-axis, and that $x_2$ is to the right of $x_3$.
@@
 Let $\theta_1$ be the angle between $x_1$ and $x_2$, $\theta_2$ be the angle between $x_2$ and $x_3$ and $\theta_3$ be the angle between $x_3$ and $x_4$.
-Now, it is easy to see that $\theta_1 + \theta_2 + \theta_3 = \pi$ and that $$\epsilon = \max_{i \in \{1,2,3\}} \cos ^2 \theta_i$$
+Now, we see that $\theta_1 + \theta_2 + \theta_3 = \pi$ (since vectors $v_1$ and $v_4 = -v_1$ are anti-parallel) and that $$\epsilon = \max_{i \in \{1,2,3\}} \cos ^2 \theta_i$$, since $|x_i \cdot x_j| = \cos \theta$ where $\theta$ is the angle between the two vectors $x_i$ and $x_j$.
 
@@
 Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \geq\theta_n$.
-Hence we want to prove that $\epsilon$ is equal to $\cos^2 \theta_n$.
+Hence our theorem statement is equivalent to proving that $\epsilon$ is equal to $\cos^2 \theta_n$.
 
@@
 Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space.
-First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$ in this case.
+First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$.
 
@@
 
-Here are diagrams of a 2-dimensional simplex:
+Figure $5$ shows diagrams of a 2-dimensional simplex.
 
@@
 \proof
-Let $\delta=\max_{i\neq j}\{v_i\cdot v_j\}$.
-We first show that $\delta\le-\frac{1}{n}$ by showing that the origin-centered simplex configuration of the vectors $v_1,\cdots v_N$ produces a value $\delta=-\frac{1}{n}$.
+Let $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$, and let $\delta_{min}$ be the minimum possible $\delta$ obtained across all unit vectors $v_1, v_2, \ldots, v_{n+1}$.
+We first show that $\delta_{min} \le-\frac{1}{n}$ by showing that the origin-centered simplex configuration of the vectors $v_1,\cdots v_{n+1}$ produces a value $\delta=-\frac{1}{n}$, and then will later show that $\delta_{min} \geq -\frac{1}{n}$ as well.
 
 We must first describe the vectors that make up the origin centered simplex.
-To center our simplex at the origin we take each vector in the original simplex, and subtract from it the centroid of the original simplex.
-The centroid of the original simplex takes the form $(\frac{1}{n},\frac{1}{n},\cdots,\frac{1}{N})$, so the points of the new simplex take the form
+To center our simplex at the origin we take each vector in the original simplex, and subtract from it the vector representing the centroid of the original simplex.
+The centroid of the original simplex takes the form $(\frac{1}{N},\frac{1}{N},\ldots,\frac{1}{N})$, so the points of the new simplex take the form
 $$
@@
 $$
-If we take the dot product of any two such vectors we get
+Now, taking the dot product between any two such vectors gives us
 $$
@@
 $$
-So an upper bound on the value of $\delta$ for all configurations of $n+1$ vectors in $n+1$ dimensional space is $\delta\le\frac{-1}{n}$.
+So an upper bound on the value of $\delta_{min}$ for all configurations of $n+1$ vectors in $n+1$ dimensional space is $\frac{-1}{n}$.
 
-To see that $\delta\ge-\frac{1}{n}$ we use the parallelogram law. Recall that the parallelogram law states that
+To see that $\delta_{min} \ge-\frac{1}{n}$ we use the parallelogram law.
+
+Recall that the parallelogram law states that
 
@@
 In our case we know three things.
-First, since the vectors are arranged in a simplex with centroid as the origin, $v_1+\cdots+v_{n+1}=0$.
+First, since the norm of any vector is greater than or equal to $0$, we can conclude that $||v_1 + \cdots + v_{n+1}||^2 \geq 0$.
 Second, $v_i\cdot v_j\le\delta$ for any $i<j$.
@@
 $$
-This implies that $\delta\ge -\frac{1}{n}$ as desired. Now we see that $\delta=-\frac{1}{n}$, so $|\delta|=\frac{1}{n}$, and therefore $\epsilon=\frac{1}{n^2}$.
+This implies that $\delta\ge -\frac{1}{n}$ for any set of unit vectors $v_1, v_2, \ldots, v_{n+1}$, from which we can conclude that $\delta_{min} \ge -\frac{1}{n}$ as well.
 
+Since $\delta_{min} \le -\frac{1}{n}$ and $\delta_{min} \ge -\frac{1}{n}$, we can conclude that $\delta_{min}=-\frac{1}{n}$.
+From this, we can see that a $\epsilon = \frac{1}{n^2}$ is achievable for the general case of when $N=n+1$ -- this leads us to conclude that the minimum possible $\epsilon$ value for $N=n+1$ is at most $\frac{1}{n^2}$.
 \section{$N>n+1$}

### 2014-12-05 project3/code/compute_epsilon_estimate.py

@@ -24,26 +24,37 @@ def compute_dp(vector1, vector2):
 
 
 def compute_epsilon_random(n, N):
-     random_vectors = get_random_vectors(n, N)
+     # random_vectors = get_random_vectors(n, N)
+     # a = 0.85
+     # b = 0.53
+     a = 0.53
+     b = 0.85
+     random_vectors = [[a, b, 0.0], [a, -b, 0], [b, 0, a], [b, 0.0, -a], [0, a, b], [0, -a, b]]
      max_dp = 0.0
      for i in xrange(len(random_vectors)):
          for j in xrange(i+1, len(random_vectors)):
              dp = compute_dp(random_vectors[i], random_vectors[j])
              if max_dp < abs(dp):
                  max_dp = abs(dp)
-     return max_dp
+     return max_dp, random_vectors
 
 
 def get_min_epsilon(n, N, num_iter):
     min_epsilon = 1.0
+    best_vectors = None
     for i in xrange(num_iter):
-        #if i%10000==0: print i
-        epsilon = compute_epsilon_random(n, N)
+        if i % 10 == 0:
+            print "Done with iteration %d..." % i
+            print "Minimum epsilon seen so far %.2f..." % min_epsilon ** 2
+            print "Best vectors so far:", best_vectors
+            print
+        epsilon, random_vectors = compute_epsilon_random(n, N)
         if epsilon < min_epsilon:
             min_epsilon = epsilon
+            best_vectors = random_vectors
     print min_epsilon ** 2
 
 
 if __name__ == '__main__':
-    get_min_epsilon(3, 6, 1000000)
-        
+    get_min_epsilon(3, 6, 100)
+ 

### 2014-12-07 project3/code/compute_epsilon_estimate.py
Changes to epsilon computation code

@@ -1,4 +1,6 @@
 import random
+from scipy.optimize import minimize
+import numpy as np
 
 
 def get_random_vectors(n, N):
@@ -23,20 +25,67 @@ def compute_dp(vector1, vector2):
     return dp
 
 
+def compute_epsilon(vectors):
+    max_dp = 0.0
+    print vectors
+    for i in xrange(len(vectors)):
+        for j in xrange(i+1, len(vectors)):
+            dp = compute_dp(vectors[i], vectors[j])
+            if max_dp < abs(dp):
+                max_dp = abs(dp)
+    print max_dp
+    return max_dp
+
+
 def compute_epsilon_random(n, N):
-     # random_vectors = get_random_vectors(n, N)
-     # a = 0.85
-     # b = 0.53
-     a = 0.53
-     b = 0.85
-     random_vectors = [[a, b, 0.0], [a, -b, 0], [b, 0, a], [b, 0.0, -a], [0, a, b], [0, -a, b]]
-     max_dp = 0.0
-     for i in xrange(len(random_vectors)):
-         for j in xrange(i+1, len(random_vectors)):
-             dp = compute_dp(random_vectors[i], random_vectors[j])
-             if max_dp < abs(dp):
-                 max_dp = abs(dp)
-     return max_dp, random_vectors
+    # random_vectors = get_random_vectors(n, N)
+    # a = 0.85
+    # b = 0.53
+    a = 0.53
+    b = 0.85
+    random_vectors = [[a, b, 0.0], [a, -b, 0], [b, 0, a], [b, 0.0, -a], [0, a, b], [0, -a, b]]
+    epsilon = compute_epsilon(random_vectors)
+    return epsilon, random_vectors
+
+# Here assume we're given 6 vectors of dimension 3
+def obj(x):
+    x1 = x[0:3]
+    x2 = x[3:6]
+    x3 = x[6:9]
+    x4 = x[9:12]
+    x5 = x[12:15]
+    x6 = x[15:]
+    return compute_epsilon([x1, x2, x3, x4, x5, x6])
+
+
+def jac(x, start, end, n, N):
+    jac = []
+    for i in xrange(n*N):
+        if (i >= start and i < end):
+            jac.append(2 * x[i])
+        else:
+            jac.append(0.0)
+    return np.array(jac)
+
+
+def construct_constraints(n, N):
+    constraints = []
+    start = 0
+    for i in xrange(N):
+        end = start + n
[…]

### 2014-12-07 project3/code/compute_epsilon_estimate.py
Added annealing

@@ -18,6 +18,20 @@ def get_random_vectors(n, N):
     return random_vectors
 
 
+def normalize_vector(vector):
+    norm = 0.0
+    for i in xrange(len(vector)):
+        norm += (vector[i] ** 2)
+    norm = norm ** 0.5
+    for i in xrange(len(vector)):
+        vector[i] /= norm
+
+
+def normalize_vectors(vectors):
+    for vector in vectors:
+        normalize_vector(vector)
+
+
 def compute_dp(vector1, vector2):
     dp = 0.0
     for i in xrange(len(vector1)):
@@ -27,14 +41,12 @@ def compute_dp(vector1, vector2):
 
 def compute_epsilon(vectors):
     max_dp = 0.0
-    print vectors
     for i in xrange(len(vectors)):
         for j in xrange(i+1, len(vectors)):
             dp = compute_dp(vectors[i], vectors[j])
             if max_dp < abs(dp):
                 max_dp = abs(dp)
-    print max_dp
-    return max_dp
+    return max_dp ** 2
 
 
 def compute_epsilon_random(n, N):
@@ -83,8 +95,6 @@ def construct_constraints(n, N):
 def find_optimal_vectors(x0):
     constraints = construct_constraints(3, 6)
     res = minimize(obj, x0, method='SLSQP', constraints=constraints)
-    print res.message
-    print res.success
     return res.x
 
 
@@ -101,12 +111,43 @@ def get_min_epsilon(n, N, num_iter):
         if epsilon < min_epsilon:
             min_epsilon = epsilon
             best_vectors = random_vectors
-    print min_epsilon ** 2
+
+
+def get_min_epsilon_with_annealing(n, N, num_iter, start_vectors):
+    min_epsilon = 1.0
+    current_vectors = start_vectors
+    i = 0
+    while (i < num_iter):
+        temperature = 1.0 / float(i + 1)
+        
+        # Get perturbing vectors
+        random_vectors = get_random_vectors(n, N)
+        # Now perturb current_vectors
+        new_vectors = list()
+        for j in xrange(N):
+            new_vector = list()
+            for k in xrange(n):
+                new_vector.append(current_vectors[j][k] + (temperature * random_vectors[j][k]))
+            new_vectors.append(new_vector)
+        normalize_vectors(new_vectors)
+        epsilon = compute_epsilon(new_vectors)
+        if epsilon < min_epsilon:
[…]

### 2014-12-08 project3/code/compute_epsilon_estimate.py
Setting up experiment to compute (n,N) values

@@ -1,6 +1,4 @@
 import random
-from scipy.optimize import minimize
-import numpy as np
 
 
 def get_random_vectors(n, N):
@@ -59,45 +57,6 @@ def compute_epsilon_random(n, N):
     epsilon = compute_epsilon(random_vectors)
     return epsilon, random_vectors
 
-# Here assume we're given 6 vectors of dimension 3
-def obj(x):
-    x1 = x[0:3]
-    x2 = x[3:6]
-    x3 = x[6:9]
-    x4 = x[9:12]
-    x5 = x[12:15]
-    x6 = x[15:]
-    return compute_epsilon([x1, x2, x3, x4, x5, x6])
-
-
-def jac(x, start, end, n, N):
-    jac = []
-    for i in xrange(n*N):
-        if (i >= start and i < end):
-            jac.append(2 * x[i])
-        else:
-            jac.append(0.0)
-    return np.array(jac)
-
-
-def construct_constraints(n, N):
-    constraints = []
-    start = 0
-    for i in xrange(N):
-        end = start + n
-        constraints.append({'type':'eq',
-            'fun': lambda x: compute_dp(x[start:end], x[start:end]) - 1,
-            'jac': lambda x: jac(x, start, end, n, N)})
-        start += n
-    return constraints
-
-
-def find_optimal_vectors(x0):
-    constraints = construct_constraints(3, 6)
-    res = minimize(obj, x0, method='SLSQP', constraints=constraints)
-    return res.x
-
-
 def get_min_epsilon(n, N, num_iter):
     min_epsilon = 1.0
     best_vectors = None
@@ -117,7 +76,10 @@ def get_min_epsilon_with_annealing(n, N, num_iter, start_vectors):
     min_epsilon = 1.0
     current_vectors = start_vectors
     i = 0
+    l = 0
     while (i < num_iter):
+        if (l > 5000):
+            break
         temperature = 1.0 / float(i + 1)
         
         # Get perturbing vectors
@@ -135,19 +97,21 @@ def get_min_epsilon_with_annealing(n, N, num_iter, start_vectors):
             min_epsilon = epsilon
             current_vectors = new_vectors
             i += 1
+        l += 1
     return min_epsilon, current_vectors
 
 
 if __name__ == '__main__':
-    min_epsilon = 1.0
-    best_vectors = None
-    n, N = 4, 12
-    for i in xrange(100):    
[…]

### 2014-12-08 project3/paper/paper.tex
Added annealing description to paper.

@@
 
-\proof
+\begin{proof}
 Let $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$, and let $\delta_{min}$ be the minimum possible $\delta$ obtained across all unit vectors $v_1, v_2, \ldots, v_{n+1}$.
@@
 From this, we can see that a $\epsilon = \frac{1}{n^2}$ is achievable for the general case of when $N=n+1$ -- this leads us to conclude that the minimum possible $\epsilon$ value for $N=n+1$ is at most $\frac{1}{n^2}$.
-\section{$N>n+1$}
+\end{proof}
 
-We conjecture that in cases where $N>n+1$ we can never have an arrangement of the vectors where all values of $|v_i\cdot v_j|^2$ are the same. Proving this remains an object of future study.
+Note that proving that the minimum possible epsilon for the case of $N = n+1$ is equal to $1/n^2$ is much harder, and we leave this for future work.
 
+\section{Simulated annealing and associated conjectures}
+Since visualizing in spaces of dimensionality greater than $3$ is extremely hard, we used an annealing approach to compute near-optimal epsilon values for different $(n, N)$ pairs.
+We describe this method in greater detail below.
+
+Our general intuition tells us that perturbing a set of nearly optimal vectors slightly could give us a new set of vectors that produce even a smaller epsilon value.
+Given this, we start off with a set of random vectors, and try to move these vectors in order to produce smaller epsilon values.
+
+At every iteration, we maintain the set of vectors (called cur\_vectors) that have produced the optimal epsilon value so far. Then we compute a set new\_vectors defined as follows,
+$$\text{new\_vector} = \text{old\_vector} + (1/i \cdot \text{random\_vector})$$
+
+We update cur\_vectors to the set new\_vectors if the epsilon value produced by new\_vectors is less than the epsilon value produced by cur\_vectors.
+$i$ is incremented whenever cur\_vectors is updated -- having the $1/i$ factor in the above expression ensures that as we move closer to an optimal value, we perturb the set of vectors by a smaller amount.
+
+Even though this approach seems simple, it actually produces remarkably precise values -- this can be verified by cross-checking with results we've already proven / conjectured in this paper.
+
+To make this more concrete, consider Table $1$.
+\begin{table}
+   \centering
+    \begin{tabular}{ | c | c | c | c |}
+    \hline
+      & $n=2$ & $n=3$ & $n=4$ \\ \hline
+    $N=2$ & $0.0$ & $-$ & $-$ \\ \hline
+    $N=3$ & $0.25$ & $0.0$ & $-$ \\ \hline
+    $N=4$ & $0.5$ & $0.11$ & $0.0$ \\ \hline
+    $N=5$ & $0.66$ & $0.21$ & $0.07$ \\ \hline
+    $N=6$ & $0.75$ & $0.22$ & $0.14$ \\ \hline
+    $N=7$ & $0.82$ & $0.36$ & $0.18$ \\ \hline
+    $N=8$ &  $0.86$ & $0.44$ & $0.24$ \\ \hline
+    $N=9$ &  $0.89$ & $0.49$ & $0.26$ \\ \hline
+    $N=10$ & $0.91$ & $0.51$ & $0.32$ \\ \hline
+    $N=11$ & $0.93$  & $0.55$ & $0.35$ \\ \hline
+    $N=12$ & $0.94$ & $0.64$ & $0.41$ \\
+    \hline
+    \end{tabular}
+    \caption {Minimum values of $\epsilon$ for different values of $(n, N)$, as computed by our annealing method}
+\end{table}
+
+The accuracy of our method can be noted by looking at the computed values for $n=2$ -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$.
 \section{Applications to sphere packing}

### 2014-12-09 project3/paper/paper.tex
Changes to the paper -- added code for algorithm, and also corrected error in proof.

@@
 \usepackage{tikz}
+\usepackage[scaled]{beramono}
+\usepackage{listings}
+
+\lstset{
+  language=Python,
+  showstringspaces=false,
+  formfeed=\newpage,
+  tabsize=4,
+  commentstyle=\itshape,
+  basicstyle=\ttfamily,
+  morekeywords={while, if}
+  captionpos=b,
+}
 
@@
 
+\renewcommand{\lstlistingname}{Algorithm}
+
 \usepackage{todonotes}
+\usepackage{caption}
+\DeclareCaptionFormat{myformat}{#1#2#3\hrulefill}
+\captionsetup{format=myformat}
+\captionsetup[lstlisting]{position=bottom,format=myformat}
 
@@
 
-We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ will be upper bounded by
+We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ is upper bounded by
 \[\frac{1}{n^2},\] 
@@
 It's clear that when $N=3$, the value of $\epsilon$ must be greater than $0$, since there is no way three dimensionality-$2$ vectors can be orthogonal to each other.
+We make this more concrete in the following theorem.
 
-In addition, we see that for $N=3$, the quantity $\epsilon$ for three unit-length vectors $v_1, v_2, v_3 \in \mathbb{R}^2$ is given by $$\max \{ |v_1 \cdot v_2|^2, |v_1 \cdot v_3|^2, |v_2 \cdot v_3|^2 \}$$
+\begin{theorem}
+For $n=2$ and $N=3$, the minimum possible $\epsilon$ is equal to $1/4$.
+\end{theorem}
 
-Since we don't care about the sign of the dot product between any two vectors, if we define the unit vectors $v_4, v_5, v_6 \in \mathbb{R}^2$ to be $-v_1, -v_2, -v_3$ respectively, then we see that $\epsilon$ can be equivalently expressed as $$\max_{i \neq j} \{|v_i \cdot v_j |^2 \}$$
-where $i, j \in \{1,2,\ldots,6\}$ -- the above result holds because $$|v \cdot v'| = |(-v) \cdot v'| = |(-v) \cdot (-v')| = |v \cdot (-v')|$$ for some arbitrary vector $v'$.
-Furthermore, we see that there now exists an axis of symmetry -- one of the diagonals passing through either $v_1$ and $v_4 = -v_1$ or $v_2$ and $v_5 = -v_2$ or $v_3$ and $v_6 = -v_3$ -- so we don't need to look at the entire unit circle anymore, and instead can focus on the four vectors contained in a semi-circle resting on one of these three diameters.
+Before proving the above theorem, we state and prove the following lemma.
 
-Without loss of generality, let $x_1 = [1, 0]^T$; then since $x_4$ is $-x_1$, we see that $x_4 = [-1, 0]^T$.
-Also, without loss of generality let us assume that $x_2$ and $x_3$ are above the $x$-axis, and that $x_2$ is to the right of $x_3$.
-
-\begin{figure}[!h]
-    \centering
-    \begin{tikzpicture}[yscale=-1] 
-        % x-axis
-        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
-        % y-axis
-        \draw [thick,->] (0, 4.5) -- (0, -4.5);
-        % origin label
-        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
-        % x-axis label
-        \node at (4.5, 0.5) {\text{$x$}};
-        % y-axis label
-        \node at (0, -5) {\text{$y$}};
-        % circle
-        \draw (0,0) circle (3cm);
-        \draw (3,0)[blue,fill=blue] circle (0.1cm);
-        \draw (1.5, -2.59)[blue,fill=blue] circle (0.1cm);
-        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);
-
-        \draw [thick,-,blue] (3, 0) -- (0, 0);
-        \draw [thick,-,blue] (1.5, -2.59) -- (0, 0);
-        \draw [thick,-,blue] (-1.5, -2.59) -- (0, 0);
-
-        \node at (4, -0.3) {\text{$x_1 = [1,0]$}};
-        \node at (2.8, -3.1) {\text{$x_2 = [\cos \pi/3,\sin \pi/3]$}};
-        \node at (-2.8, -3.1) {\text{$x_3 = [\cos 2\pi/3,\sin 2\pi/3]$}};
-    \end{tikzpicture}
-    \caption{A configuration of three unit vectors that produce the optimum $\epsilon$ for $n=2, N=3$}
-\end{figure}
-
-Let $\theta_1$ be the angle between $x_1$ and $x_2$, $\theta_2$ be the angle between $x_2$ and $x_3$ and $\theta_3$ be the angle between $x_3$ and $x_4$.
-Now, we see that $\theta_1 + \theta_2 + \theta_3 = \pi$ (since vectors $v_1$ and $v_4 = -v_1$ are anti-parallel) and that $$\epsilon = \max_{i \in \{1,2,3\}} \cos ^2 \theta_i$$, since $|x_i \cdot x_j| = \cos \theta$ where $\theta$ is the angle between the two vectors $x_i$ and $x_j$.
-
-Before proceeding, we state and prove the following lemma.
 \begin{lemma}
@@
 $\theta_1 + \theta_2 + \ldots + \theta_n = \pi$.
-Then $\epsilon = \max_i \cos^2 \theta_i$ must equal $\cos^2 \theta_j$ where $j = \text{argmin }\theta_i$.
+Then $\alpha = \max_i \cos^2 \theta_i$ must equal $\cos^2 \theta_j$ where $j = \text{argmin }\theta_i$.
 \end{lemma}
@@
 
-Our arguments will hinge on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing -- this is easy to see from Figure $2$.
+Our arguments will hinge on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing -- this is easy to see from Figure $1$.
 Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \geq\theta_n$.
-Hence our theorem statement is equivalent to proving that $\epsilon$ is equal to $\cos^2 \theta_n$.
+Hence our theorem statement is equivalent to proving that $\alpha$ is equal to $\cos^2 \theta_n$.
 
@@
 \begin{itemize}
-\item $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$: In this case, it is easy to see that $\epsilon = \cos^2 \theta_n$ from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.
+\item $\theta_1, \theta_2, \ldots, \theta_n \in [0, \pi/2]$: In this case, it is easy to see that $\alpha = \cos^2 \theta_n$ from the fact that $\cos \theta$ is a decreasing function in $\theta$ if $\theta \in [0, \pi/2]$.
 
@@
 
-In addition, we see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\epsilon$ is equal to $\cos^2 \theta_n$.
+In addition, we see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\alpha$ is equal to $\cos^2 \theta_n$.
 
@@
 \end{proof}
+
+Given this lemma, we now prove the above theorem.
+
+\begin{proof}
+Observe that for $N=3$, the quantity $\epsilon$ for three unit-length vectors $v_1, v_2, v_3 \in \mathbb{R}^2$ is given by $$\max \{ |v_1 \cdot v_2|^2, |v_1 \cdot v_3|^2, |v_2 \cdot v_3|^2 \}$$
+
+Since we don't care about the sign of the dot products between any two vectors (we're looking at the squares of the dot products in the above equation), and since $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin.
+
+Without loss of generality, let $v_1 = [1, 0]^T$.
+Also, without loss of generality let us assume that $v_2$ and $v_3$ are above the $x$-axis, and that $v_1$, $v_2$ and $v_3$ are in anti-clockwise order.
+
+\begin{figure}[!h]
+    \centering
+    \begin{tikzpicture}[yscale=-1] 
+        % x-axis
+        \draw [thick,->] (-4.5, 0) -- (4.5, 0);
+        % y-axis
+        \draw [thick,->] (0, 4.5) -- (0, -4.5);
+        % origin label
+        \node at (-0.5, 0.3) {\text{$(0, 0)$}};
+        % x-axis label
+        \node at (4.5, 0.5) {\text{$x$}};
+        % y-axis label
+        \node at (0, -5) {\text{$y$}};
+        % circle
+        \draw (0,0) circle (3cm);
+        \draw (3,0)[blue,fill=blue] circle (0.1cm);
+        \draw (1.5, -2.59)[blue,fill=blue] circle (0.1cm);
+        \draw (-1.5, -2.59)[blue,fill=blue] circle (0.1cm);
+
+        \draw [thick,-,blue] (3, 0) -- (0, 0);
+        \draw [thick,-,blue] (1.5, -2.59) -- (0, 0);
+        \draw [thick,-,blue] (-1.5, -2.59) -- (0, 0);
+
+        \node at (4, -0.3) {\text{$v_1 = [1,0]$}};
+        \node at (2.8, -3.1) {\text{$v_2 = [\cos \pi/3,\sin \pi/3]$}};
+        \node at (-2.8, -3.1) {\text{$v_3 = [\cos 2\pi/3,\sin 2\pi/3]$}};
+    \end{tikzpicture}
+    \caption{A configuration of three unit vectors that produce the optimum $\epsilon$ for $n=2, N=3$}
+\end{figure}
+
+Let $\theta_1$ be the angle between $v_1$ and $v_2$ and $\theta_2$ be the angle between $v_2$ and $v_3$.
+Let us now define $\theta_3$ such that $\theta_1 + \theta_2 + \theta_3 = \pi$.
+Observe that $|v_1 \cdot v_3|^2 = \cos^2(\theta_1 + \theta_2) = \cos^2 (\pi - \theta_3) = \cos^2 \theta_3$, hence we can conclude that
+
+$$\epsilon = \max_{i \in \{1,2,3\}} \cos ^2 \theta_i$$
 
@@
 Figure $1$ shows this optimum configuration.
+
+\end{proof}
 
@@
 
-Again without loss of generality, we can assume that the vectors $x_1, x_2, \ldots, x_N$ are on or above the $x$-axis, and that $x_1 = [1,0]^T$ and that $x_{N+1} = [-1,0]^T$ -- if any vector $x_i$ were not above the $x$-axis, then we could just consider $-x_i$ instead.
+Again without loss of generality, we can assume that the vectors $v_1, v_2, \ldots, v_N$ are on or above the $x$-axis, and that $v_1 = [1,0]^T$ -- if any vector $v_i$ were not above the $x$-axis, then we could just consider $-v_i$ (the reflection of $v_i$ about the $x$-axis) instead.
 
-Let us define $\theta_i$ as the angle between the vectors $x_i$ and $x_{i+1}$.
-Then, we quickly see that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$.
-If $j = \text{argmin } \theta_i$, $\epsilon = \cos^2 \theta_j$ by the above Lemma.
+Let us define $\theta_i$ as the angle between the vectors $v_i$ and $v_{i+1}$ for $i \in \{1, 2, \ldots, N-1\}$, and let $\theta_N$ be the angle such that $\theta_1 + \theta_2 + \ldots + \theta_N = \pi$.
 
-Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$ (again making use of the fact that the cosine function is decreasing).
+Observe that as before, we are interested in maximizing the square of the cosine of the angle between any two vectors in $v_1, v_2, \ldots, v_N$.
+Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\cos^2 (\sum_{i=j} ^{k-1} \theta_i)$.
+Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ not in $\{j, j+1, \ldots, k-1\}$.
+
+From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by the above Lemma.
+
+Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$ (again making use of the fact that the cosine function is decreasing between $0$ and $\pi/2$).
 \end{proof}
@@
 
-\section{The Parallelogram Law}
+\section{$N = n+1$}
+
+We now look to calculate the optimum value of $\epsilon$ for the specific case of $N = n+1$.
+Before proving an upper bound on what the optimum value of $\epsilon$ can be, we look at the parallelogram law.
+
+\subsection{The Parallelogram Law}
 
@@
 
-\section{$N = n+1$}
+\subsection{Upper bound on minimum possible $\epsilon$}
 Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space.
@@
 \section{Simulated annealing and associated conjectures}
+
 Since visualizing in spaces of dimensionality greater than $3$ is extremely hard, we used an annealing approach to compute near-optimal epsilon values for different $(n, N)$ pairs.
@@
 Our general intuition tells us that perturbing a set of nearly optimal vectors slightly could give us a new set of vectors that produce even a smaller epsilon value.
-Given this, we start off with a set of random vectors, and try to move these vectors in order to produce smaller epsilon values.
+Given this, we start off with a set of random vectors, and try to move these vectors towards smaller epsilon values.
 
-At every iteration, we maintain the set of vectors (called cur\_vectors) that have produced the optimal epsilon value so far. Then we compute a set new\_vectors defined as follows,
-$$\text{new\_vector} = \text{old\_vector} + (1/i \cdot \text{random\_vector})$$
+At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector.
 
-We update cur\_vectors to the set new\_vectors if the epsilon value produced by new\_vectors is less than the epsilon value produced by cur\_vectors.
-$i$ is incremented whenever cur\_vectors is updated -- having the $1/i$ factor in the above expression ensures that as we move closer to an optimal value, we perturb the set of vectors by a smaller amount.
+\begin{lstlisting}
+def get_min_epsilon_annealing(n, N, num_iter, start_vectors):
+    min_epsilon = 1.0  # Start off with the worst possible epsilon
+    current_vectors = start_vectors
+    i = 0
+    while (i < num_iter):
+        temperature = 1.0 / float(i + 1)
+
+        # Get perturbing vectors
+        random_vectors = get_random_vectors(n, N)
+
+        # Now perturb current_vectors
+        new_vectors = list()
+        for j in xrange(N):
+            new_vector = list()
+            for k in xrange(n):
+                new_vector.append(
+                    current_vectors[j][k] + (
+                    	temperature * random_vectors[j][k]))
+            new_vectors.append(new_vector)
+        normalize_vectors(new_vectors)
+
+        # Accept change only if new_vectors produces a better epsilon
+        epsilon = compute_epsilon(new_vectors)
+        if epsilon < min_epsilon:
+            min_epsilon = epsilon
+            current_vectors = new_vectors
+            i += 1
+
+    return min_epsilon, current_vectors
+\end{lstlisting}
+\captionof{lstlisting}{The annealing algorithm used to generate optimum epsilon values}
+
+To make this more precise, we present Python code for the algorithm above.
 
@@
 To make this more concrete, consider Table $1$.
+
+The accuracy of our method can be noted by looking at the computed values for $n=2$ -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$, that we already proved in Section $2$, which makes us confident about the accuracy of this numeric method.
+\section{Applications to sphere packing}
+We should be able to prove something about the radius of spheres you can pack in the case of $N=n+1$, taking as intuition square and hexagonal tilings of the plane and cubic and tetrahedral tilings of 3-space.
+
 \begin{table}
@@
 
-The accuracy of our method can be noted by looking at the computed values for $n=2$ -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$.
-\section{Applications to sphere packing}
-We should be able to prove something about the radius of spheres you can pack in the case of $N=n+1$, taking as intuition square and hexagonal tilings of the plane and cubic and tetrahedral tilings of 3-space.
-
 \end{document}

### 2014-12-09 project3/paper/paper.tex
added the lower bound section

@@
 The accuracy of our method can be noted by looking at the computed values for $n=2$ -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$, that we already proved in Section $2$, which makes us confident about the accuracy of this numeric method.
+
+\section{A lower bound}
+
+We can use a well-known lemma to provide a useful lower bound on $\epsilon$.
+Our presentation of the result follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.
+
+\begin{lemma}
+Let $A=(a_{i,j})$ be an $n$-by-$n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
+\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
+\end{lemma}
+\begin{proof}
+We define $\lambda_1,\ldots, \lambda_n$ to be the eigenvalues of $A$; write $k$ as the number of nonzero eigenvalues.
+Their sum is the trace of $A$, equal to $n$.
+Recall that the number of nonzero eigenvalues of a complex matrix is less than or equal to the rank of the matrix: so we have that $d\geq k$.
+
+Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
+\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
+In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
+\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
+so that
+\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
+In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
+\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
+But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
+\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
+We obtain, therefore, that
+\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
+and it clearly follows that
+\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
+\end{proof}
+We can now prove our lower bound. 
+\begin{theorem}
+\label{lower_bound}
+For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
+\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
+\end{theorem}
+\begin{proof}
+Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Let us define the $n$-by-$N$ matrix
+\[A=\left(\begin{matrix}
+| & | & \cdots & | & |\\
+v_1 & v_2 & \cdots & v_{N-1} & v_N\\
+| & | & \cdots & | & | \end{matrix}\right).\] 
+Now consider the $N$-by-$N$ Gram matrix:
+\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
+v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
+v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
+\vdots & \vdots & \ddots & \vdots\\
+v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
+Let $d$ be the rank of $G$; since $A$ has rank at most $n$, we have that $d\leq n$.
+Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$.
+We can apply the lemma to conclude
+\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
+Since $d\leq n$, we have as desired
+\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
+\end{proof}
+For $N=n+1$, for instance, the theorem provides us a bound of 
+\[\epsilon\geq \frac{(n+1)-n}{n(n+1-1)}=\frac 1{n^2}\]
+implying that the $n$-simplex is indeed an optimal configuration. Similarly, for $n=3, N=6$, the theorem gives us
+\[\epsilon\geq \frac{6-3}{3(6-1)}=\frac{1}{5}.\]
+Again, the configuration we obtained was optimal. We present now a table of the bounds the theorem gives to us for varying $n$ and $N$:
+\begin{table}[h]
+   \centering
+    \begin{tabular}{ | c | c | c | c |}
+    \hline
+      & $n=2$ & $n=3$ & $n=4$ \\ \hline
+    $N=2$ & $0$ & $-$ & $-$ \\ \hline
+    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
+    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
+    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
+    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
+    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
+    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
+    \hline
+    \end{tabular}
+    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
+    \end{table}
+
+Note that the bounds for $n=2$ are not tight, as we proved in Section 2 that the minimum $\epsilon$ was given by $\cos^2(\pi/N)$.
+One direction for further research would be checking whether the bounds provided by Theorem \ref{lower_bound} are indeed tight for other combinations of $n$ and $N$.
+
 \section{Applications to sphere packing}

### 2014-12-09 project3/paper/paper.tex
moved the section + minor edits

@@
 Note by taking reflections of $x_2$ and $x_4$ in Figure $3$, we get five vectors that form a regular $5$-gon -- this can be seen in Figure $4$.
-A similar transformation can be done for all odd $N$; unfortunately, even $N$ cannot produce regular polygons.
+A similar transformation can be done for all odd $N$; unfortunately, even $N$ do not produce regular polygons.
+
+\section{A lower bound}
+
+We can use a well-known lemma to provide a useful lower bound on $\epsilon$.
+Our proof of the lemma follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.
+
+\begin{lemma}
+Let $A=(a_{i,j})$ be an $n$-by-$n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
+\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
+\end{lemma}
+\begin{proof}
+We define $\lambda_1,\ldots, \lambda_n$ to be the eigenvalues of $A$; write $k$ as the number of nonzero eigenvalues.
+Their sum is the trace of $A$, equal to $n$.
+We know that the number of nonzero eigenvalues of a complex matrix is less than or equal to the rank of the matrix: so we have that $d\geq k$.
+
+Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
+\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
+In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
+\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
+so that
+\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
+In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
+\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
+But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
+\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
+We obtain, therefore, that
+\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
+and it clearly follows that
+\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
+\end{proof}
+Using this lemma, we can now prove our lower bound on $\epsilon$.
+\begin{theorem}
+\label{lower_bound}
+For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
+\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
+\end{theorem}
+\begin{proof}
+Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Let us define the $n$-by-$N$ matrix
+\[A=\left(\begin{matrix}
+| & | & \cdots & | & |\\
+v_1 & v_2 & \cdots & v_{N-1} & v_N\\
+| & | & \cdots & | & | \end{matrix}\right).\] 
+Now consider the $N$-by-$N$ Gram matrix:
+\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
+v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
+v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
+\vdots & \vdots & \ddots & \vdots\\
+v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
+Let $d$ be the rank of $G$; since $A$ has rank at most $n$, we have that $d\leq n$.
+Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$.
+We can apply the lemma to conclude
+\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
+Since $d\leq n$, we have as desired
+\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
+\end{proof}
+We present now a table of the bounds the theorem gives to us for varying $n$ and $N$:
+\begin{table}[h]
+   \centering
+    \begin{tabular}{ | c | c | c | c |}
+    \hline
+      & $n=2$ & $n=3$ & $n=4$ \\ \hline
+    $N=2$ & $0$ & $-$ & $-$ \\ \hline
+    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
+    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
+    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
+    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
+    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
+    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
+    \hline
+    \end{tabular}
+    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
+    \end{table}
+
+Note that the bounds for $n=2$ are not tight, as we proved in Section 2 that the minimum $\epsilon$ was given by $\cos^2(\pi/N)$.
+We shall see, however, that the bounds for $\epsilon$ are in fact tight in the cases where $N=n+1$ and where $n=3, N=6$.
+One direction for further research might be checking whether the bounds provided by Theorem \ref{lower_bound} are indeed attainable for other combinations of $n$ and $N$.
 
@@
 
-\section{A lower bound}
-
-We can use a well-known lemma to provide a useful lower bound on $\epsilon$.
-Our presentation of the result follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.
-
-\begin{lemma}
-Let $A=(a_{i,j})$ be an $n$-by-$n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
-\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
-\end{lemma}
-\begin{proof}
-We define $\lambda_1,\ldots, \lambda_n$ to be the eigenvalues of $A$; write $k$ as the number of nonzero eigenvalues.
-Their sum is the trace of $A$, equal to $n$.
-Recall that the number of nonzero eigenvalues of a complex matrix is less than or equal to the rank of the matrix: so we have that $d\geq k$.
-
-Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
-\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
-In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
-\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
-so that
-\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
-In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
-\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
-But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
-\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
-We obtain, therefore, that
-\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
-and it clearly follows that
-\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
-\end{proof}
-We can now prove our lower bound. 
-\begin{theorem}
-\label{lower_bound}
-For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
-\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
-\end{theorem}
-\begin{proof}
-Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Let us define the $n$-by-$N$ matrix
-\[A=\left(\begin{matrix}
-| & | & \cdots & | & |\\
-v_1 & v_2 & \cdots & v_{N-1} & v_N\\
-| & | & \cdots & | & | \end{matrix}\right).\] 
-Now consider the $N$-by-$N$ Gram matrix:
-\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
-v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
-v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
-\vdots & \vdots & \ddots & \vdots\\
-v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
-Let $d$ be the rank of $G$; since $A$ has rank at most $n$, we have that $d\leq n$.
-Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$.
-We can apply the lemma to conclude
-\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
-Since $d\leq n$, we have as desired
-\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
-\end{proof}
-For $N=n+1$, for instance, the theorem provides us a bound of 
-\[\epsilon\geq \frac{(n+1)-n}{n(n+1-1)}=\frac 1{n^2}\]
-implying that the $n$-simplex is indeed an optimal configuration. Similarly, for $n=3, N=6$, the theorem gives us
-\[\epsilon\geq \frac{6-3}{3(6-1)}=\frac{1}{5}.\]
-Again, the configuration we obtained was optimal. We present now a table of the bounds the theorem gives to us for varying $n$ and $N$:
-\begin{table}[h]
-   \centering
-    \begin{tabular}{ | c | c | c | c |}
-    \hline
-      & $n=2$ & $n=3$ & $n=4$ \\ \hline
-    $N=2$ & $0$ & $-$ & $-$ \\ \hline
-    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
-    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
-    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
-    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
-    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
-    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
-    \hline
-    \end{tabular}
-    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
-    \end{table}
-
-Note that the bounds for $n=2$ are not tight, as we proved in Section 2 that the minimum $\epsilon$ was given by $\cos^2(\pi/N)$.
-One direction for further research would be checking whether the bounds provided by Theorem \ref{lower_bound} are indeed tight for other combinations of $n$ and $N$.
-
 \section{Applications to sphere packing}

### 2014-12-09 project3/paper/paper.tex
edits to intro

@@
 For $n=N$, it is clear that the minimum value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis.
-For fixed $n$, it is also easy to see that the minimum value of $\epsilon$ is nondecreasing as $N$ increases: that is, if $\epsilon_1$ is the minimum value of $\epsilon$ for $N$ unit vectors in $\mathbb R^n$, and $\epsilon_2$ is the minimum value of $\epsilon$ for $N+1$ unit vectors in $\mathbb R^n$, we must have $\epsilon_1\leq \epsilon_2$.
+For fixed $n$, it is also easy to see that the minimum value of $\epsilon$ is nondecreasing as $N$ increases: that is, if $\epsilon_1$ is the minimum value of $\epsilon$ for $N$ unit vectors in $\mathbb R^n$, and $\epsilon_2$ is the minimum value of $\epsilon$ for $N+1$ unit vectors in $\mathbb R^n$, we must have
+\[\epsilon_1\leq \epsilon_2.\] 
 Otherwise, given a configuration of $N+1$ vectors in $\mathbb R^n$ with $\epsilon=\epsilon_2$, we could simply remove one vector and obtain a configuration of $N$ vectors whose $\epsilon$ would be less than or equal to $\epsilon_1$.
 
-Furthermore, given a configuration of $N$ vectors in $n$-space with corresponding $\epsilon$, we can obtain a collection of $N+1$ vectors in $\mathbb R^{n+1}$ with the same $\epsilon$: simply embed the first $N$ vectors in $\mathbb R^{n}$, then add any vector perpendicular to the hyperplane in which the $N$ vectors lie.
-Hence the minimum $\epsilon$ for collections of $N+1$ vectors in $\mathbb R^{n+1}$ is upper-bounded by the corresponding $\epsilon$ for $N$ vectors in $\mathbb R^n$.
+Along the same lines, given a configuration of $N$ vectors in $n$-space with corresponding $\epsilon$, we can obtain a collection of $N+1$ vectors in $\mathbb R^{n+1}$ with the same $\epsilon$: simply embed the first $N$ vectors in $\mathbb R^{n}$, then add any vector perpendicular to the hyperplane in which the $N$ vectors lie.
+Hence the minimum $\epsilon_3$ for collections of $N+1$ vectors in $\mathbb R^{n+1}$ is upper-bounded by the corresponding $\epsilon$ for $N$ vectors in $\mathbb R^n$.
 
-In this paper, we shall see that, for $n=2$, the minimum value of $\epsilon$ will be given by 
+In this paper, we shall see that for $n=2$, the minimum value of $\epsilon$ will be given by 
 \[\epsilon=\cos^2\pi/N.\] 
-
-We shall also see that, for $N=n+1$, the minimum value of $\epsilon$ is upper bounded by
-\[\frac{1}{n^2},\] 
+We will also derive a lower bound on $\epsilon$ for any choice of $n$ and $N$. In addition, we shall see that, for $N=n+1$, the minimum value of $\epsilon$ is given by
+\[\epsilon=\frac{1}{n^2},\] 
 and this configuration is produced when the $n+1$ vectors point to the vertices of the $n$-simplex.
+We shall also exhibit an optimal configuration of six vectors in $\mathbb R^3$, with corresponding $\epsilon$ given by
+\[\epsilon=\frac 1 5.\]
 

### 2014-12-10 project3/paper/paper.tex
Paper changes

@@
 Along the same lines, given a configuration of $N$ vectors in $n$-space with corresponding $\epsilon$, we can obtain a collection of $N+1$ vectors in $\mathbb R^{n+1}$ with the same $\epsilon$: simply embed the first $N$ vectors in $\mathbb R^{n}$, then add any vector perpendicular to the hyperplane in which the $N$ vectors lie.
-Hence the minimum $\epsilon_3$ for collections of $N+1$ vectors in $\mathbb R^{n+1}$ is upper-bounded by the corresponding $\epsilon$ for $N$ vectors in $\mathbb R^n$.
+Hence the minimum $\epsilon$ for collections of $N+1$ vectors in $\mathbb R^{n+1}$ is upper-bounded by the corresponding $\epsilon$ for $N$ vectors in $\mathbb R^n$.
 
@@
 Observe that as before, we are interested in maximizing the square of the cosine of the angle between any two vectors in $v_1, v_2, \ldots, v_N$.
-Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\cos^2 (\sum_{i=j} ^{k-1} \theta_i)$.
+Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$.
 Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ not in $\{j, j+1, \ldots, k-1\}$.
@@
 Note by taking reflections of $x_2$ and $x_4$ in Figure $3$, we get five vectors that form a regular $5$-gon -- this can be seen in Figure $4$.
-A similar transformation can be done for all odd $N$; unfortunately, even $N$ do not produce regular polygons.
+A similar transformation can be done for all odd $N$.
 
@@
 
-\section{Simulated annealing and associated conjectures}
+\section{Numerical analysis and associated conjectures}
 
-Since visualizing in spaces of dimensionality greater than $3$ is extremely hard, we used an annealing approach to compute near-optimal epsilon values for different $(n, N)$ pairs.
+Since visualizing in spaces of dimensionality greater than $3$ is extremely hard, we used a computational approach to compute near-optimal epsilon values for different $(n, N)$ pairs.
 We describe this method in greater detail below.
 
-Our general intuition tells us that perturbing a set of nearly optimal vectors slightly could give us a new set of vectors that produce even a smaller epsilon value.
-Given this, we start off with a set of random vectors, and try to move these vectors towards smaller epsilon values.
-
-At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector.
-
 \begin{lstlisting}
-def get_min_epsilon_annealing(n, N, num_iter, start_vectors):
+def get_min_epsilon(n, N, num_iter, start_vectors):
     min_epsilon = 1.0  # Start off with the worst possible epsilon
@@
 \end{lstlisting}
-\captionof{lstlisting}{The annealing algorithm used to generate optimum epsilon values}
+\captionof{lstlisting}{The numerical method used to generate optimum epsilon values -- note that we call this method multiple times with different start\_vectors to obtain a reasonable epsilon estimate}
+
+Our general intuition tells us that perturbing a set of nearly optimal vectors slightly could give us a new set of vectors that produce even a smaller epsilon value.
+Given this, we start off with a set of random vectors, and try to move these vectors towards smaller epsilon values.
+
+At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector.
 
@@
 
-Even though this approach seems simple, it actually produces remarkably precise values -- this can be verified by cross-checking with results we've already proven / conjectured in this paper.
+Even though this approach seems simple, it actually produces remarkably precise values -- this can be verified by cross-checking with results we've already proven / conjectured in this paper, this can be seen from values in Table $2$.
 
-To make this more concrete, consider Table $1$.
-
-The accuracy of our method can be noted by looking at the computed values for $n=2$ -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$, that we already proved in Section $2$, which makes us confident about the accuracy of this numeric method.
+In particular, the computed values for $n=2$ are pretty accurate -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$, that we already proved in Section $2$, which makes us confident about the accuracy of this numeric method.
 

### 2014-12-10 project3/paper/paper.tex
resctructured, added 3,6 and n,n+1'

@@
 \maketitle
+\tableofcontents
 
@@
 
-\section{$N = n+1$}
-
-We now look to calculate the optimum value of $\epsilon$ for the specific case of $N = n+1$.
-Before proving an upper bound on what the optimum value of $\epsilon$ can be, we look at the parallelogram law.
-
-\subsection{The Parallelogram Law}
-
-First we give the definition of the norm on vectors in $\R^n$ in terms of the dot product:
-
-\begin{definition}
-For a vector $v\in\R^n$ the norm of $v$ is the positive value of $||v||$ given by the equation
+\section{Some Optimal Configurations}
+Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
+\subsection{n=3,N=6}
+In this situation we have a lower bound of $\epsilon\ge\frac{1}{5}$. Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
 $$
-||v||^2=v\cdot v.
+(a,b,0),(a,0,b),(b,0,a)
 $$
-\end{definition}
-\begin{theorem}[Parallelogram Law]
-For vectors $v_1,\cdots,v_n\in\R^n$, the following identity holds
 $$
-||v_1+\cdots+v_n||^2=\displaystyle\sum_{i=1}^n||v_i||^2+2\displaystyle\sum_{1\le i<j\le n}v_i\cdot v_j.
+(a,-b,0),(a,0,-b),(-b,0,a)
 $$
-\proof
-First consider the following identity from the definition of the norm for vectors $u,v\in\R^n$:
+There are four possible dot products we get from these six vectors: $1,ab,-ab,a^2-b^2$.
+This gives us three possible values of $\epsilon$: $1,(a^2-b^2)^2,(ab)^2$.
+Let us consider the solution we get when $a^2-b^2=ab$.
+If we solve the resulting system according to the constraint $a^2+b^2=1$, we get
 $$
-||u+v||^2=(u+v)\cdot (u+v)=u\cdot u+u\cdot v+ v\cdot u + v\cdot v=||u||^2+||v||^2+2u\cdot v.
+a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
 $$
-Now if we let $u=v_1$ and $v= v_2+\cdots+v_n$ we get 
-$$
-||v_1+v_2+\cdots+v_n||^2=||v_1||^2+||v_2+\cdots +v_n||^2+2\sum_{i=2}^n v_1\cdot v_i.
-$$
-If we recurse on $||v_2+\cdots +v_n||^2$ using induction we get the desired result. 
-\end{theorem}
-
-\subsection{Upper bound on minimum possible $\epsilon$}
+If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we have that $\epsilon=\frac{1}{5}$, implying that these six points give us an optimal configuration.
+\subsection{N=n+1}
 Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space.
+From Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we will find a configuration of vectors that achieves this lower bound.
 First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$.
@@
 \begin{theorem}
-If $N=n+1$ and the vectors in question are denoted $v_1,\cdots, v_N$, then over all configurations of these vectors the maximum value of $v_i\cdot v_j$ for $i\neq j$ is $\frac{-1}{n}$.
-This is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin, and in this configuration $\epsilon=\frac{1}{n^2}$.
+If $N=n+1$ and the vectors in question are denoted $v_1,\cdots, v_N$, then over all configurations of these vectors the minimum value of $\epsilon$ for $i\neq j$ is $\frac{1}{^2n}$.
+This is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
+\label{simplex}
 \end{theorem}
@@
 \begin{proof}
-Let $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$, and let $\delta_{min}$ be the minimum possible $\delta$ obtained across all unit vectors $v_1, v_2, \ldots, v_{n+1}$.
-We first show that $\delta_{min} \le-\frac{1}{n}$ by showing that the origin-centered simplex configuration of the vectors $v_1,\cdots v_{n+1}$ produces a value $\delta=-\frac{1}{n}$, and then will later show that $\delta_{min} \geq -\frac{1}{n}$ as well.
+Let $\epsilon$ be as before for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$.
+We show that the origin-centered simplex configuration of the vectors $v_1,\cdots v_{n+1}$ produces a value $\epsilon=\frac{1}{n^2}$.
 
@@
 $$
-So an upper bound on the value of $\delta_{min}$ for all configurations of $n+1$ vectors in $n+1$ dimensional space is $\frac{-1}{n}$.
-
-To see that $\delta_{min} \ge-\frac{1}{n}$ we use the parallelogram law.
-
-Recall that the parallelogram law states that
-
-$$
-||v_1+\cdots+v_{n+1}||^2=\displaystyle\sum_{i=1}^{n+1}||v_i||^2+2\displaystyle\sum_{1\le i<j\le n+1}v_i\cdot v_j.
-$$
-In our case we know three things.
-First, since the norm of any vector is greater than or equal to $0$, we can conclude that $||v_1 + \cdots + v_{n+1}||^2 \geq 0$.
-Second, $v_i\cdot v_j\le\delta$ for any $i<j$.
-Third, $||v_i||=1$.
-Therefore
-$$
-0\le n+1+n(n+1)\delta.
-$$
-This implies that $\delta\ge -\frac{1}{n}$ for any set of unit vectors $v_1, v_2, \ldots, v_{n+1}$, from which we can conclude that $\delta_{min} \ge -\frac{1}{n}$ as well.
-
-Since $\delta_{min} \le -\frac{1}{n}$ and $\delta_{min} \ge -\frac{1}{n}$, we can conclude that $\delta_{min}=-\frac{1}{n}$.
-From this, we can see that a $\epsilon = \frac{1}{n^2}$ is achievable for the general case of when $N=n+1$ -- this leads us to conclude that the minimum possible $\epsilon$ value for $N=n+1$ is at most $\frac{1}{n^2}$.
+So for this configuration $\epsilon=\frac{1}{n^2}$. By Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we have achieved the lower bound. 
 \end{proof}
-
-Note that proving that the minimum possible epsilon for the case of $N = n+1$ is equal to $1/n^2$ is much harder, and we leave this for future work.
-
 \section{Numerical analysis and associated conjectures}
@@
 
-\section{Applications to sphere packing}
-We should be able to prove something about the radius of spheres you can pack in the case of $N=n+1$, taking as intuition square and hexagonal tilings of the plane and cubic and tetrahedral tilings of 3-space.
-
 \begin{table}
@@
 
+\section{Variations}
+
+We now look to calculate the optimum value of $\epsilon$ for the specific case of $N = n+1$.
+Before proving an upper bound on what the optimum value of $\epsilon$ can be, we introduce the parallelogram law.
+
+\subsection{The Parallelogram Law}
+
+First we give the definition of the norm on vectors in $\R^n$ in terms of the dot product:
+
+\begin{definition}
+For a vector $v\in\R^n$ the norm of $v$ is the positive value of $||v||$ given by the equation
+$$
+||v||^2=v\cdot v.
+$$
+\end{definition}
+\begin{theorem}[Parallelogram Law]
+For vectors $v_1,\cdots,v_n\in\R^n$, the following identity holds
+$$
+||v_1+\cdots+v_n||^2=\displaystyle\sum_{i=1}^n||v_i||^2+2\displaystyle\sum_{1\le i<j\le n}v_i\cdot v_j.
+$$
+\proof
+First consider the following identity from the definition of the norm for vectors $u,v\in\R^n$:
+$$
+||u+v||^2=(u+v)\cdot (u+v)=u\cdot u+u\cdot v+ v\cdot u + v\cdot v=||u||^2+||v||^2+2u\cdot v.
+$$
+Now if we let $u=v_1$ and $v= v_2+\cdots+v_n$ we get 
+$$
+||v_1+v_2+\cdots+v_n||^2=||v_1||^2+||v_2+\cdots +v_n||^2+2\sum_{i=2}^n v_1\cdot v_i.
+$$
+If we recurse on $||v_2+\cdots +v_n||^2$ using induction we get the desired result. 
+\end{theorem}
+
+\subsection{Variations: Minimizing the dot product}
+Let us consider the situation in which we try to minimize $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for $n+1$ vectors in an $n$ dimensional space.
+
+\begin{theorem}
+If $N=n+1$ and the vectors in question are denoted $v_1,\cdots, v_N$, then over all configurations of these vectors the minimum value of $v_i\cdot v_j$ for $i\neq j$ is $\frac{-1}{n}$.
+This is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
+\end{theorem}
+
+\begin{proof}
+Let $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$, and let $\delta_{min}$ be the minimum possible $\delta$ obtained across all unit vectors $v_1, v_2, \ldots, v_{n+1}$.
+We computed in the proof of Theorem ~\ref{simplex} that for the origin centered simplex, we obtain a value of $\delta_{\min}=\frac{-1}{n}$, so $\delta_{\min}\le\frac{-1}{n}$.
+So we need only show that $\delta_{min} \geq -\frac{1}{n}$ as well.
+
+To see that $\delta_{min} \ge-\frac{1}{n}$ we use the parallelogram law.
+
+Recall that the parallelogram law states that
+
+$$
+||v_1+\cdots+v_{n+1}||^2=\displaystyle\sum_{i=1}^{n+1}||v_i||^2+2\displaystyle\sum_{1\le i<j\le n+1}v_i\cdot v_j.
+$$
+In our case we know three things.
+First, since the norm of any vector is greater than or equal to $0$, we can conclude that $||v_1 + \cdots + v_{n+1}||^2 \geq 0$.
+Second, $v_i\cdot v_j\le\delta$ for any $i<j$.
+Third, $||v_i||=1$.
+Therefore
+$$
+0\le n+1+n(n+1)\delta.
+$$
+This implies that $\delta\ge -\frac{1}{n}$ for any set of unit vectors $v_1, v_2, \ldots, v_{n+1}$, from which we can conclude that $\delta_{min} \ge -\frac{1}{n}$ as well.
+
+Since $\delta_{min} \le -\frac{1}{n}$ and $\delta_{min} \ge -\frac{1}{n}$, we can conclude that $\delta_{min}=-\frac{1}{n}$.
+From this, we can see that a $\epsilon = \frac{1}{n^2}$ is achievable for the general case of when $N=n+1$ -- this leads us to conclude that the minimum possible $\epsilon$ value for $N=n+1$ is at most $\frac{1}{n^2}$.
+\end{proof}
+
+Note that proving that the minimum possible epsilon for the case of $N = n+1$ is equal to $1/n^2$ is much harder, and we leave this for future work.
+
 \end{document}

### 2014-12-10 project3/paper/paper.tex
Edits to paper

@@
 \maketitle
-\tableofcontents
 
@@
 \[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
-In this paper, we consider the problem of minimizing $\epsilon$ for various $n$ and $N$.
+In this paper, we consider the problem of minimizing $\epsilon$ for different values of $n$ and $N$.
 
@@
 \[\epsilon_1\leq \epsilon_2.\]
+
 Otherwise, given a configuration of $N+1$ vectors in $\mathbb R^n$ with $\epsilon=\epsilon_2$, we could simply remove one vector and obtain a configuration of $N$ vectors whose $\epsilon$ would be less than or equal to $\epsilon_1$.
@@
 \[\epsilon=\cos^2\pi/N.\]
+
 We will also derive a lower bound on $\epsilon$ for any choice of $n$ and $N$. In addition, we shall see that, for $N=n+1$, the minimum value of $\epsilon$ is given by
@@
 and this configuration is produced when the $n+1$ vectors point to the vertices of the $n$-simplex.
-We shall also exhibit an optimal configuration of six vectors in $\mathbb R^3$, with corresponding $\epsilon$ given by
+We shall also exhibit an optimal configuration of six vectors in $\mathbb R^3$; this optimal configuration produces an $\epsilon$ equal to
 \[\epsilon=\frac 1 5.\]
@@
 
-However, we see that the problem gets harder for larger $N$. Let us first consider the specific case of $N=3$; once we build some intuition for the problem, we will try to generalize to generic $N$.
+However, we see that the problem gets harder for larger $N$. Let us first consider the specific case of $N=3$; once we build some intuition for the problem, we will try to generalize to arbitrary $N$.
 \\
@@
 
-Our arguments will hinge on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing -- this is easy to see from Figure $1$.
-Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \geq\theta_n$.
-Hence our theorem statement is equivalent to proving that $\alpha$ is equal to $\cos^2 \theta_n$.
+Our argument hinges on the fact that for all $\theta \in [0, \pi/2]$, the function $\cos \theta$ is decreasing -- this is easy to see from Figure $1$.
+
+Without loss of generality, let us assume that $\theta_1 \geq \theta_2 \geq \ldots \geq\theta_n$. Hence our theorem statement is equivalent to proving that $\alpha$ is equal to $\cos^2 \theta_n$.
 
@@
 
-Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$.
+Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ which is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$.
 Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$.
@@
 
-Given this lemma, we now prove the above theorem.
+Given this lemma, we now prove the theorem stated above.
 
@@
 
-Since we don't care about the sign of the dot products between any two vectors (we're looking at the squares of the dot products in the above equation), and since $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin.
+Since we don't care about the sign of the dot products between any two vectors (we're looking at the squares of the dot products in the above equation), and since $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin (multiplied by $-1$).
 
@@
 Since we're interested in the smallest such $\epsilon$ and since the cosine function is a decreasing function in $\theta$ between $0$ and $\pi/2$, we conclude that the optimum value of $\epsilon$ for $n=2$ and $N=3$ is $\cos^2 \pi / 3$.
-Figure $1$ shows this optimum configuration.
+Figure $2$ shows this optimum configuration.
 
@@
 Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$.
-Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ not in $\{j, j+1, \ldots, k-1\}$.
+Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\}$$, where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$.
 
-From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by the above Lemma.
+From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.
 
@@
 
-We can use a well-known lemma to provide a useful lower bound on $\epsilon$.
+We can use a well-known lemma to provide a useful lower bound on $\epsilon$ for arbitrary $(n, N)$ pairs.
 Our proof of the lemma follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.
@@
 Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
-\subsection{n=3,N=6}
-In this situation we have a lower bound of $\epsilon\ge\frac{1}{5}$. Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
+\subsection{$n=3,N=6$}
+In this situation we have a lower bound of $\epsilon\ge\frac{1}{5}$ from the above theorem. Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
 $$
@@
 $$
-There are four possible dot products we get from these six vectors: $1,ab,-ab,a^2-b^2$.
-This gives us three possible values of $\epsilon$: $1,(a^2-b^2)^2,(ab)^2$.
-Let us consider the solution we get when $a^2-b^2=ab$.
-If we solve the resulting system according to the constraint $a^2+b^2=1$, we get
+There are three possible dot products we get from computing dot products between non-equal vectors chosen from these six vectors: $ab,-ab,a^2-b^2$.
+This gives us two possible values of $\epsilon$: $(a^2-b^2)^2,(ab)^2$.
+It would seem that an optimal configuration can be obtained by setting $a^2-b^2=ab$.
+If we add in the constraint that $a^2+b^2=1$ (since the six vectors we're interested in must live on the unit sphere), we get
 $$
@@
 $$
-If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we have that $\epsilon=\frac{1}{5}$, implying that these six points give us an optimal configuration.
-\subsection{N=n+1}
-Let us consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space.
-From Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we will find a configuration of vectors that achieves this lower bound.
-First we introduce the definition of a simplex, and then we will proceed to characterize $\epsilon$.
+If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$; coupling this knowledge with the fact that for $(n, N) = (3, 6)$, we see that these six points give us an optimal configuration.
+\subsection{$N=n+1$}
+Let us now consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space.
+From Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so if we show that there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$, then we can conclude that $1/n^2$ is the optimal value of $\epsilon$ for $N = n+1$.
+
+To do this, we first define what we mean by an $n$-dimensional simplex.
 
@@
 
-Before proceeding with the theorem, we notice that when $n=2$, the configuration of $N=3$ vectors that produced the minimum value of $\epsilon$ made up a simplex centered at the origin.
+Before proceeding with the theorem, we notice that when $n=2$, the configuration of $N=3$ vectors that produced the minimum value of $\epsilon$ made up a simplex centered at the origin (an equilateral triangle).
 
@@
 
-Since visualizing in spaces of dimensionality greater than $3$ is extremely hard, we used a computational approach to compute near-optimal epsilon values for different $(n, N)$ pairs.
+Since visualizing in spaces of dimensionality $3$ or greater is extremely hard, we used a computational approach to compute near-optimal epsilon values for different $(n, N)$ pairs.
 We describe this method in greater detail below.
@@
 
-At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector.
-
-To make this more precise, we present Python code for the algorithm above.
+At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector. We present pseudocode for this algorithm above.
 
@@
 
-We now look to calculate the optimum value of $\epsilon$ for the specific case of $N = n+1$.
-Before proving an upper bound on what the optimum value of $\epsilon$ can be, we introduce the parallelogram law.
+We now consider a variation of the above problem, where we actually try to place $N$ vectors in $n$-space as far apart as possible from each other. More concretely, if we define $\delta$ as
+
+$$\delta = \max_{i \neq j} v_i \cdot v_j$$,
+
+then we can formalize our problem as finding the $N$ vectors $v_1, v_2, \ldots, v_N$ that minimize $\delta$.
+
+Before proving what the optimum value of $\delta$ is for $N= n+1$, we introduce the parallelogram law.
 
@@
 \subsection{Variations: Minimizing the dot product}
-Let us consider the situation in which we try to minimize $\delta= \max_{i\neq j}\{v_i\cdot v_j\}$ for $n+1$ vectors in an $n$ dimensional space.
+Given the above bound, we proceed to state and prove the following theorem.
 
@@
 Since $\delta_{min} \le -\frac{1}{n}$ and $\delta_{min} \ge -\frac{1}{n}$, we can conclude that $\delta_{min}=-\frac{1}{n}$.
-From this, we can see that a $\epsilon = \frac{1}{n^2}$ is achievable for the general case of when $N=n+1$ -- this leads us to conclude that the minimum possible $\epsilon$ value for $N=n+1$ is at most $\frac{1}{n^2}$.
 \end{proof}
 
-Note that proving that the minimum possible epsilon for the case of $N = n+1$ is equal to $1/n^2$ is much harder, and we leave this for future work.
-
 \end{document}

### 2014-12-10 project3/paper/paper.tex
more edits to intro

@@
 \[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
-In this paper, we consider the problem of minimizing $\epsilon$ for different values of $n$ and $N$.
+What is the minimal attainable $\epsilon$?
 
-For $n=N$, it is clear that the minimum value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis.
+For $n=N$, it is clear that the optimal value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis.
 For fixed $n$, it is also easy to see that the minimum value of $\epsilon$ is nondecreasing as $N$ increases: that is, if $\epsilon_1$ is the minimum value of $\epsilon$ for $N$ unit vectors in $\mathbb R^n$, and $\epsilon_2$ is the minimum value of $\epsilon$ for $N+1$ unit vectors in $\mathbb R^n$, we must have
@@
 
-In this paper, we shall see that for $n=2$, the minimum value of $\epsilon$ will be given by 
+In Section 2, we shall see that for $n=2$, the minimum value of $\epsilon$ is given by 
 \[\epsilon=\cos^2\pi/N.\]
 
-We will also derive a lower bound on $\epsilon$ for any choice of $n$ and $N$. In addition, we shall see that, for $N=n+1$, the minimum value of $\epsilon$ is given by
+We will then derive in Section 3 a lower bound on $\epsilon$ for any choice of $n$ and $N$. In Section 4 we shall see that, for $N=n+1$, the minimum value of $\epsilon$ is given by
 \[\epsilon=\frac{1}{n^2},\] 
 and this configuration is produced when the $n+1$ vectors point to the vertices of the $n$-simplex.
-We shall also exhibit an optimal configuration of six vectors in $\mathbb R^3$; this optimal configuration produces an $\epsilon$ equal to
+We will also exhibit an optimal configuration of six vectors in $\mathbb R^3$; this optimal configuration produces an $\epsilon$ equal to
 \[\epsilon=\frac 1 5.\]
+
+In Section 5, we present the results of a computational approach to computing minimal values of $\epsilon$. Finally, in Section 6 we consider the related problem of minimizing 
+\[\delta=\max_{i\neq j}v_i\cdot v_j.\]
+In particular, we will show that in the case where $N=n+1$, the simplex remains the optimal configuration.
 

### 2014-12-10 project3/paper/paper.tex
more edits

@@
 For $n=N$, it is clear that the optimal value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis.
+And for $N>n$, we must have $\epsilon>0$ as any collection of $N$ vectors is linearly dependent, while any collection of orthonormal vectors is linearly independent.
+
 For fixed $n$, it is also easy to see that the minimum value of $\epsilon$ is nondecreasing as $N$ increases: that is, if $\epsilon_1$ is the minimum value of $\epsilon$ for $N$ unit vectors in $\mathbb R^n$, and $\epsilon_2$ is the minimum value of $\epsilon$ for $N+1$ unit vectors in $\mathbb R^n$, we must have
@@
 
-It is obvious that for $N=2$, we can find an $\epsilon$ equal to $0$, since it easy to pick two unit vectors $v_1$ and $v_2$ of unit length that are orthogonal to each other -- a simple example of such vectors is $[1, 0]^T$ and $[0, 1]^T$.
+As stated previously, when $N=2$ we can easily obtain an $\epsilon$ equal to $0$; a simple example of such a configuration is $[1, 0]^T$ and $[0, 1]^T$.
 
-However, we see that the problem gets harder for larger $N$. Let us first consider the specific case of $N=3$; once we build some intuition for the problem, we will try to generalize to arbitrary $N$.
-\\
-
-It's clear that when $N=3$, the value of $\epsilon$ must be greater than $0$, since there is no way three dimensionality-$2$ vectors can be orthogonal to each other.
-We make this more concrete in the following theorem.
+The problem, however, gets harder for larger $N$. Let us first consider the specific case of $N=3$. Once we build some intuition for the problem, we will generalize the result for arbitrary $N$.
 
@@
 
-We can use a well-known lemma to provide a useful lower bound on $\epsilon$ for arbitrary $(n, N)$ pairs.
+In the search for a lower bound on $\epsilon$ for arbitrary $(n, N)$ pairs, it is natural to consider the so-called \emph{Gram matrix} of our collection of vectors. 
+\begin{definition}
+The Gram matrix of a collection of $N$ vectors $v_1,\ldots, v_N$ in $\mathbb R^n$ is the $N$-by-$N$ matrix given by
+\[G(v_1,\ldots, v_N)=\left(\begin{matrix}
+v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
+v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
+\vdots & \vdots & \ddots & \vdots\\
+v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
+\end{definition}
+Note that the Gram matrix can be written as the product $A^TA$, where $A$ is the $n$-by-$N$ matrix
+\[A=\left(\begin{matrix}
+| & | & \cdots & | & |\\
+v_1 & v_2 & \cdots & v_{N-1} & v_N\\
+| & | & \cdots & | & | \end{matrix}\right).\]
+We can use a well-known lemma on the rank of certain real symmetric matrices to derive a useful lower bound.
 Our proof of the lemma follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.
@@
 \begin{proof}
-Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Let us define the $n$-by-$N$ matrix
+Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Recall the definitions of $A$ and $G$:
 \[A=\left(\begin{matrix}
@@
 | & | & \cdots & | & | \end{matrix}\right).\] 
-Now consider the $N$-by-$N$ Gram matrix:
 \[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
@@
 v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
-Let $d$ be the rank of $G$; since $A$ has rank at most $n$, we have that $d\leq n$.
+Let $d$ be the rank of $G$.
+Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we must have that $d\leq n$.
 Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$.

### 2014-12-10 project3/code/compute_epsilon_estimate.py
Typo fix

@@ -102,7 +102,8 @@ def get_min_epsilon_with_annealing(n, N, num_iter, start_vectors):
 
 
 if __name__ == '__main__':
-    values = [(2, 2), (2, 3), (2, 4), (2, 5), (2, 6), (3, 3), (3, 4), (3, 5), (3, 6), (4, 4), (4, 5), (4, 6), (4, 7), (4, 8), (4, 9), (4, 10), (4, 11), (4, 12), (16, 144)]
+    # values = [(2, 2), (2, 3), (2, 4), (2, 5), (2, 6), (3, 3), (3, 4), (3, 5), (3, 6), (4, 4), (4, 5), (4, 6), (4, 7), (4, 8), (4, 9), (4, 10), (4, 11), (4, 12), (16, 144)]
+    values = [(2, 7), (2, 8), (2, 9), (2, 10), (2, 11), (2, 12), (3, 7), (3, 8), (3, 9), (3, 10), (3, 11), (3, 12)]
     for (n, N) in values:
         min_epsilon = 1.0
         best_vectors = None

### 2014-12-10 project3/paper/paper.tex
Typo fix
more edits

@@
 
-Before proving the above theorem, we state and prove the following lemma.
+Before proving the above theorem, we state and prove the following lemma, which will be useful for the general case as well.
 
@@
 Let $d$ be the rank of $G$.
-Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we must have that $d\leq n$.
+Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we have that $d\leq n$.
 Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$.

### 2014-12-11 project3/paper/paper.tex
more edits through part of Section 4

@@
 
-\item One of $\theta_1, \theta_2, \dots, \theta_n$ is greater than $\pi/2$:
-
-Then, if $\theta_1 > \pi/2$ we see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ which is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$.
+\item One of $\theta_1, \theta_2, \dots, \theta_n$ is greater than $\pi/2$: Then in particular $\theta_1 > \pi/2$.
+We see that $\cos^2 \theta_1 = \cos^2 (\pi - \theta_1)$ which is equal to $\cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n)$.
 Furthermore, $\pi - \theta_1 = \theta_2 + \theta_3 + \ldots + \theta_n < \pi/2$, which means $\cos^2 \theta_1 = \cos^2 (\theta_2 + \theta_3 + \ldots + \theta_n) < \cos^2 \theta_n$.
@@
 
-In addition, we see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that in this case too, $\alpha$ is equal to $\cos^2 \theta_n$.
+In addition, we see that for all $i \in \{2, 3, \ldots, n-1\}$, $\cos^2 \theta_i < \cos^2 \theta_n$, from which we can conclude that $\alpha$ is still equal to $\cos^2 \theta_n$.
 
@@
 
-Since we don't care about the sign of the dot products between any two vectors (we're looking at the squares of the dot products in the above equation), and since $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin (multiplied by $-1$).
+Because the sign of the dot product of any two vectors does not matter, and because $$|v \cdot v'| = |v \cdot (-v')|$$ for some arbitrary vector $v'$, we see that it's easy to transform the three vectors $v_1$, $v_2$ and $v_3$ so that they all lie in the same semi-circle -- to accomplish this, at most one vector needs to be reflected about the origin (multiplied by $-1$).
 
@@
 | & | & \cdots & | & | \end{matrix}\right).\]
-We can use a well-known lemma on the rank of certain real symmetric matrices to derive a useful lower bound.
+We can use a well-known lemma on the rank of certain real symmetric matrices to derive a useful lower bound on $\epsilon$.
 Our proof of the lemma follows that of a paper by Noga Alon\footnote{Lemma 2.2, \emph{Perturbed identity matrices have high rank: proof and applications}, Noga Alon.}.
@@
 \end{proof}
-We present now a table of the bounds the theorem gives to us for varying $n$ and $N$:
+We present now a table of the bounds the theorem gives us for varying $n$ and $N$:
 \begin{table}[h]
@@
 \subsection{$n=3,N=6$}
-In this situation we have a lower bound of $\epsilon\ge\frac{1}{5}$ from the above theorem. Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
+Let us consider collections of six vectors in $\mathbb R^3$.
+As can be seen in Table 1, the lower bound provided in this case by Theorem \ref{lower_bound} is $\frac 1 5$ as well.
+We shall now see, in fact, that this lower bound is attainable.
+\begin{theorem}
+For $n=3$ and $N=6$, the optimum value of $\epsilon$ is given by $\frac 1 5$.
+\end{theorem}
+\begin{proof}Consider the following 6 vectors on the unit sphere in $\R^3$ with variables $a$ and $b$. 
 $$
@@
 $$
-If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$; coupling this knowledge with the fact that for $(n, N) = (3, 6)$, we see that these six points give us an optimal configuration.
+If we compute $\epsilon=(a^2-b^2)^2=(ab)^2$, we see that $\epsilon=\frac{1}{5}$.
+Coupled with our knowledge that $\frac 1 5$ is a lower bound on $\epsilon$, we conclude that these six points give us an optimal configuration.\end{proof}
 \subsection{$N=n+1$}
 Let us now consider the situation in which we try to minimize $\epsilon$ for $n+1$ vectors in an $n$ dimensional space.
-From Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so if we show that there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$, then we can conclude that $1/n^2$ is the optimal value of $\epsilon$ for $N = n+1$.
+From Theorem \ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so it suffices to show there exist some $n+1$ vectors in $n$-space that achieve an $\epsilon$ equal to $1/n^2$.
 
-To do this, we first define what we mean by an $n$-dimensional simplex.
+To do this, we first introduce the notion of an $n$-dimensional simplex.
 
@@
 \begin{theorem}
-If $N=n+1$ and the vectors in question are denoted $v_1,\cdots, v_N$, then over all configurations of these vectors the minimum value of $\epsilon$ for $i\neq j$ is $\frac{1}{^2n}$.
-This is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
+If $N=n+1$, then the minimum value of $\epsilon$ for $i\neq j$ is $\frac{1}{n^2}$. This $\epsilon$ is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
 \label{simplex}
@@
 $$
-So for this configuration $\epsilon=\frac{1}{n^2}$. By Theorem ~\ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we have achieved the lower bound. 
+So for this configuration $\epsilon=\frac{1}{n^2}$. By Theorem \ref{lower_bound} we know that $\epsilon\ge \frac{1}{n^2}$, so we have achieved the lower bound. 
 \end{proof}

### 2014-12-11 project3/paper/paper.tex
more edits through part of Section 6

@@
 
-\section{Some Optimal Configurations}
+\section{Some optimal configurations}
 Below we provide choices of $n$ and $N$ for which we can achieve the lower bound given by Theorem ~\ref{lower_bound}.
@@
 
-Since visualizing in spaces of dimensionality $3$ or greater is extremely hard, we used a computational approach to compute near-optimal epsilon values for different $(n, N)$ pairs.
+Because of the difficulty involved with visualization in dimensions 3 and above, we used a computational approach to compute near-optimal epsilon values for different $(n, N)$ pairs.
 We describe this method in greater detail below.
@@
 
-Even though this approach seems simple, it actually produces remarkably precise values -- this can be verified by cross-checking with results we've already proven / conjectured in this paper, this can be seen from values in Table $2$.
-
-In particular, the computed values for $n=2$ are pretty accurate -- the computed values for different values of $N$ are very close to the expected value of $\cos^2 \pi/N$, that we already proved in Section $2$, which makes us confident about the accuracy of this numeric method.
+Though this approach seems simple, it produces  results which seem to agree with the $\epsilon$ we derived for $(2, N)$ in Section 2, and with the $\epsilon$ we derived for $(3, 6)$ and $(n, n+1)$ in Section 4.
+This gives us some confidence in the accuracy of our algorithm.
+However, it must be noted that the data do not agree with the bounds produced by Theorem \ref{lower_bound}.
+There are multiple interpretations of this fact.
+First, it is possible that the lower bounds produced by Theorem \ref{lower_bound} are not always tight; they in fact are not for the case $(2, N)$.
+Alternatively, it is possible that our code simply becomes inaccurate for larger $n$ and $N$.
+This could be a fruitful direction for further research.
 
@@
 
-\section{Variations}
+\section{A variation of the problem}
 
-We now consider a variation of the above problem, where we actually try to place $N$ vectors in $n$-space as far apart as possible from each other. More concretely, if we define $\delta$ as
+We now consider a variation of the above problem, where we actually try to place $N$ unit vectors in $n$-space as far apart as possible from each other. More concretely, we define $\delta$ as
+$$\delta = \max_{i \neq j} v_i \cdot v_j,$$
+and we formalize our problem as finding the $N$ vectors $v_1, v_2, \ldots, v_N$ that minimize $\delta$.
 
-$$\delta = \max_{i \neq j} v_i \cdot v_j$$,
-
-then we can formalize our problem as finding the $N$ vectors $v_1, v_2, \ldots, v_N$ that minimize $\delta$.
-
-Before proving what the optimum value of $\delta$ is for $N= n+1$, we introduce the parallelogram law.
+Our results on this section concern the case when $N=n+1$. We shall see, in fact, that the optimal value of $\delta$ is given by $-\frac{1}{n}$, we introduce the parallelogram law.
 

### 2014-12-11 project3/paper/paper.tex
More changes to paper

@@
 \[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
-What is the minimal attainable $\epsilon$?
+In this paper, we try to address the question of what the minimal attainable $\epsilon$ is.
 
 For $n=N$, it is clear that the optimal value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis.
-And for $N>n$, we must have $\epsilon>0$ as any collection of $N$ vectors is linearly dependent, while any collection of orthonormal vectors is linearly independent.
+And for $N>n$, we must have $\epsilon>0$ as any collection of $N$ vectors in $n$-space is linearly dependent, while any collection of orthonormal vectors is linearly independent.
 
@@
 
-Again without loss of generality, we can assume that the vectors $v_1, v_2, \ldots, v_N$ are on or above the $x$-axis, and that $v_1 = [1,0]^T$ -- if any vector $v_i$ were not above the $x$-axis, then we could just consider $-v_i$ (the reflection of $v_i$ about the $x$-axis) instead.
+Again without loss of generality, we can assume that the vectors $v_1, v_2, \ldots, v_N$ are on or above the $x$-axis, and that $v_1 = [1,0]^T$ -- if any vector $v_i$ were not above the $x$-axis, then we could just consider $-v_i$ (the reflection of $v_i$ about the origin) instead.
 
@@
 
-Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
+Recall the Cauchy-Schwarz inequality for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
 \[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
@@
 
-Note that the bounds for $n=2$ are not tight, as we proved in Section 2 that the minimum $\epsilon$ was given by $\cos^2(\pi/N)$.
+Note that the bounds for $n=2$ are not tight, as we proved in Section 2 that the minimum $\epsilon$ is given by $\cos^2(\pi/N)$ when $n=2$.
 We shall see, however, that the bounds for $\epsilon$ are in fact tight in the cases where $N=n+1$ and where $n=3, N=6$.
@@
 \begin{theorem}
-If $N=n+1$, then the minimum value of $\epsilon$ for $i\neq j$ is $\frac{1}{n^2}$. This $\epsilon$ is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
+If $N=n+1$, then the minimum value of $\epsilon$ for $\frac{1}{n^2}$. This $\epsilon$ is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
 \label{simplex}
@@
 \begin{proof}
-Let $\epsilon$ be as before for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$.
+Let $\epsilon$ be defined as before for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$.
 We show that the origin-centered simplex configuration of the vectors $v_1,\cdots v_{n+1}$ produces a value $\epsilon=\frac{1}{n^2}$.
@@
 This gives us some confidence in the accuracy of our algorithm.
-However, it must be noted that the data do not agree with the bounds produced by Theorem \ref{lower_bound}.
-There are multiple interpretations of this fact.
-First, it is possible that the lower bounds produced by Theorem \ref{lower_bound} are not always tight; they in fact are not for the case $(2, N)$.
+However, it must be noted that the data does not agree with the bounds produced by Theorem \ref{lower_bound}.
+There are multiple interpretations for this fact.
+First, it is possible that the lower bounds produced by Theorem \ref{lower_bound} are not always tight -- this is easy to see for the case of $(2, N)$.
 Alternatively, it is possible that our code simply becomes inaccurate for larger $n$ and $N$.
@@
 
-Our results on this section concern the case when $N=n+1$. We shall see, in fact, that the optimal value of $\delta$ is given by $-\frac{1}{n}$, we introduce the parallelogram law.
+Our results on this section concern the case when $N=n+1$. We shall see, in fact, that the optimal value of $\delta$ is given by $-\frac{1}{n}$. To begin, we introduce the paralleogram law.
 
@@
 
-First we give the definition of the norm on vectors in $\R^n$ in terms of the dot product:
+First we give the definition of the norm of a sum of vectors in $\R^n$ in terms of the dot product:
 
@@
 
-\subsection{Variations: Minimizing the dot product}
+\subsection{Variation: Minimizing the dot product}
 Given the above bound, we proceed to state and prove the following theorem.

### 2014-12-11 project3/paper/paper.tex
More changes

@@
 
-From the above lemma, we can conclude that if $j = \text{argmin } \theta_i$ and $\theta_1 + \theta_2 + \theta_3 = \pi$, then $\epsilon = \cos^2 \theta_j$.
+From the above lemma, if $j = \text{argmin } \theta_i$ and $\theta_1 + \theta_2 + \theta_3 = \pi$, then $\epsilon = \cos^2 \theta_j$.
 
 Furthermore, we can obtain an upper bound on $\theta_j$ by observing that $\theta_ 1 + \theta_ 2 + \theta_3 \geq \theta_ j + \theta_j + \theta_j = 3 \theta_j \Rightarrow \theta_j \leq \pi/3$.
-Since we're interested in the smallest such $\epsilon$ and since the cosine function is a decreasing function in $\theta$ between $0$ and $\pi/2$, we conclude that the optimum value of $\epsilon$ for $n=2$ and $N=3$ is $\cos^2 \pi / 3$.
+Since we're interested in the smallest such $\epsilon$ and since the cosine function is a decreasing function in $\theta$ between $0$ and $\pi/2$, the optimum value of $\epsilon$ for $n=2$ and $N=3$ is $\cos^2 \pi / 3$.
 Figure $2$ shows this optimum configuration.
@@
 Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\}$$, where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$.
+(the above set must be non-empty since it must at least contain the element $N$)
 
@@
 
-Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we conclude that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$ (again making use of the fact that the cosine function is decreasing between $0$ and $\pi/2$).
+Since $\theta_ 1 + \theta_2 + \ldots + \theta_N \geq N \cdot \theta_j \Rightarrow \theta_j \leq \pi/N$, we see that the optimum $\epsilon$ value is in fact equal to $\cos^2 \pi/N$ (again making use of the fact that the cosine function is decreasing between $0$ and $\pi/2$).
 \end{proof}
@@
 \begin{lemma}
-Let $A=(a_{i,j})$ be an $n$-by-$n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
+Let $A=(a_{i,j})$ be a $n$-by-$n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
 \[\epsilon\geq \frac{n-d}{d(n-1)}.\]

### 2014-12-11 project3/paper/paper.tex
edits

--- 2014-12-11 project3/paper/paper.tex
@@
 
-\title{Almost orthogonal vectors}
+\title{Almost Orthogonal Vectors}
 
@@
 
-\section{$n=2$, generic $N$}
-In this section, we consider the problem of almost orthogonal vectors of dimensionality $2$.
+\section{$N$ vectors in $\R^2$}
+In this section, we consider the problem of almost orthogonal vectors in $\R^2$.
 
@@
 Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$.
-Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\}$$, where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$.
+Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\},$$ where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$.
 
-From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.
+From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.
 
@@
 
-\section{A lower bound}
+\section{A lower bound on $\epsilon$}
 
@@
 \begin{lemma}
-Let $A=(a_{i,j})$ be an $n$-by-$n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
+Let $A=(a_{i,j})$ be an $n\times n$ real symmetric matrix with $a_{i,i}=1$ for all $i$ and $|a_{i,j}|\leq \sqrt{\epsilon}$ for all $i\neq j$. Let $d$ be the rank of $A$. Then
 \[\epsilon\geq \frac{n-d}{d(n-1)}.\]
@@
 v_1 & v_2 & \cdots & v_{N-1} & v_N\\
-| & | & \cdots & | & | \end{matrix}\right).\] 
+| & | & \cdots & | & | \end{matrix}\right)\] 
 \[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
@@
 \vdots & \vdots & \ddots & \vdots\\
-v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
+v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right)\]
 Let $d$ be the rank of $G$.
@@
 $$
-a=\sqrt{\frac{1}{10}(5-\sqrt{5}}, b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
+a=\sqrt{\frac{1}{10}(5-\sqrt{5}}$$$$ b=-\frac{1+\sqrt{5}}{2}\sqrt{\frac{1}{10}(5-\sqrt{5}}. 
 $$
@@
 
-Figure $5$ shows diagrams of a 2-dimensional simplex.
+Figure $5$ shows diagrams of a 2-dimensional simplex. Notice that an $n$ dimensional simplex is constructed in $n+1$ dimensions, but exists in a hyperplane of $\R^{n+1}$.
 
@@
 
+As another note, in some instances the algorithm sometimes finds extrema that are not the extrema we have identified in cases where we provide optimal configurations.
+For example, for $n=3,N=4$, the $\epsilon$ value attained is very close to the optimal epsilon value of $\frac{1}{9}$, however the corresponding configuration is not a tetrahedron.
+This implies that in some cases there may be multiple configurations of vectors that are inequivalent up to rotation yet produce the optimal $\epsilon$ value.
+
 \begin{table}
@@
 
-\section{A variation of the problem}
+\section{A variation}
 

### 2014-12-11 project3/paper/paper.tex
More changes

--- 2014-12-11 project3/paper/paper.tex
@@
 Note that here, the angle between any two vectors in $v_j$ and $v_k$ in $v_1, v_2, \ldots, v_N$ can be expressed as $\sum_{i=j} ^{k-1} \theta_i$.
-Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\}$$, where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$.
+Note that, however, if $\sum_{i=j}^{k-1} \theta_i \leq \pi/2$, then $\cos^2 \theta_j \geq \cos^2(\sum_{i=j}^{k-1} \theta_i)$ and that if $\sum_{i=j}^{k-1} \theta_i > \pi/2$, then $\cos^2 (\sum_{i=j}^{k-1} \theta_i) = \cos^2 (\pi - \sum_{i=j}^{k-1} \theta_i)$ which is less than $\cos^2 \theta_i$ for any $i$ in $$\{1, 2, \ldots, N-1, N\} \setminus \{j, j+1, \ldots, k-1\},$$ where $j, k \in \{1, 2, \ldots, N-1\}$ and $j \neq k$.
 (the above set must be non-empty since it must at least contain the element $N$)
 
-From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, then we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.
+From this we can conclude that $\epsilon$ is equal to $\max_i \cos^2 \theta_i$. Then, if $j = \text{argmin } \theta_i$, we see that $\epsilon = \cos^2 \theta_j$ by Lemma $2.2$.
 
@@
 
+\begin{table}
+   \centering
+    \begin{tabular}{ | c | c | c | c |}
+    \hline
+      & $n=2$ & $n=3$ & $n=4$ \\ \hline
+    $N=2$ & $0.0$ & $-$ & $-$ \\ \hline
+    $N=3$ & $0.25$ & $0.0$ & $-$ \\ \hline
+    $N=4$ & $0.5$ & $0.11$ & $0.0$ \\ \hline
+    $N=5$ & $0.66$ & $0.21$ & $0.07$ \\ \hline
+    $N=6$ & $0.75$ & $0.22$ & $0.14$ \\ \hline
+    $N=7$ & $0.82$ & $0.36$ & $0.18$ \\ \hline
+    $N=8$ &  $0.86$ & $0.44$ & $0.24$ \\ \hline
+    $N=9$ &  $0.89$ & $0.49$ & $0.26$ \\ \hline
+    $N=10$ & $0.91$ & $0.51$ & $0.32$ \\ \hline
+    $N=11$ & $0.93$  & $0.55$ & $0.35$ \\ \hline
+    $N=12$ & $0.94$ & $0.64$ & $0.41$ \\
+    \hline
+    \end{tabular}
+    \caption {Minimum values of $\epsilon$ for different values of $(n, N)$, as computed by our numerical method}
+\end{table}
+
+Our general intuition tells us that perturbing a set of nearly optimal vectors slightly could give us a new set of vectors that produce even a smaller epsilon value.
+Given this, we start off with a set of random vectors, and try to move these vectors towards smaller epsilon values.
+
+At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector. We present pseudocode for this algorithm below.
+
+Though this approach seems simple, it produces  results which seem to agree with the $\epsilon$ we derived for $(2, N)$ in Section 2, and with the $\epsilon$ we derived for $(3, 6)$ and $(n, n+1)$ in Section 4.
+This gives us some confidence in the accuracy of our algorithm.
+However, it must be noted that the data does not agree with the bounds produced by Theorem \ref{lower_bound}.
+There are multiple interpretations for this fact.
+First, it is possible that the lower bounds produced by Theorem \ref{lower_bound} are not always tight -- this is easy to see for the case of $(2, N)$.
+Alternatively, it is possible that our code simply becomes inaccurate for larger $n$ and $N$.
+This could be a fruitful direction for further research.
+
+As another note, in some instances the algorithm sometimes finds extrema that are not the extrema we have identified in cases where we provide optimal configurations.
+For example, for $n=3,N=4$, the $\epsilon$ value attained is very close to the optimal epsilon value of $\frac{1}{9}$, however the corresponding configuration is not a tetrahedron.
+This implies that in some cases there may be multiple configurations of vectors that are not equivalent up to rotation yet produce the optimal $\epsilon$ value.
+
 \begin{lstlisting}
@@
 
-Our general intuition tells us that perturbing a set of nearly optimal vectors slightly could give us a new set of vectors that produce even a smaller epsilon value.
-Given this, we start off with a set of random vectors, and try to move these vectors towards smaller epsilon values.
-
-At every iteration, we try adding a vector whose norm becomes smaller with every passing iteration, to the already computed optimal vector. We present pseudocode for this algorithm above.
-
-Though this approach seems simple, it produces  results which seem to agree with the $\epsilon$ we derived for $(2, N)$ in Section 2, and with the $\epsilon$ we derived for $(3, 6)$ and $(n, n+1)$ in Section 4.
-This gives us some confidence in the accuracy of our algorithm.
-However, it must be noted that the data does not agree with the bounds produced by Theorem \ref{lower_bound}.
-There are multiple interpretations for this fact.
-First, it is possible that the lower bounds produced by Theorem \ref{lower_bound} are not always tight -- this is easy to see for the case of $(2, N)$.
-Alternatively, it is possible that our code simply becomes inaccurate for larger $n$ and $N$.
-This could be a fruitful direction for further research.
-
-\begin{table}
-   \centering
-    \begin{tabular}{ | c | c | c | c |}
-    \hline
-      & $n=2$ & $n=3$ & $n=4$ \\ \hline
-    $N=2$ & $0.0$ & $-$ & $-$ \\ \hline
-    $N=3$ & $0.25$ & $0.0$ & $-$ \\ \hline
-    $N=4$ & $0.5$ & $0.11$ & $0.0$ \\ \hline
-    $N=5$ & $0.66$ & $0.21$ & $0.07$ \\ \hline
-    $N=6$ & $0.75$ & $0.22$ & $0.14$ \\ \hline
-    $N=7$ & $0.82$ & $0.36$ & $0.18$ \\ \hline
-    $N=8$ &  $0.86$ & $0.44$ & $0.24$ \\ \hline
-    $N=9$ &  $0.89$ & $0.49$ & $0.26$ \\ \hline
-    $N=10$ & $0.91$ & $0.51$ & $0.32$ \\ \hline
-    $N=11$ & $0.93$  & $0.55$ & $0.35$ \\ \hline
-    $N=12$ & $0.94$ & $0.64$ & $0.41$ \\
-    \hline
-    \end{tabular}
-    \caption {Minimum values of $\epsilon$ for different values of $(n, N)$, as computed by our annealing method}
-\end{table}
-
-\section{A variation of the problem}
+\section{A variation}
 
